### 2009-01-27 trunk/talk/draft1.tex

\documentclass[9pt]{beamer}

\usepackage{amsmath,amssymb,amstext,amsfonts,setspace,dsfont,amsmath,amsthm,array,nomencl,hyperref,pgf,xcolor}

\usepackage{beamerthemesplit}

%\usetheme{Dresden}
\usetheme{Ilmenau}

\setbeamertemplate{navigation symbols}{} %no navigation symbols

\setbeamertemplate{footline}
{%
\leavevmode%
\hbox{%

\begin{beamercolorbox}[wd=.3\paperwidth,ht=2.5ex,dp=1.125ex,left]{author
in head/foot}%
\usebeamerfont{author in head/foot}\hspace{.3cm}\insertshortauthor\hspace{.3cm}
\end{beamercolorbox}%

\begin{beamercolorbox}[wd=.5\paperwidth,ht=2.5ex,dp=1.125ex,center]{title
in head/foot}%
\usebeamerfont{title in head/foot}\hspace{.3cm}\insertshorttitle
\end{beamercolorbox}%

\begin{beamercolorbox}[wd=.2\paperwidth,ht=2.5ex,dp=1.125ex,right]{author
in head/foot}%
\usebeamerfont{author in head/foot}Slide\hspace{.1cm}\insertframenumber\hspace{.1cm}of\hspace{.1cm}\inserttotalframenumber\hspace{.3cm}
\end{beamercolorbox}%
}%
\vskip0pt%
}

\setbeamertemplate{blocks}[rounded][shadow=true]

\title{Linear and Differential Cryptanalysis of DES}
\author{Slava Chernyak and Sourav Sen Gupta}
\institute{University of Washington}
\date{February 2, 2009}

\begin{document}

\begin{frame}
\titlepage
\end{frame}

\section{Introduction}
\begin{frame}{Introduction}
\begin{block}{Quack!}
Introduction to DES and the two attacks
\end{block}
\end{frame}

\subsection{Data Encrytption Standard}
\begin{frame}
DES structure

\end{frame}

\begin{frame}
DES pieces

\end{frame}

\begin{frame}
DES mathematics

\end{frame}

\subsection{Linear Cryptanalysis}
\begin{frame}
Vulnerabilities of DES

\end{frame}

\begin{frame}
How to exploit those - Linear attack idea

\end{frame}

\subsection{Differential Cryptanalysis}
\begin{frame}
Vulnerabilities of DES

\end{frame}

\begin{frame}
How to exploit those - Differential attack idea

\end{frame}

\section{Mathematical Framework}
\begin{frame}
Mathematical description of the attacks

\end{frame}

\subsection{Simplification WLOG}
\begin{frame}
Reducing DES down to SPN

\end{frame}

\begin{frame}
Introducing SPN mathematically

\end{frame}

\subsection{Linear Attack on SPN}
\begin{frame}
Basic math concepts behind Linear attack ...

\end{frame}

\begin{frame}
Intro to linear / affine approximation ...

\end{frame}

\begin{frame}
Linear characteristic / approximation of SBox ...

\end{frame}

\begin{frame}
Probabilities for the approximations ...

\end{frame}

\begin{frame}
Piling Up lemma ...

\end{frame}

\begin{frame}
Combining across SPN rounds ...

\end{frame}

\begin{frame}
The algorithm for the attack ...

\end{frame}

\begin{frame}
Probabilistic justification ...

\end{frame}

\subsection{Differential Attack on SPN}
\begin{frame}
Basic math concepts behind Differential attack ...

\end{frame}

\begin{frame}
Intro to differentials ...

\end{frame}

\begin{frame}
Differential characteristic / approximation of SBox ...

\end{frame}

\begin{frame}
Probabilities for the approximations ...

\end{frame}

\begin{frame}
Analogue of Piling Up lemma ...

\end{frame}

\begin{frame}
Combining across SPN rounds ...

\end{frame}

\begin{frame}
The algorithm for the attack ...

\end{frame}

\begin{frame}
Probabilistic justification ...

\end{frame}

\subsection{Actual Attacks}
\begin{frame}
Demo of attacks on SPN ... Runtime comparisons with Brute Force ...

\end{frame}

\begin{frame}
Break ... After coming back from the break, we should refresh the basic ideas behind the attacks once again ...

\end{frame}

\section{Attacks on DES}
\begin{frame}
Back to DES ... Pull back from SPN ...

\end{frame}

\subsection{Linear Attack: 3-round DES}
\begin{frame}
Linear Attack idea ...

\end{frame}

\begin{frame}
Matsui NS for SBoxes ...

\end{frame}

\begin{frame}
Linear approximations for SBoxes ...

\end{frame}

\begin{frame}
Piling up over 3 rounds ...

\end{frame}

\begin{frame}
Demo of the attack ...

\end{frame}

\subsection{Differential Attack: 3-round DES}
\begin{frame}
Differential Attack idea ...

\end{frame}

\begin{frame}
Differentials for SBoxes ...

\end{frame}

\begin{frame}
Differential characteristics for SBoxes ...

\end{frame}

\begin{frame}
Piling up over 3 rounds ...

\end{frame}

\begin{frame}
Demo of the attack ...

\end{frame}

\subsection{Extension to full DES}
\begin{frame}
Extension of the same ideas to 6 / 8 / 16 round DES ... Probabilistic explanation ...

\end{frame}

\section{Conclusion}
\begin{frame}
Concluding remarks about the project ...

\end{frame}

\subsection{Independence of S-Boxes}
\begin{frame}
Refresh math idea ... Independence assumption ...

\end{frame}

\begin{frame}
What if that is false ? ... Discussion ...

\end{frame}

\subsection{Number of Pairs Required}
\begin{frame}
Probabilistic possibility of the attacks ... Pairs of data points needed ...

\end{frame}

\begin{frame}
Generalized attack idea on any given system ... Discussion ...

\end{frame}

\subsection{Links between the Attacks}
\begin{frame}
Links between Linear and Differential Attacks ...

\end{frame}

\begin{frame}
Mathematically and while implementing ...

\end{frame}

\subsection{Take home Lesson}
\begin{frame}
If you are mounting the attacks ... To Do / Not To Do ...

\end{frame}

\begin{frame}
If you are building a cryptosystem ... To Do / Not To Do ...

\end{frame}

\begin{frame}
Questions ... Doubts ... Coding ... Misc ... Discussion ...

\end{frame}

\end{document}

### 2009-01-29 trunk/talk/draft1.tex

@@
 \begin{frame}
-How to exploit those - Linear attack idea
+Linear Attack on DES idea:\\
+\begin{itemize}
+\item S-Boxes depend on a relatively small number of bits (6) which allows us to write down linear (or afine) expressions that approximate S-Boxes \\
+\item The effects of one round do not diffuse quickly over following rounds. Thus linear expresions (as above) that hold per-round can be combined across rounds.
+\end{itemize}
+\end{frame}
 
+\begin{frame}
+Specifically, if $P_{i}$ are plaintext bits, $C_{i}$ are cyphertext bits, and $K_{i}$ are subkey bits, then we wish to find an expression of the form
+\[ P_{i_1} \oplus P_{i_2} \cdots P_{i_j} \oplus C_{i_1} \oplus C_{i_2} \cdots C_{i_k} = K_{i_1} \oplus K_{i_2} \oplus K_{i_m} \]
+Such that this expression has a high \textit{or} low probability of occurence.\\
+\vspace{5mm}
+Consider: No such obvious expression should exist, otherwise the cipher is trivially weak.
+If we were to randomly select bits for the above expression, it would hold exactly $1/2$ the time.
+If we find an expression such as above that displays a high \textit{bias}, that is, it holds much more or less freqently than $1/2$ the time, we can exploit this.
 \end{frame}
@@
 
+We can remove much of the complexity of DES without affecting the discussion of the techniques of Linear Cryptanalysis.
+We define a simple Substitution-Permutation Network Cipher (SPN) which has structural similarity to DES.
+
+\begin{itemize}
+\item SPN is a 16-bit block cipher (16-bit plaintext to 16-bit ciphertext) with 16-bit round keys.
+\item Each round consists of a substitution and a permutation (much like in DES)
+\item The substitution is the result of splitting the 16-bit input block to a round in to 4 4-bit sub-blocks.
+Each 4-bit sub-block is then mapped across a 4-bit to 4-bit S-box.
+We use the same S-box throughout.
+\item The permutation is applied to all 16-bits of the round following the substitution.
+\item Finally, key-mixing is achieved by XOR-ing the round key with every input block to a round as well as at the end of the last round (so that we can't simply ignore the last round)
+\item The SPN cipher we will look at will consist of 4 rounds.
+\end{itemize}
+
+\end{frame}
+
+\begin{frame}
+SPN Visually
+%\pgfimage[scale=0.75]{SPN.PNG}
 \end{frame}
@@
 \begin{frame}
-Basic math concepts behind Linear attack ...
-
+\large{Basic math concepts behind the Linear attack:}\\
+\vspace{5 mm}
+Linear cryptanalysis tries to take advantage of high probability occurrences of  linear expressions involving plaintext bits, ciphertext bits and subkey bits.  It is a known plaintext attack.\\
+\vspace{5 mm}
+Define $P_i$ where $i = 1, 2, \dots 16$ as the $i$-th plaintext bit.\\
+Define $C_i$ where $i = 1, 2, \dots 16$ as the $i$-th cyphertext bit.\\
+Define $U_{j,i}$ where $j = 1, 2, 3, 4$ and $i = 1, 2, \dots 16$ be the $i$-th input bit to the $j$-th round of SPN.\\
+Define $V_{j,i}$ where $j = 1, 2, 3, 4$ and $i = 1, 2, \dots 16$ be the $i$-th output bit of the $j$-th round of SPN.\\
 \end{frame}

### 2009-01-29 talk/draft1.tex
1st commit of Linear slides

--- 2009-01-29 trunk/talk/draft1.tex
@@
 \begin{frame}
-Introducing SPN mathematically
+SPN S-Box and Permutation:\\
+\vspace{5mm}
+SPN uses a single 4-bit S-Box that has the following structure:
+\begin{tabular}{|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|}
+\hline
+Input: & 0 & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & A & B & C & D & E & F \\
+\hline
+Output:& E & 4 & D & 1 & 2 & F & B & 8 & 3 & A & 6 & C & 5 & 9 & 0 & 7 \\
+\hline
+\end{tabular}\\
+\vspace{5mm}
+And the following 16-bit permutation:
+\begin{tabular}{|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|}
+\hline
+1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 & 15 & 16\\
+\hline
+1 & 5 & 9 & 13 & 2 & 6 & 10 & 14 & 3 & 7 & 11 & 15 & 4 & 8 & 12 & 16 \\
+\hline
+\end{tabular}
+\vspace{5mm}
 
@@
 \begin{frame}
-\large{Basic math concepts behind the Linear attack:}\\
+\large{Basic Definitions for the Linear attack:}\\
 \vspace{5 mm}
@@
 \begin{frame}
-Intro to linear / affine approximation ...
-
+Linear and Affine approximation of S-Box:\\
+\vspace{5mm}
+How do we come up with the desired expression for the entire cipher? We start by looking at the only non-linear component, the S-Box.\\
+\vspace{5mm}
+To find the linear/afine approximation of the S-Box we simply consider every possible expression of the input bits $X_i$ and output bits $Y_j$. Thus the expression has the form
+\[ \bigoplus_{i \in U} X_i = \bigoplus_{j \in V} Y_j \]
+Where $U$ and $V$ range over all possible subsets of $\{1, 2, 3 ,4\}$. We then compare how often this expression coincides with the S-Box. \\
+\vspace{5mm}
+Note that there are 16 possibilities for $U$ and $V$, hence 256 total possible expressions. (this is for a 4-bit S-box)
 \end{frame}
@@
 \begin{frame}
-Linear characteristic / approximation of SBox ...
-
+S-Box approximatin example:\\
+\vspace{5mm}
+We can take the following expression as an example:
+\[ X_2 \oplus X_3 = Y_1 \oplus Y_3 \oplus Y_4 \]
+Applying all possible values for the input $X$ bits it turns out that the expression holds in 12 out of the 16 cases. Hence, this expression has a bias of $12/16 - 1/2 = 1/4$.\\
+\vspace{5mm}
+Similarly the following expression
+\[ X_3 \oplus X_4 = Y_1 \oplus Y_4 \]
+Has a bias of $-3/8$, which is higher in magnitude than the previous one. It is the magnitude of the bias that is important.
 \end{frame}
@@
 \begin{frame}
-Probabilities for the approximations ...
+The number of agreements (minus 8) between the S-Box and every possible expression is summarized in the table below. Thus to get the bias, one must only divide by 16.\\
+\vspace{5mm}
+\tiny
+\begin{tabular}{|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|}
+\hline
+ &0&1&2&3&4&5&6&7&8&9&A&B&C&D&E&F\\
+\hline
+0&+8&0&0&0&0&0&0&0&0&0&0&0&0&0&0&0 \\
+\hline
+1&0&0&-2&-2&0&0&-2&\textbf{+6}&+2&+2&0&0&+2&+2&0&0 \\
+\hline
+2&0&0&-2&-2&0&0&-2&-2&0&0&+2&+2&0&0&\textbf{-6}&+2 \\
+\hline
+3&0&0&0&0&0&0&0&0&+2&\textbf{-6}&-2&-2&+2&+2&-2&-2 \\
+\hline
+4&0&+2&0&-2&-2&-4&-2&0&0&-2&0&+2&+2&-4&+2&0 \\
+\hline
+5&0&-2&-2&0&-2&0&+4&+2&-2&0&-4&+2&0&-2&-2&0  \\
+\hline
+6&0&+2&-2&+4&+2&0&0&+2&0&-2&+2&+4&-2&0&0&-2 \\
+\hline
+7&0&-2&0&+2&+2&-4&+2&0&-2&0&+2&0&+4&+2&0&+2  \\
+\hline
+8&0&0&0&0&0&0&0&0&-2&+2&+2&-2&+2&-2&-2&\textbf{-6} \\
+\hline
+9&0&0&-2&-2&0&0&-2&-2&-4&0&-2&+2&0&+4&+2&-2 \\
+\hline
+A&0&+4&-2&+2&-4&0&+2&-2&+2&+2&0&0&+2&+2&0&0 \\
+\hline
+B&0&+4&0&-4&+4&0&+4&0&0&0&0&0&0&0&0&0 \\
+\hline
+C&0&-2&+4&-2&-2&0&+2&0&+2&0&+2&+4&0&+2&0&-2 \\ 
+\hline
+D&0&+2&+2&0&-2&+4&0&+2&-4&-2&+2&0&+2&0&0&+2 \\
+\hline
+E&0&+2&+2&0&-2&-4&0&+2&-2&0&0&-2&-4&+2&-2&0 \\
+\hline
+F&0&-2&-4&-2&-2&0&+2&0&0&-2&+4&-2&-2&0&+2&0 \\
+\hline
+\end{tabular}
+\small
+\end{frame}
+
+\begin{frame}
+What this means for 1 Round:\\
+\vspace{5mm}
+Note, from the previous table, that the expression $X_3 \oplus X_4 = Y_1 \oplus Y_4$ has a bias of $-3/8$
 

### 2009-01-29 talk/draft1.tex
More Linear SPN slides...

@@
 \vspace{5mm}
-Note, from the previous table, that the expression $X_3 \oplus X_4 = Y_1 \oplus Y_4$ has a bias of $-3/8$
-
+Note, from the previous table, that the expression $X_1 \oplus X_3 \oplus X_4 = Y_2$ has a bias of $+1/4$. \\
+\vspace{5mm}
+Note also that $U_1 = P \oplus K_1$. 
+\vspace{5mm} We can now write down the following linear approximation across the 1st round of SPN:
+\begin{eqnarray*}
+V_{1,6} & = & U_{1,5} \oplus U_{1,7} \oplus U_{1,8} \mbox{  $\leftarrow$ S-Box approximation above}\\
+		  & = & (P_5\oplus K_{1,5}) \oplus (P_7\oplus K_{1,7}) \oplus (P_8\oplus K_{1,8})
+\end{eqnarray*}
+This expression holds with probability of $3/4$ (bias of $+1/4$)
 \end{frame}
@@
 \begin{frame}
-Piling Up lemma ...
+We can get expressions that hold with some non-$1/2$ probability for every round.
+But we must somehow combine them to write an expression relating the plaintext and ciphertext bits $\Rightarrow$ We use the Piling Up Principle:\\
+\vspace{5mm}
+Note: \\
+$X_1 \oplus X_2 = 0$ is a linear expression equivalent to $X_1 = X_2$\\
+$X_1 \oplus X_2 = 1$ is an affine expression equivalent to $X_1 \neq X_2$\\
+\vspace{5mm}
+Assume the following probability distribution
+\[ Pr(X_1 = i)  = \left\{ \begin{array}{ll}
+         p_1 & \mbox{$i = 0$};\\
+         (1 - p_1) & \mbox{$i = 1$}.\end{array} \right. \] 
+\[ Pr(X_2 = i)  = \left\{ \begin{array}{ll}
+         p_2 & \mbox{$i = 0$};\\
+         (1 - p_2) & \mbox{$i = 1$}.\end{array} \right. \] 
+\end{frame}
 
+\begin{frame}
+Piling Up Principle Continued:\\
+\vspace{5mm}
+Assuming that $X_1$ and $X_2$ are independant:
+\[ Pr(X_1 = i, X_2 = j)  = \left\{ \begin{array}{ll}
+         p_1 p_2 & \mbox{$i = 0, j = 0$};\\
+         (1 - p_1) p_2 & \mbox{$i = 1, j = 0$} \\
+			p_1 (1 - p_2) & \mbox{$i = 0, j = 1$} \\
+			(1 - p_1)(1 - p_2) & \mbox{$i = 1, j = 1$}.\end{array} \right. \] 
+Now, note that $X_1 \oplus X_2 = 0 \Rightarrow X_1 = 1, X_2 =1$ OR $X_1 = 0, X_2 = 0$. Hence:
+\begin{eqnarray*}
+ Pr(X_1 \oplus X_2 = 0) & = & Pr(X_1=0,X_2=0) + Pr(X_1=1,X_2=1) \\
+								& = & p_1p_2 + (1 - p_1)(1-p_2)
+\end{eqnarray*}
+If we write $p_k = 1/2 + \epsilon_k$ we have
+\[ Pr(X_1 \oplus X_2 = 0) = 1/2 + 2\epsilon_1 \epsilon_2 \]
+Hence the bias of $X_1 \oplus X_2 = 0$ is 
+\[ 2 \epsilon_1 \epsilon_2 \]
+That is, twice the product of the bias of the original expressions.
+\end{frame}
+
+\begin{frame}
+Piling Up Principle Continued - The Piling Up Lemma:\\
+\vspace{5mm}
+We extend the idea on the previous slide to the general case: For $n$ independant random binary variables, $X_1, X_2, \dots, X_n$, 
+\[ Pr(X_1 \oplus X_2 \cdots \oplus X_n = 0) = 1/2 + 2^{n-1} \prod_{i=1}^n \epsilon_i \]
+In terms of bias,
+\[ \epsilon_{1,2,\dots,n} = 2^{n-1} \prod_{i = 1}^n \epsilon_i \]
+Where $\epsilon_{1,2,\dots,n}$ is the total bias of $X_1 \oplus X_2 \cdots \oplus X_n = 0$. \\
+\vspace{5mm}
+Note that if some $\epsilon_k = 0$ (that is, some $X_k$ has no bias) then the total bias will be $0$.  
 \end{frame}

### 2009-01-29 trunk/talk/draft1.tex

@@
 \section{Introduction}
-\begin{frame}{Introduction}
-\begin{block}{Quack!}
-Introduction to DES and the two attacks
-\end{block}
+
+\begin{frame}
+\begin{beamercolorbox}[ht=2.5ex,dp=1.125ex,center,rounded=true,shadow=true]{author in head/foot}
+A quick review of DES
+\end{beamercolorbox}
 \end{frame}
@@
 \begin{frame}
-DES structure
+\begin{columns}
+\begin{column}{0.6\textwidth}
+\begin{block}{DES Function}
+\[ DES : (\mbox{Plaintext }(P), \mbox{ Key }(K))  \mapsto \mbox{Cipher }(C) \]
+\end{block}
 
+\begin{block}{DES Unit Blocks}
+\begin{align*}
+\sigma_{i+1} : \: & (0,1)^{64} \rightarrow (0,1)^{64} \\
+& (L_i, R_i) \mapsto (L_i \oplus f(R_i,K_{i+1}), R_i) \\
+\tau : \: & (0,1)^{64} \rightarrow (0,1)^{64} \\
+& (L_{i+1},R_{i+1}) \mapsto (R_{i+1},L_{i+1})
+\end{align*}
+\end{block}
+
+\begin{itemize}
+\item{Input for round $i+1$: $(L_i,R_i)$}
+\item{Subkey for round $i+1$: $K_{i+1}$}
+\item{Feistel function ($f$)}
+\item{16-rounds for $i = 0, 1, ..., 15$}
+\end{itemize}
+
+\end{column}
+\begin{column}{0.4\textwidth}
+\begin{figure}
+\includegraphics[totalheight=0.8\textheight]{des_algo.jpg}
+\caption{DES Algorithm}
+\end{figure}
+\end{column}
+\end{columns}
 \end{frame}
@@
 \begin{frame}
-DES pieces
+\begin{columns}
+\begin{column}{0.45\textwidth}
+\begin{block}{Feistel Function}
+\begin{align*}
+f : \: & (0,1)^{32} \times (0,1)^{48} \rightarrow (0,1)^{32} \\
+& (R,K)  \mapsto P (S_{Box}(E(R) \oplus K)) 
+\end{align*}
+\end{block}
 
+\begin{itemize}
+\item{Right half of the plaintext ($R$)}
+\item{Expansion function \\ $E : (0,1)^{32} \rightarrow (0,1)^{48}$}
+\item{Key for the round ($K$)}
+\item{Confusion function (S-Boxes) \\ $S_i: (0,1)^6 \rightarrow (0,1)^4$}
+\item{Diffusion function \\ $P : (0,1)^{32} \rightarrow (0,1)^{32}$}
+\end{itemize}
+\end{column}
+\begin{column}{0.55\textwidth}
+\begin{figure}
+\includegraphics[totalheight=0.6\textheight]{des_feistel.jpg}
+\caption{Feistel Round for DES}
+\end{figure}
+\end{column}
+\end{columns}
 \end{frame}
@@
 \begin{frame}
-DES mathematics
+Mathematically, the different pieces of DES are:
+\begin{itemize}
+\item{Initial Permutation $IP : (0,1)^{64} \rightarrow (0,1)^{64}$}
+\item{DES Unit Block functions $\sigma_{i+1} (L_i,R_i)$ for $i = 0,1,...,15$}
+\item{Transpositions $\tau : (L_{i+1},R_{i+1}) \mapsto (R_{i+1},L_{i+1})$ for $i = 0,1,...,14$}
+\item{Inverse Initial Permutation $IP^{-1} : (0,1)^{64} \rightarrow (0,1)^{64}$}
+\end{itemize}
 
+\begin{block}{Algebraic Representation of DES}
+\begin{align*}
+C &= DES_K (P) = IP^{-1} \sigma_{16} \tau \cdots \tau \sigma_1 IP (P) \\
+P &= DES^{-1}_K (C) = IP^{-1} \sigma_{1} \tau \cdots \tau \sigma_{16} IP (C)
+\end{align*}
+\end{block}
+
+Note: The inverse DES comes from the fact that $\tau^2 = \sigma_i^2 = 1$
 \end{frame}
@@
 \begin{frame}
-Vulnerabilities of DES
-
+\begin{beamercolorbox}[ht=2.5ex,dp=1.125ex,center,rounded=true,shadow=true]{author in head/foot}
+Basic Idea of Linear Cryptanalysis
+\end{beamercolorbox}
 \end{frame}
@@
 \begin{frame}
-Vulnerabilities of DES
-
+\begin{beamercolorbox}[ht=2.5ex,dp=1.125ex,center,rounded=true,shadow=true]{author in head/foot}
+Basic Idea of Differential Cryptanalysis
+\end{beamercolorbox}
 \end{frame}
@@
 \begin{frame}
-Mathematical description of the attacks
-
+\begin{beamercolorbox}[ht=2.5ex,dp=1.125ex,center,rounded=true,shadow=true]{author in head/foot}
+Dive into the Mathematics
+\end{beamercolorbox}
 \end{frame}
@@
 \subsection{Simplification WLOG}
+
 \begin{frame}
-Reducing DES down to SPN
+\begin{columns}
+\begin{column}{0.6\textwidth}
+\begin{figure}
+\includegraphics[totalheight=0.55\textheight]{des_feistel.jpg}
+\end{figure}
+Let us simplify DES by
+\begin{itemize}
+\item{Reducing the bitlength from $64$ to $16$}
+\item{Removing expansion function $E(R)$}
+\item{Taking identical S-Boxes of size $4$ bits}
+\item{Repeating over less rounds}
+\end{itemize}
+\end{column}
 
+\begin{column}{0.4\textwidth}
+\begin{figure}
+\includegraphics[totalheight=0.9\textheight]{spn.jpg}
+\end{figure}
+\end{column}
+\end{columns}
+\end{frame}
+
+\begin{frame}
+\begin{beamercolorbox}[ht=2.5ex,dp=1.125ex,center,rounded=true,shadow=true]{author in head/foot}
+Simple Substitution-Permutation Network
+\end{beamercolorbox}
+\end{frame}
+
+\begin{frame}
 We can remove much of the complexity of DES without affecting the discussion of the techniques of Linear Cryptanalysis.
@@
 \end{itemize}
-
 \end{frame}
@@
 \begin{frame}
-SPN Visually
-%\pgfimage[scale=0.75]{SPN.PNG}
-\end{frame}
+SPN S-Box and Permutation:\\
+\vspace{5mm}
+SPN uses a single 4-bit S-Box that has the following structure:
+\begin{figure}
+\includegraphics[width=0.9\textwidth]{spn_sbox.jpg}
+\end{figure}
 
-\begin{frame}
-Introducing SPN mathematically
+\vspace{5mm}
+And the following 16-bit permutation:
+\begin{figure}
+\includegraphics[width=0.9\textwidth]{spn_perm.jpg}
+\end{figure}
+
+\vspace{5mm}
 
@@
 \begin{frame}
-\large{Basic math concepts behind the Linear attack:}\\
+\begin{beamercolorbox}[ht=2.5ex,dp=1.125ex,center,rounded=true,shadow=true]{author in head/foot}
+Mathematics of Linear Cryptanalysis
+\end{beamercolorbox}
+\end{frame}
+
+\begin{frame}
+\large{Basic Definitions for the Linear attack:}\\
 \vspace{5 mm}
@@
 \begin{frame}
-Probabilities for the approximations ...
-
+The number of agreements (minus 8) between the S-Box and every possible expression is summarized in the table below. Thus to get the bias, one must only divide by 16.\\
+\vspace{5mm}
+\begin{figure}
+\includegraphics[height=0.7\textheight]{spn_linear_approx.jpg}
+\end{figure}
 \end{frame}
@@
 \begin{frame}
-Piling Up lemma ...
+What this means for 1 Round:\\
+\vspace{5mm}
+Note, from the previous table, that the expression $X_1 \oplus X_3 \oplus X_4 = Y_2$ has a bias of $+1/4$. \\
+\vspace{5mm}
+Note also that $U_1 = P \oplus K_1$. 
+\vspace{5mm} We can now write down the following linear approximation across the 1st round of SPN:
+\begin{eqnarray*}
+V_{1,6} & = & U_{1,5} \oplus U_{1,7} \oplus U_{1,8} \mbox{  $\leftarrow$ S-Box approximation above}\\
+		  & = & (P_5\oplus K_{1,5}) \oplus (P_7\oplus K_{1,7}) \oplus (P_8\oplus K_{1,8})
+\end{eqnarray*}
+This expression holds with probability of $3/4$ (bias of $+1/4$)
+\end{frame}
 
+\begin{frame}
+We can get expressions that hold with some non-$1/2$ probability for every round.
+But we must somehow combine them to write an expression relating the plaintext and ciphertext bits $\Rightarrow$ We use the Piling Up Principle:\\
+\vspace{5mm}
+Note: \\
+$X_1 \oplus X_2 = 0$ is a linear expression equivalent to $X_1 = X_2$\\
+$X_1 \oplus X_2 = 1$ is an affine expression equivalent to $X_1 \neq X_2$\\
+\vspace{5mm}
+Assume the following probability distribution
+\[ Pr(X_1 = i)  = \left\{ \begin{array}{ll}
+         p_1 & \mbox{$i = 0$};\\
+         (1 - p_1) & \mbox{$i = 1$}.\end{array} \right. \] 
+\[ Pr(X_2 = i)  = \left\{ \begin{array}{ll}
+         p_2 & \mbox{$i = 0$};\\
+         (1 - p_2) & \mbox{$i = 1$}.\end{array} \right. \] 
+\end{frame}
+
+\begin{frame}
+Piling Up Principle Continued:\\
+\vspace{5mm}
+Assuming that $X_1$ and $X_2$ are independant:
+\[ Pr(X_1 = i, X_2 = j)  = \left\{ \begin{array}{ll}
+         p_1 p_2 & \mbox{$i = 0, j = 0$};\\
+         (1 - p_1) p_2 & \mbox{$i = 1, j = 0$} \\
+			p_1 (1 - p_2) & \mbox{$i = 0, j = 1$} \\
+			(1 - p_1)(1 - p_2) & \mbox{$i = 1, j = 1$}.\end{array} \right. \] 
+Now, note that $X_1 \oplus X_2 = 0 \Rightarrow X_1 = 1, X_2 =1$ OR $X_1 = 0, X_2 = 0$. Hence:
+\begin{eqnarray*}
+ Pr(X_1 \oplus X_2 = 0) & = & Pr(X_1=0,X_2=0) + Pr(X_1=1,X_2=1) \\
+								& = & p_1p_2 + (1 - p_1)(1-p_2)
+\end{eqnarray*}
+If we write $p_k = 1/2 + \epsilon_k$ we have
+\[ Pr(X_1 \oplus X_2 = 0) = 1/2 + 2\epsilon_1 \epsilon_2 \]
+Hence the bias of $X_1 \oplus X_2 = 0$ is 
+\[ 2 \epsilon_1 \epsilon_2 \]
+That is, twice the product of the bias of the original expressions.
+\end{frame}
+
+\begin{frame}
+Piling Up Principle Continued - The Piling Up Lemma:\\
+\vspace{5mm}
+We extend the idea on the previous slide to the general case: For $n$ independant random binary variables, $X_1, X_2, \dots, X_n$, 
+\[ Pr(X_1 \oplus X_2 \cdots \oplus X_n = 0) = 1/2 + 2^{n-1} \prod_{i=1}^n \epsilon_i \]
+In terms of bias,
+\[ \epsilon_{1,2,\dots,n} = 2^{n-1} \prod_{i = 1}^n \epsilon_i \]
+Where $\epsilon_{1,2,\dots,n}$ is the total bias of $X_1 \oplus X_2 \cdots \oplus X_n = 0$. \\
+\vspace{5mm}
+Note that if some $\epsilon_k = 0$ (that is, some $X_k$ has no bias) then the total bias will be $0$.  
 \end{frame}
@@
 \subsection{Differential Attack on SPN}
+\begin{frame}
+\begin{beamercolorbox}[ht=2.5ex,dp=1.125ex,center,rounded=true,shadow=true]{author in head/foot}
+Mathematics of Differential Cryptanalysis
+\end{beamercolorbox}
+\end{frame}
+
 \begin{frame}
@@
 \begin{frame}
+\begin{beamercolorbox}[ht=2.5ex,dp=1.125ex,center,rounded=true,shadow=true]{author in head/foot}
+Let's get our hands dirty!
+\end{beamercolorbox}
+\end{frame}
+
+\begin{frame}
 Demo of attacks on SPN ... Runtime comparisons with Brute Force ...
@@
 \begin{frame}
-Break ... After coming back from the break, we should refresh the basic ideas behind the attacks once again ...
-
+\begin{beamercolorbox}[ht=2.5ex,dp=1.125ex,center,rounded=true,shadow=true]{author in head/foot}
+It's time for a Break!
+\end{beamercolorbox}
 \end{frame}
@@
 \begin{frame}
-Back to DES ... Pull back from SPN ...
-
+\begin{beamercolorbox}[ht=2.5ex,dp=1.125ex,center,rounded=true,shadow=true]{author in head/foot}
+Do these work for DES?
+\end{beamercolorbox}
 \end{frame}
@@
 \begin{frame}
-Concluding remarks about the project ...
-
+\begin{beamercolorbox}[ht=2.5ex,dp=1.125ex,center,rounded=true,shadow=true]{author in head/foot}
+Did we miss anything important?
+\end{beamercolorbox}
 \end{frame}
@@
 
-\subsection{Take home Lesson}
+\subsection{Take Home}
+\begin{frame}
+\begin{beamercolorbox}[ht=2.5ex,dp=1.125ex,center,rounded=true,shadow=true]{author in head/foot}
+`Take Home'
+\end{beamercolorbox}
+\end{frame}
+
 \begin{frame}

### 2009-01-30 trunk/talk/draft1.tex
More Linear SPN slides

@@
 \begin{frame}
-Combining across SPN rounds ...
-
+Now, come back to Per-round approximations. We can write down 4 approximations (Let $S_{i,j}$ repesent the $j$-th S-Box in the $i$-th round):
+\begin{eqnarray*}
+S_{1,2}: X_1 \oplus X_3 \oplus X_4 & = & Y_2 \\
+S_{2,2}: X_2 = Y_2 \oplus Y_4 \\
+S_{3,2}: X_2 = Y_2 \oplus Y_4 \\
+S_{3,4}: X_2 = Y_2 \oplus Y_4 \\
+\end{eqnarray*}
+Each of these has a probability bias magnitude of $1/4$. We can use the Piling Up Principle to combine them into a single expression relating plaintext bits to cyphertext bits.
 \end{frame}
@@
 \begin{frame}
-The algorithm for the attack ...
+Consider the first 2 rounds:\\
+\vspace{5mm}
+\[  X_1 \oplus X_3 \oplus X_4 = Y_2 \Rightarrow \]
+\[ V_{1,6} = (P_5\oplus K_{1,5}) \oplus (P_7\oplus K_{1,7}) \oplus (P_8\oplus K_{1,8}) \]
+\vspace{2mm}
+\[ X_2 = Y_2 \oplus Y_4 \Rightarrow \]
+\[ V_{2,6} \oplus V_{2,8} = V_{1,6} \oplus K_{2,6} \]
+\vspace{2mm}
+Each of these has a bias of magnitude $1/4$. We can combine:
+\[  V_{2,6} \oplus V_{2,8} \oplus P_5 \oplus P_7 \oplus P_8 \oplus K_{1,5} \oplus K_{1,7} \oplus K_{1,8} \oplus K_{2,6} = 0 \]
+By the Piling Up Principle this expression holds with bias:
+\[ 2 * 1/4 * 1/4 = 1/8 \]
+\end{frame}
 
+\begin{frame}
+Using this principle we can write the following equation over 3 rounds of SPN:
+\[ U_{4,6} \oplus U_{4,8} \oplus U_{4,14} \oplus U_{4, 16} \oplus P_5 \oplus P_7 \oplus P_8 = \Sigma K \]
+Where $\Sigma K$ is the sum over some key bits. Note that Since the key is fixed, $\Sigma K = 0$ or $1$ and thus we can ignore it since we only care about the bias.\\
+\vspace{5mm}
+The magnitude of the bias of the above expression, by the Piling Up Principle, is $1/32$. \\
+\vspace{5mm}
+Next we show how we can extract key bits using this information.
+\end{frame}
+
+\begin{frame}
+Attack Idea:\\
+\vspace{3mm}
+We have an expression that links plaintext bits to input bits to the 4th round of SPN that holds with high bias ($1/32$). We can partially undo the last round by guessing the last key. \\
+\vspace{3mm}
+\pause
+If we guess correctly, the equation will hold with high bias. If we guess wrong, the equation will probabily hold with probability close to 1/2, that is, a bias close to 0.\\
+\vspace{3mm}
+BUT To do this, we don't need to guess the entire key for the last round! Our expression only involves 4 key bits, output from 2 S-Boxes. Thus we only need to guess $2^8 = 256$ values.\\
+\vspace{3mm}
+For each value of the guessed \textit{target partial subkey} we can undo the last round and determine the bias of the equation. Highest bias indicates likely correct guess.
 \end{frame}

### 2009-01-30 talk/draft1.tex

@@
 \begin{frame}
-How to exploit those - Differential attack idea
+\begin{definition}[Differential]
+Suppose two plaintext inputs to the system be $X$ and $X'$ with corresponding output ciphertexts $Y$ and $Y'$ respectively.
+Then the pair of input difference ($\Delta X = X \oplus X'$) and the output difference ($\Delta Y = Y \oplus Y'$) is called a {\it differential} for the system.
+\end{definition}
 
+Vulnerability of DES:\\
+Differential Cryptanalysis exploits the high probability of certain occurrences of differential patterns in the cipher. In DES, this highly probable differentials occur due to
+\begin{itemize}
+\item{S-Boxes show a strong tendency to produce certain differential pairs with high probability, rather than being random.}
+\item{The differences do not get diffused fast enough.}
+\item{Differentials are not affected by the round keys.}
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+In an $k$-round DES, we can trace the path of a certain input difference to get an output difference with a high probability. This tracing path is known as a {\it differential characteristic}.
+
+\[ \Delta P \rightarrow \Delta X_1 \rightarrow \Delta X_2 \rightarrow \cdots \rightarrow \Delta X_{k-1} \rightarrow \Delta C \]
+
+In an ideal situation, one will expect the probability
+\[ P \left(\Delta C | \Delta P \right) = \frac{1}{2^n} \]
+for an $n$-bit cipher system. Differential attacks seek to exploit a scenario where
+\[ P \left(\Delta C | \Delta P \right) = p_D \gg \frac{1}{2^n} \]
+
+This is essentially a chosen-plaintext attack as we want the specific input difference to occur for every pair.
 \end{frame}
@@
 \begin{frame}
-Mathematical description of the attacks
-
+\begin{beamercolorbox}[ht=2.5ex,dp=1.125ex,center,rounded=true,shadow=true]{author in head/foot}
+Dive into the Mathematics
+\end{beamercolorbox}
 \end{frame}
 
-\subsection{Simplification WLOG}
+\subsection{Substitution-Permutation Network}
+
 \begin{frame}
-Reducing DES down to SPN
+\begin{columns}
+\begin{column}{0.6\textwidth}
+\begin{figure}
+\includegraphics[totalheight=0.5\textheight]{des_feistel.jpg}
+\end{figure}
+Let us simplify DES by
+\begin{itemize}
+\item{Reducing the bitlength from $64$ to $16$}
+\item{Removing expansion function $E(R)$}
+\item{Taking identical S-Boxes of size $4$ bits}
+\item{Repeating over less rounds}
+\end{itemize}
+\end{column}
 
-We can remove much of the complexity of DES without affecting the discussion of the techniques of Linear Cryptanalysis.
+\begin{column}{0.4\textwidth}
+\begin{figure}
+\includegraphics[totalheight=0.8\textheight]{spn.jpg}
+\end{figure}
+\end{column}
+\end{columns}
+
+\begin{center}
+{\tiny Disclaimer: This simplification does not affect the discussion of the techniques of Linear/Differential Cryptanalysis.}
+\end{center}
+\end{frame}
+
+\begin{frame}
+\begin{beamercolorbox}[ht=2.5ex,dp=1.125ex,center,rounded=true,shadow=true]{author in head/foot}
+Simple Substitution-Permutation Network
+\end{beamercolorbox}
+\end{frame}
+
+\begin{frame}
 We define a simple Substitution-Permutation Network Cipher (SPN) which has structural similarity to DES.
@@
 \end{itemize}
-
-\end{frame}
-
-\begin{frame}
-SPN Visually
-%\pgfimage[scale=0.75]{SPN.PNG}
 \end{frame}
@@
 SPN uses a single 4-bit S-Box that has the following structure:
-\begin{tabular}{|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|}
-\hline
-Input: & 0 & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & A & B & C & D & E & F \\
-\hline
-Output:& E & 4 & D & 1 & 2 & F & B & 8 & 3 & A & 6 & C & 5 & 9 & 0 & 7 \\
-\hline
-\end{tabular}\\
+\begin{figure}
+\includegraphics[width=0.9\textwidth]{spn_sbox.jpg}
+\end{figure}
+
 \vspace{5mm}
 And the following 16-bit permutation:
-\begin{tabular}{|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|}
-\hline
-1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 & 15 & 16\\
-\hline
-1 & 5 & 9 & 13 & 2 & 6 & 10 & 14 & 3 & 7 & 11 & 15 & 4 & 8 & 12 & 16 \\
-\hline
-\end{tabular}
+\begin{figure}
+\includegraphics[width=0.9\textwidth]{spn_perm.jpg}
+\end{figure}
+
 \vspace{5mm}
@@
 \subsection{Linear Attack on SPN}
+\begin{frame}
+\begin{beamercolorbox}[ht=2.5ex,dp=1.125ex,center,rounded=true,shadow=true]{author in head/foot}
+Mathematics of Linear Cryptanalysis
+\end{beamercolorbox}
+\end{frame}
+
 \begin{frame}
@@
 \vspace{5mm}
-\tiny
-\begin{tabular}{|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|}
-\hline
- &0&1&2&3&4&5&6&7&8&9&A&B&C&D&E&F\\
-\hline
-0&+8&0&0&0&0&0&0&0&0&0&0&0&0&0&0&0 \\
-\hline
-1&0&0&-2&-2&0&0&-2&\textbf{+6}&+2&+2&0&0&+2&+2&0&0 \\
-\hline
-2&0&0&-2&-2&0&0&-2&-2&0&0&+2&+2&0&0&\textbf{-6}&+2 \\
-\hline
-3&0&0&0&0&0&0&0&0&+2&\textbf{-6}&-2&-2&+2&+2&-2&-2 \\
-\hline
-4&0&+2&0&-2&-2&-4&-2&0&0&-2&0&+2&+2&-4&+2&0 \\
-\hline
-5&0&-2&-2&0&-2&0&+4&+2&-2&0&-4&+2&0&-2&-2&0  \\
-\hline
-6&0&+2&-2&+4&+2&0&0&+2&0&-2&+2&+4&-2&0&0&-2 \\
-\hline
-7&0&-2&0&+2&+2&-4&+2&0&-2&0&+2&0&+4&+2&0&+2  \\
-\hline
-8&0&0&0&0&0&0&0&0&-2&+2&+2&-2&+2&-2&-2&\textbf{-6} \\
-\hline
-9&0&0&-2&-2&0&0&-2&-2&-4&0&-2&+2&0&+4&+2&-2 \\
-\hline
-A&0&+4&-2&+2&-4&0&+2&-2&+2&+2&0&0&+2&+2&0&0 \\
-\hline
-B&0&+4&0&-4&+4&0&+4&0&0&0&0&0&0&0&0&0 \\
-\hline
-C&0&-2&+4&-2&-2&0&+2&0&+2&0&+2&+4&0&+2&0&-2 \\ 
-\hline
-D&0&+2&+2&0&-2&+4&0&+2&-4&-2&+2&0&+2&0&0&+2 \\
-\hline
-E&0&+2&+2&0&-2&-4&0&+2&-2&0&0&-2&-4&+2&-2&0 \\
-\hline
-F&0&-2&-4&-2&-2&0&+2&0&0&-2&+4&-2&-2&0&+2&0 \\
-\hline
-\end{tabular}
-\small
+\begin{figure}
+\includegraphics[height=0.7\textheight]{spn_linear_approx.jpg}
+\end{figure}
 \end{frame}

### 2009-01-30 trunk/talk/draft1.tex
SPN Linear slides DONE

@@
 
-\subsection{Simplification WLOG}
+\subsection{Substitution-Permutation Network}
 
@@
 \begin{figure}
-\includegraphics[totalheight=0.55\textheight]{des_feistel.jpg}
+\includegraphics[totalheight=0.5\textheight]{des_feistel.jpg}
 \end{figure}
@@
 \begin{figure}
-\includegraphics[totalheight=0.9\textheight]{spn.jpg}
+\includegraphics[totalheight=0.8\textheight]{spn.jpg}
 \end{figure}
@@
 \end{columns}
+
+\begin{center}
+{\tiny Disclaimer: This simplification does not affect the discussion of the techniques of Linear/Differential Cryptanalysis.}
+\end{center}
 \end{frame}
@@
 \begin{frame}
-We can remove much of the complexity of DES without affecting the discussion of the techniques of Linear Cryptanalysis.
 We define a simple Substitution-Permutation Network Cipher (SPN) which has structural similarity to DES.
@@
 \[ 2 * 1/4 * 1/4 = 1/8 \]
+Note the implicit assumption that S-Boxes are independant. More on this later.
 \end{frame}
@@
 \vspace{3mm}
+\pause
 BUT To do this, we don't need to guess the entire key for the last round! Our expression only involves 4 key bits, output from 2 S-Boxes. Thus we only need to guess $2^8 = 256$ values.\\
 \vspace{3mm}
+\pause
 For each value of the guessed \textit{target partial subkey} we can undo the last round and determine the bias of the equation. Highest bias indicates likely correct guess.
@@
 \begin{frame}
-Probabilistic justification ...
+\begin{columns}
+\begin{column}{0.45\textwidth}
+\begin{block}{SPN Linear Approximation}
+\begin{align*}
+ U_{4,6} & \oplus U_{4,8} \oplus U_{4,14} \oplus U_{4, 16} \\
+         & \oplus P_5 \oplus P_7 \oplus P_8 = 0 
+\end{align*}
+\end{block}
+\end{column}
+\begin{column}{0.55\textwidth}
+\begin{figure}
+\includegraphics[totalheight=0.85\textheight]{SPN_linear.PNG}
+\caption{Linear Attack on SPN}
+\end{figure}
+\end{column}
+\end{columns}
+\end{frame}
 
+\begin{frame}
+Probabalistic Justification:\\
+\vspace{5mm}
+In general, the number of plaintext-ciphertext pairs needed is related inversely-quadratically to the bias. That is
+\[ N_{pairs} \approx 1 / \epsilon^2 \]
+For the SPN cipher approximation with bias $1/32$ we need about $1000$ pairs. To perform the attack with near full probability of success.
 \end{frame}

### 2009-01-30 trunk/talk/draft1.tex
Most of DES Linear slides

@@
 \begin{frame}
-Linear Attack idea ...
+\begin{itemize}
+\item{Structurally and cryptographically, DES is similar to SPN.}
+\item{We will adapt the linear attack on SPN to a linear attack on a 3-round DES algorithm and mount this attack.}
+\item{We will also show how to extend this attack to a full 16-round DES algorithm and show it is faster than an exhaustive search.}
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+The first step, as before, is to come up with a linear approximation for the S-Boxes used by DES.\\
+\vspace{5mm}
+Note that the S-Boxes here are 6-bit to 4-bit maps.
+We can use the same technique as before.
+For a given S-Box $a \in \{1, 2, \dots 8\}, 1 \leq \alpha \leq 63, 1 \leq \beta \leq 15$ we define $NS_{a}(\alpha, \beta)$ as the number of times (out of 64) for S-Box $a$ that for all the input patterns masked by $\alpha$ the output pattern masked by $\beta$ agrees with the value of S-Box $a$.
+\[ NS_{a}(\alpha, \beta) = | \{ X | 0 \leq X < 64, ( \bigoplus_{i=0}^5 X[i]\cdot\alpha[i]) = (\bigoplus_{j=0}^3 S_a(X)[j] \cdot \beta[j])\} | \]
+This looks more complicated, but is infact the same expression as for SPN, except it takes into account all possible S-Boxes.
+\end{frame}
+
+\begin{frame}
+Using $NS$ to find a linear approximation for the S-Boxes.\\
+\vspace{5mm}
+With 6-bit S-Boxes things are not as easy as with 4-bit S-Boxes.
+\begin{figure}
+\includegraphics[totalheight=0.5\textheight]{NS_S5.PNG}
+\caption{Bias for S-Box 5 }
+\end{figure}
+It turns out that the highest bias is for S-Box 5 with $\alpha = 16$ and $\beta = 15$
+\end{frame}
+
+\begin{frame}
+From previous result, the S-Box approximation with highest bias for S-Box 5. Let $X$ be the input bits to the S-Box and let $Y$ be the output bits. Then:
+\[ X_{2} = Y_{1} \oplus Y_{2} \oplus Y_{3} \oplus Y_{4} \]
+Holds with bias of absolute value $\sim 0.31$\\
+\vspace{5mm}
+Tracing through the Feistel function to the right input bits $R$, round key bits $K$, and Feistel function output bits $F$ we get
+\[ R_{17} \oplus F_{25} \oplus F_{14} \oplus F_{8} \oplus F_{3} = K_{26} \]
 
@@
 \begin{frame}
-Matsui NS for SBoxes ...
-
-\end{frame}
-
-\begin{frame}
-Linear approximations for SBoxes ...
-
-\end{frame}
-
-\begin{frame}
-Piling up over 3 rounds ...
-
-\end{frame}
-
-\begin{frame}
-Demo of the attack ...
-
+Let $P_L$ and $P_R$ be the left and right plaintext bits. Let $C_L$ and $C_R$ be the left and right ciphertext bits. Let $K_i$ be the round key for round $i$. \\
+\vspace{5mm}
+Using the S-Box 5 approximation and the Piling Up Principle we can combine the approximations to get
+\begin{align*}
+ P_L{25} & \oplus P_L{14} \oplus P_L{8} \oplus P_L{3} \oplus \\
+         & C_L{25} \oplus C_L{14} \oplus C_L{8} \oplus C_L{3} \oplus \\
+			& P_R{17} \oplus C_R{17} = K_1{26} \oplus K_3{26} \\
+\end{align*}
+The bias magnitude for this expression is for this expression is 
+\[ (12/64)^2 + (1 - 12/64)^2 - 1/2 \approx 0.30 \]
+This is the best expression for the 3-round DES.
 \end{frame}

### 2009-01-30 trunk/src/DesImpl.java
Linear cryptanalysis for 4-round DES

@@ -2,6 +2,8 @@ import java.util.BitSet;
 
 
 public class DesImpl {
+  public static final boolean DES_INITIAL_PERMUTATION = false;
+  
   // S Boxes for DES
   public static final int[][][] S = new int[][][] {
     { { 14,  4, 13,  1,  2, 15, 11,  8,  3, 10,  6, 12,  5,  9,  0,  7 },
@@ -56,6 +58,42 @@ public class DesImpl {
   public static final int[] C_PERM = new int[] {
     16,  7, 20, 21, 29, 12, 28, 17,  1, 15, 23, 26,  5, 18, 31, 10,
     2,   8, 24, 14, 32, 27,  3,  9, 19, 13, 30,  6, 22, 11,  4, 25
+  };
+  
+  // for cryptanalysis
+  public static final int[] C_PERM_INV = new int[] {
+	9,//1
+	17,//2
+	23,//3
+	31,//4
+	13,//5
+	28,//6
+	2,//7
+	18,//8
+	24,//9
+	16,//10
+	30,//11
+	6,//12
+	26,//13
+	20,//14
+	10,//15
+	1,//16
+	8,//17
+	14,//18
+	25,//19
+	3,//20
+	4,//21
+	29,//22
+	11,//23
+	19,//24
+	32,//25
+	12,//26
+	22,//27
+	7,//28
+	5,//29
+	27,//30
+	15,//31
+	21,//32
   };
   
   // PC-1 box for DES Key Schedule
@@ -164,17 +202,26 @@ public class DesImpl {
   
   public BitSet SBox(BitSet ER) {
     // compute results of s-boxes
-    BitSet C = new BitSet(48);
+    BitSet C = new BitSet(32);
     for (int i = 0; i < 8; i++) {
-      
-      int row = (ER.get(0 + (6 * i)) ? 2 : 0) + (ER.get(5 + (6 * i)) ? 1 : 0);
-      int col = (ER.get(1 + (6 * i)) ? 8 : 0) + (ER.get(2 + (6 * i)) ? 4 : 0)
-              + (ER.get(3 + (6 * 1)) ? 2 : 0) + (ER.get(4 + (6 * i)) ? 1 : 0);
-      int sval = S[i][row][col];
-      for (int j = 0; j < 4; j++) {
-        C.set(i * 4 + j, (((sval >> (3 - j)) & 1) != 0 )); 
-      }
[…]

### 2009-01-30 trunk/src/LinearCryptanalysis.java

@@ -1,13 +1,24 @@
+import java.util.ArrayList;
 import java.util.BitSet;
+import java.util.List;
+import java.util.Random;
 
 
 public class LinearCryptanalysis {
+
+
+  public static void main(String[] args) {
+	  DesLinearAttack4Round();
+	  //PrintNSForAll();
+	  //MatsuiEqn5Distribution();
+	  //testFinalRound();
+  }
   
   // Prints the Linear Approximation for all Sboxes
   /**
  * @param args
  */
-public static void main(String[] args) {
+  public static void PrintNSForAll() {
     int max = 0;
     int maxAlpha = -1;
     int maxBeta = -1;
@@ -96,5 +107,282 @@ public static void main(String[] args) {
     
     return count;
   }
+  
+  public static void MatsuiEqn5Distribution() {
+	  BitSet k1 = Util.toBitSet(0xFEEEED, 24);
+	  BitSet k2 = Util.toBitSet(0x123456, 24);
+	  BitSet K = Util.concatenate(k1, 24, k2, 24);
+	  
+	  // generate random X-input values
+	  Random r = new Random();
+	  DesImpl des = new DesImpl();
+	  
+	  int numXs = 10000;
+	  int count = 0;
+	  for (int i = 0; i < numXs; i++) {
+	      BitSet x1 = Util.toBitSet(r.nextInt(65536), 16);
+	      BitSet x2 = Util.toBitSet(r.nextInt(65536), 16);
+	      BitSet X = Util.concatenate(x1, 16, x2, 16);
+	      
+	      BitSet F = des.Feistel(X, K);
+	      
+	      // now compute eqn 5
+	      //  X[15] + F(X,K)[7, 18, 24, 29]  = K[22] 
+	      //  probability of holding should be ~ 0.19
+	      boolean lhs = X.get(31 - 15) ^ F.get(31 - 7) ^ F.get(31 - 18) ^ F.get(31 - 24) ^ F.get(31 - 29);
+	      if (lhs == false) count++;
+	  }
+	  
+	  System.out.println("Eqn5 frequency of holding:");
+	  System.out.println(count / ((double)numXs));
+	  
+  }
+  
+  public static void testFinalRound() {
+	  BitSet k1 = Util.toBitSet(0xFEEEED, 24);
+	  BitSet k2 = Util.toBitSet(0x123456, 24);
+	  BitSet K = Util.concatenate(k1, 24, k2, 24);
+	  
+      BitSet r1 = Util.toBitSet(0xDEAD, 16);
[…]

### 2009-01-30 trunk/src/SpnLinearCryptanalysis.java

@@ -56,7 +56,7 @@ public class SpnLinearCryptanalysis {
         Util.toBitSet(0xDEAD, 16),
         Util.toBitSet(0x5678, 16),
         Util.toBitSet(0xACE1, 16),
-        Util.toBitSet(0xF2E4, 16),
+        Util.toBitSet(0xFEED, 16),
     };
     
     // 1 - generate 10000 pairs

### 2009-01-30 trunk/talk/draft1.tex
DES Linear slides complete

@@
 
-\subsection{Linear Attack: 3-round DES}
+\subsection{Linear Attack: 4-round DES}
 \begin{frame}
@@
 This is the best expression for the 3-round DES.
+\end{frame}
+
+\begin{frame}
+We perform the attack on the 4-round DES similiarly to how we performed the attack on 4-round SPN, by "undoing" the last round.\\
+\vspace{5mm}
+\pause
+Note that in the last round, $C_L$ bits are not changed, only $C_R$ bits are affected. Note also that there is only one $C_R$ bit the 3-round equation. I
+\vspace{5mm}
+\pause
+This bit, $C_R{17}$ is affected by the result of S-Box 1 in round 4.
+If we guess correctly for the 6 partial subkey bits $K_{4,1}$ to $K_{4,6}$, the 3-round equation will hold with high bias, otherwise it will likely hold with close to 0 bias.\\
+\vspace{5mm}
+\pause
+Thus, this attack will give us $6$ of the $56$ key bits.
+What this means in terms of time saved over exhaustive search is this: If it takes a year to find a key through exhaustive search of $2^56$ bits, after this attack it will only take $5$ days to exhaustively search the remaining $2^50$ bits.
+\end{frame}
+
+\begin{frame}
+How many pairs do we need? Probabalistic justification:\\
+\vspace{3mm}
+Matsui Lemma 5: Let $N$ be the number of given random plaintexts and $\epsilon$, the bias, be sufficiently small.
+Let $q^{(i)}$ be the probability that the following equation holds for a subkey candidate $K_n^{(i)}$ and a random variable X:
+\[ F_n(X,K_n)[l_1, l_2,\dots,l_d] = F_n(X, K_n^{(i)})[l_1, l_2, \dots, l_d].\]
+Then if $q^{(i)}$'s are independent, the success rate of the attack is
+\[ \int_{x=-2\sqrt{N}\epsilon}^\infty \left( \prod_{K_n^{(i)} \neq K_n} \int_{-x-4\sqrt{N}\epsilon q^{(i)}}^{x+4\sqrt{N}\epsilon (1 - q^{(i)})} \frac{1}{\sqrt{2\pi}} e^{-y^2/2} dy \right) \frac{1}{\sqrt{2\pi}} e^{-x^2/2} dx \]
+Where the product is taken over all subkey candidates except $K_n$.\\
+\vspace{3mm}
+Let $d = 1$ and $l_1 = 17$ in the above expression - this corresponds to our attack (distribution over guessing for one bit, $CR_{17}$ in the equation).
+The following table summarizes the sucess rates:
+\begin{center}
+\begin{tabular}{|c|r|r|r|r|}
+\hline
+N & 2$\epsilon^{-2}$ & $4\epsilon^{-2}$ & $8\epsilon^{-2}$ & $16\epsilon^{-2}$ \\
+\hline
+Success Rate & 0.486 & 0.785 & 0.967 & 0.999 \\
+\hline
+\end{tabular} 
+\end{center}
+\end{frame}
+
+\begin{frame}
+How many pairs do we need? Continued\\
+\vspace{3mm}
+Using the calculation from the previous slide, we see that for a 4-round DES, our equation holds with bias of about 1 in 4.
+Thus we will need about $16\times 4^2 = 256$ pairs for the attack to have near perfect chance of sucess.\\
+\vspace{3mm}
+For a full 16-round DES, the best approximation has bias $\sim 1.19\times 10^{-22}$, thus we would need on the order of $2^{47}$ plaintext-ciphertext pairs.
 \end{frame}

### 2009-01-30 talk/draft1.tex
More spelling mistakes...

@@
 
-\subsection{Linear Attack: 3-round DES}
+\subsection{Linear Attack: 4-round DES}
 \begin{frame}
-Linear Attack idea ...
+\begin{itemize}
+\item{Structurally and cryptographically, DES is similar to SPN.}
+\item{We will adapt the linear attack on SPN to a linear attack on a 3-round DES algorithm and mount this attack.}
+\item{We will also show how to extend this attack to a full 16-round DES algorithm and show it is faster than an exhaustive search.}
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+The first step, as before, is to come up with a linear approximation for the S-Boxes used by DES.\\
+\vspace{5mm}
+Note that the S-Boxes here are 6-bit to 4-bit maps.
+We can use the same technique as before.
+For a given S-Box $a \in \{1, 2, \dots 8\}, 1 \leq \alpha \leq 63, 1 \leq \beta \leq 15$ we define $NS_{a}(\alpha, \beta)$ as the number of times (out of 64) for S-Box $a$ that for all the input patterns masked by $\alpha$ the output pattern masked by $\beta$ agrees with the value of S-Box $a$.
+\[ NS_{a}(\alpha, \beta) = | \{ X | 0 \leq X < 64, ( \bigoplus_{i=0}^5 X[i]\cdot\alpha[i]) = (\bigoplus_{j=0}^3 S_a(X)[j] \cdot \beta[j])\} | \]
+This looks more complicated, but is infact the same expression as for SPN, except it takes into account all possible S-Boxes.
+\end{frame}
+
+\begin{frame}
+Using $NS$ to find a linear approximation for the S-Boxes.\\
+\vspace{5mm}
+With 6-bit S-Boxes things are not as easy as with 4-bit S-Boxes.
+\begin{figure}
+\includegraphics[totalheight=0.5\textheight]{NS_S5.PNG}
+\caption{Bias for S-Box 5 }
+\end{figure}
+It turns out that the highest bias is for S-Box 5 with $\alpha = 16$ and $\beta = 15$
+\end{frame}
+
+\begin{frame}
+From previous result, the S-Box approximation with highest bias for S-Box 5. Let $X$ be the input bits to the S-Box and let $Y$ be the output bits. Then:
+\[ X_{2} = Y_{1} \oplus Y_{2} \oplus Y_{3} \oplus Y_{4} \]
+Holds with bias of absolute value $\sim 0.31$\\
+\vspace{5mm}
+Tracing through the Feistel function to the right input bits $R$, round key bits $K$, and Feistel function output bits $F$ we get
+\[ R_{17} \oplus F_{25} \oplus F_{14} \oplus F_{8} \oplus F_{3} = K_{26} \]
 
@@
 \begin{frame}
-Matsui NS for SBoxes ...
-
+Let $P_L$ and $P_R$ be the left and right plaintext bits. Let $C_L$ and $C_R$ be the left and right ciphertext bits. Let $K_i$ be the round key for round $i$. \\
+\vspace{5mm}
+Using the S-Box 5 approximation and the Piling Up Principle we can combine the approximations to get
+\begin{align*}
+ P_L{25} & \oplus P_L{14} \oplus P_L{8} \oplus P_L{3} \oplus \\
+         & C_L{25} \oplus C_L{14} \oplus C_L{8} \oplus C_L{3} \oplus \\
+			& P_R{17} \oplus C_R{17} = K_1{26} \oplus K_3{26} \\
+\end{align*}
+The bias magnitude for this expression is
+\[ (12/64)^2 + (1 - 12/64)^2 - 1/2 \approx 0.30 \]
+This is the best expression for the 3-round DES.
 \end{frame}
@@
 \begin{frame}
-Linear approximations for SBoxes ...
-
+We perform the attack on the 4-round DES similiarly to how we performed the attack on 4-round SPN, by "undoing" the last round.\\
+\vspace{5mm}
+\pause
+Note that in the last round, $C_L$ bits are not changed, only $C_R$ bits are affected. Note also that there is only one $C_R$ bit the 3-round equation.\\
+\vspace{5mm}
+\pause
+This bit, $C_R{17}$ is affected by the result of S-Box 1 in round 4.
+If we guess correctly for the 6 partial subkey bits $K_{4,1}$ to $K_{4,6}$, the 3-round equation will hold with high bias, otherwise it will likely hold with close to 0 bias.\\
+\vspace{5mm}
+\pause
+Thus, this attack will give us $6$ of the $56$ key bits.
+What this means in terms of time saved over exhaustive search is this: If it takes a year to find a key through exhaustive search of $2^{56}$ bits, after this attack it will only take $5$ days to exhaustively search the remaining $2^{50}$ bits.
 \end{frame}
@@
 \begin{frame}
-Piling up over 3 rounds ...
-
+How many pairs do we need? Probabalistic justification:\\
+\vspace{3mm}
+Matsui Lemma 5: Let $N$ be the number of given random plaintexts and $\epsilon$, the bias, be sufficiently small.
+Let $q^{(i)}$ be the probability that the following equation holds for a subkey candidate $K_n^{(i)}$ and a random variable X:
+\[ F_n(X,K_n)[l_1, l_2,\dots,l_d] = F_n(X, K_n^{(i)})[l_1, l_2, \dots, l_d].\]
+Then if $q^{(i)}$'s are independent, the success rate of the attack is
+\[ \int_{x=-2\sqrt{N}\epsilon}^\infty \left( \prod_{K_n^{(i)} \neq K_n} \int_{-x-4\sqrt{N}\epsilon q^{(i)}}^{x+4\sqrt{N}\epsilon (1 - q^{(i)})} \frac{1}{\sqrt{2\pi}} e^{-y^2/2} dy \right) \frac{1}{\sqrt{2\pi}} e^{-x^2/2} dx \]
+Where the product is taken over all subkey candidates except $K_n$.\\
+\vspace{3mm}
+Let $d = 1$ and $l_1 = 17$ in the above expression - this corresponds to our attack (distribution over guessing for one bit, $CR_{17}$ in the equation).
+The following table summarizes the sucess rates:
+\begin{center}
+\begin{tabular}{|c|r|r|r|r|}
+\hline
+N & 2$\epsilon^{-2}$ & $4\epsilon^{-2}$ & $8\epsilon^{-2}$ & $16\epsilon^{-2}$ \\
+\hline
+Success Rate & 0.486 & 0.785 & 0.967 & 0.999 \\
+\hline
+\end{tabular} 
+\end{center}
 \end{frame}
@@
 \begin{frame}
-Demo of the attack ...
-
+How many pairs do we need? Continued\\
+\vspace{3mm}
+Using the calculation from the previous slide, we see that for a 4-round DES, our equation holds with bias of about 1 in 4.
+Thus we will need about $16\times 4^2 = 256$ pairs for the attack to have near perfect chance of sucess.\\
+\vspace{3mm}
+For a full 16-round DES, the best approximation has bias $\sim 1.19\times 10^{-22}$, thus we would need on the order of $2^{47}$ plaintext-ciphertext pairs.
 \end{frame}

### 2009-01-30 trunk/talk/draft1.tex
Spelling fixes and etc.
Starting conclusion

@@
 \end{align*}
-The bias magnitude for this expression is for this expression is 
+The bias magnitude for this expression is
 \[ (12/64)^2 + (1 - 12/64)^2 - 1/2 \approx 0.30 \]
@@
 \pause
-Note that in the last round, $C_L$ bits are not changed, only $C_R$ bits are affected. Note also that there is only one $C_R$ bit the 3-round equation. I
+Note that in the last round, $C_L$ bits are not changed, only $C_R$ bits are affected. Note also that there is only one $C_R$ bit the 3-round equation.\\
 \vspace{5mm}
@@
 Thus, this attack will give us $6$ of the $56$ key bits.
-What this means in terms of time saved over exhaustive search is this: If it takes a year to find a key through exhaustive search of $2^56$ bits, after this attack it will only take $5$ days to exhaustively search the remaining $2^50$ bits.
+What this means in terms of time saved over exhaustive search is this: If it takes a year to find a key through exhaustive search of $2^{56}$ bits, after this attack it will only take $5$ days to exhaustively search the remaining $2^{50}$ bits.
 \end{frame}
@@
 \begin{frame}
-Refresh math idea ... Independence assumption ...
+Note that throughout this discussion we have assumed that the S-Boxes were independant. This allowed us to use the Piling Up Principle to combine linear approximations of S-Boxes across rounds. \\
+\vspace{3mm}
+This assumption worked well for us in practice, but is not necessarily true. We could have proceeded differently without this assumption - [see John Manferdelli's Boolean Functions slides]
 

### 2009-01-30 trunk/talk/draft1.tex
SPN Diff slides

--- 2009-01-30 trunk/talk/draft1.tex
@@
 Vulnerability of DES:\\
-Differential Cryptanalysis exploits the high probability of certain occurrences of differential patterns in the cipher. In DES, this highly probable differentials occur due to
 \begin{itemize}
@@
 
-\[ \Delta P \rightarrow \Delta X_1 \rightarrow \Delta X_2 \rightarrow \cdots \rightarrow \Delta X_{k-1} \rightarrow \Delta C \]
+\[ \Delta P \rightarrow \Delta C^{(1)} \rightarrow \Delta C^{(2)} \rightarrow \cdots \rightarrow \Delta C^{(k-1)} \rightarrow \Delta C \]
 
@@
 \begin{frame}
-Basic math concepts behind Differential attack ...
+Differential Cryptanalysis tries to exploit the high probability of certain occurrences of differential characteristics $(\Delta P \rightarrow \Delta C)$ in the cipher.
+This is a chosen ciphertext attack.
+
+[Differential: $(\Delta P, \Delta C) = (P \oplus P', C \oplus C') $]
+
+We have to analyze the cipher in pieces. The basic idea is
+\begin{itemize}
+\item{Analyze the difference pattern and their probabilities of the S-Box}
+\item{Combine the S-Boxes over the rounds to build the differential characteristic of the cipher}
+\item{Choose the characteristic with maximum probability and use that to retrieve subkey bits of the cipher}
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+Analyzing the S-Box
+
+Let us consider the S-Box of the SPN cipher as we constructed it:
+\begin{columns}
+\begin{column}{0.5\textwidth}
+\begin{itemize}
+\item{$4 \times 4$ S-Box}
+\item{Input: $X = [X_1 \: X_2 \: X_3 \: X_4]$}
+\item{Output: $Y = [Y_1 \: Y_2 \: Y_3 \: Y_4]$}
+\item{Difference pair: $(\Delta X, \Delta Y)$}
+\end{itemize}
+\end{column}
+\begin{column}{0.5\textwidth}
+\begin{figure}
+\includegraphics[totalheight=0.3\textheight]{spn_sbox_unkeyed.jpg}
+\end{figure}\end{column}
+\end{columns}
+
+Algorithm
+\begin{enumerate}
+\item{Fix a certain input difference $\Delta X = 0100$, say}
+\item{Iterate through the first input $X = \{0000, ... ,1111 \}$}
+\item{This fixes the second input $X' = \{0100, ... ,1011\}$ (ordering does not matter)}
+\item{Obtain corresponding output difference $\Delta Y = Y \oplus Y'$}
+\item{Iterate through steps 2 to 4 for $\Delta X = \{0000, ... ,1111 \}$} 
+\end{enumerate}
+\end{frame}
+
+\begin{frame}
+Sample Difference Pairs for the S-Box
+
+\begin{figure}
+\includegraphics[totalheight=0.6\textheight]{spn_sbox_diffpairs.jpg}
+\end{figure}
+
+Distribution of differences
+
+Note that, for $\Delta X = 1011$, $\Delta Y = 0010$ occurs $8$ times, out of the possible $16$ times.
+So, the difference pair $(1011,0010)$ has a probability of occurrence $\frac{8}{16} = \frac{1}{2}$.
+So, we can tabulate the complete data for the S-Box to check this probabilities.
+\end{frame}
+
+\begin{frame}
+Difference Distribution Table
+
+\begin{figure}
+\includegraphics[totalheight=0.6\textheight]{spn_sbox_disttable.jpg}
+\end{figure}
+
+Note that
+\begin{itemize}
+\item{In an ideal S-Box, we would like all entries to be $1$}
+\item{But this table clearly shows bias towards certain pairs}
+\item{Just divide the entries by $2^4 = 16$ to get the probabilities, and exploit the scenario of highest probability of occurrence}
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+Why can not we make an ideal S-Box?
+
+Notice that in the difference table
+\begin{itemize}
+\item{The $(0,0)$ entry is $16$, just because identical inputs ($\Delta X = 0 \Rightarrow X = X'$) should produce identical outputs ($Y = Y' \Rightarrow \Delta Y = 0$)}
+\item{All the entries are even, because $\Delta X$ is the same for both the input pairs $(X,X')$ and $(X',X)$, producing same $\Delta Y$}
+\item{Sum of all entries in a row is $2^4 = 16$}
+\end{itemize}
+
+As an ideal S-Box expects to give out no information of $\Delta Y$ given $\Delta X$, we wish it could have all entries $1$, i.e,
+\[ Pr(\Delta Y | \Delta X) = \frac{1}{16} \quad \forall \Delta X  \]
+But, from our discussion about the distribution table, this is infeasible.
+\end{frame}
+
+\begin{frame}
+What happens for a Keyed S-Box?
+
+Let us consider the S-Box of the SPN cipher, with the round keys this time:
+\begin{columns}
+\begin{column}{0.5\textwidth}
+\begin{itemize}
+\item{$4 \times 4$-bit S-Box}
+\item{Input: $W = [W_1 \: W_2 \: W_3 \: W_4]$}
+\item{Round Key: $K = [K_1 \: K_2 \: K_3 \: K_4]$}
+\item{Output: $Y = [Y_1 \: Y_2 \: Y_3 \: Y_4]$}
+\item{Difference pair: $(\Delta W, \Delta Y)$}
+\end{itemize}
+\end{column}
+\begin{column}{0.5\textwidth}
+\begin{figure}
+\includegraphics[totalheight=0.3\textheight]{spn_sbox_keyed.jpg}
+\end{figure}
+\end{column}
+\end{columns}
+
+Note that for each pair of inputs $(W,W')$, the actual inputs to the S-Box are $(X,X') = (W \oplus K, W' \oplus K)$. Hence, the input difference for the keyed S-Box is 
+\[ \Delta W = W \oplus W' = (X \oplus K) \oplus (X' \oplus K) = X \oplus X' = \Delta X \]
+and hence $\Delta W \rightarrow \Delta Y$ corresponds identically to $\Delta X \rightarrow \Delta Y$.
+
+Thus, the keyed S-Box has the same difference distribution table as the unkeyed S-Box.
+\end{frame}
+
+\begin{frame}
+Constructing a Differential Characteristic
+
+\begin{columns}
+\begin{column}{0.6\textwidth}
+
+The diagram beside illustrates the tracing of the non-zero bits of the input difference through the SPN structure.
+
+Input Diff: $\Delta P = 0B00$ \\
+
+Propagating Difference Pairs
+\begin{itemize}
+\item{$S_{12}$: $1011 \rightarrow 0010$ [$Pr = \frac{8}{16}$]}
+\item{$S_{23}$: $0100 \rightarrow 0110$ [$Pr = \frac{6}{16}$]}
+\item{$S_{32}$: $0010 \rightarrow 0101$ [$Pr = \frac{6}{16}$]}
+\item{$S_{33}$: $0010 \rightarrow 0101$ [$Pr = \frac{6}{16}$]}
+\item{$S_{ij}$: $0000 \rightarrow 0000$ [$Pr = \frac{16}{16}$]}
+\end{itemize}
+
+Differential Characteristic
+\[ 0B00 \rightarrow 0040 \rightarrow 0220 \rightarrow 0606 \]
+\end{column}
+\begin{column}{0.4\textwidth}
+\begin{figure}
+\includegraphics[totalheight=0.8\textheight]{spn_diff_char.jpg}
+\end{figure}
+\end{column}
+\end{columns}
+\end{frame}
+
+\begin{frame}
+Piling up Probabilities
+
+Notation
+$\Delta U_i$: Input difference to $i$-th round
+$\Delta V_i$: Output difference from $i$-th round
+$K_i$: Round Key for $i$-th round
+$X_{i,j}$: $j$-th bit of $X_i$
+
+Here, we have the following in the differential characteristic
+\begin{itemize}
+\item{Round 1: $\Delta P = \Delta U_1 = 0B00 \rightarrow 0040 = \Delta V_1$ [$Pr = \frac{8}{16}$]}
+\item{Round 2: $\Delta V_1 = \Delta U_2 = 0040 \rightarrow 0220 = \Delta V_2$ [$Pr = \frac{6}{16}$]}
+\item{Round 3: $\Delta V_2 = \Delta U_3 = 0220 \rightarrow 0606 = \Delta V_3$ [$Pr = \left(\frac{6}{16}\right)^2$]}
+\item{Round 4: $\Delta V_3 = \Delta U_4 = 0606$}
+\end{itemize}
+
+Hence, assuming the S-Boxes to be independent over the rounds, we obtain
+\[ \Delta P = 0B00 \rightarrow 0606 = \Delta U_4 \]
+with probability $Pr(\Delta U_4 = 0606 | \Delta P = 0B00) = \frac{8}{16} \times \frac{6}{16} \times \left(\frac{6}{16}\right)^2 = \frac{27}{1024}$.
+
+Note: The probability of this characteristic $Pr = \frac{27}{2^{10}} \gg \frac{1}{2^{16}}$
+
+Now, lets see if we can exploit this situation to extract the key bits.
+\end{frame}
+
+\begin{frame}
+Attack Idea
+\begin{itemize}
+\item{We have a differential characteristic that links plaintext difference $\Delta P$ to input difference $\Delta U_4$ of the 4th round output of SPN and holds with high probability ($\frac{27}{1024}$)}
+\item{We can partially undo the last round by guessing the last key $K_5$}
+\item{If we guess correctly, the differential characteristic will hold with high probability, close to $\frac{27}{1024}$. If we guess wrong, it will hold with low probability}
+\item{But in this case, we don't need to guess the entire key $K_5$ for the last round!}
+\item{The differential output only involves 2 S-Boxes, $S_{42}$ and $S_{44}$. Thus we only need to guess $2^8 = 256$ values}
+\item{For each value of the guessed \textit{target partial subkey} we can undo the last round and determine the probability of the differential characteristic}
+\item{Highest probability indicates likely correct guess}
+\end{itemize}
 
@@
 \begin{frame}
-Intro to differentials ...
+\begin{columns}
+\begin{column}{0.6\textwidth}
+Differential Characteristic flow
 
-\end{frame}
+Input Diff: $\Delta P = 0B00$
 
-\begin{frame}
-Differential characteristic / approximation of SBox ...
+Characteristic
+\[ \Delta P = 0B00 \rightarrow 0606 = \Delta U_4 \]
 
-\end{frame}
+Probability $Pr(\Delta U_4 = 0606 | \Delta P = 0B00) = \frac{27}{1024}$
 
-\begin{frame}
-Probabilities for the approximations ...
+Partial Subkey guess only for
+\[ [K_{5,5} ... K_{5,8}, K_{5,13} ... K_{5,16}] \]
+\end{column}
 
-\end{frame}
-
-\begin{frame}
-Analogue of Piling Up lemma ...
-
-\end{frame}
-
-\begin{frame}
-Combining across SPN rounds ...
-
-\end{frame}
-
-\begin{frame}
-The algorithm for the attack ...
-
-\end{frame}
-
-\begin{frame}
-Probabilistic justification ...
+\begin{column}{0.4\textwidth}
+\begin{figure}
+\includegraphics[totalheight=0.8\textheight]{spn_diff_char.jpg}
+\end{figure}
+\end{column}
+\end{columns}
 

### 2009-01-30 trunk/talk/draft1.tex
Fixed r21 break

@@
 \begin{frame}
-Refresh math idea ... Independence assumption ...
-
+Note that throughout this discussion we have assumed that the S-Boxes were independant. This allowed us to use the Piling Up Principle to combine linear approximations of S-Boxes across rounds. \\
+\vspace{3mm}
+This assumption worked well for us in practice, but is not necessarily true. We could have proceeded differently without this assumption - [see John Manferdelli's Boolean Functions slides]
 \end{frame}

### 2009-01-30 src/DesImpl.java
Readability

@@ -1,374 +1,405 @@
 import java.util.BitSet;
 
-
+/**
+ * Implementation of variable-round DES
+ * 
+ * @author Slava Chernyak and Sourav Sen Gupta
+ * 
+ */
 public class DesImpl {
-  public static final boolean DES_INITIAL_PERMUTATION = false;
-  
-  // S Boxes for DES
-  public static final int[][][] S = new int[][][] {
-    { { 14,  4, 13,  1,  2, 15, 11,  8,  3, 10,  6, 12,  5,  9,  0,  7 },
-      {  0, 15,  7,  4, 14,  2, 13,  1, 10,  6, 12, 11,  9,  5,  3,  8 },
-      {  4,  1, 14,  8, 13,  6,  2, 11, 15, 12,  9,  7,  3, 10,  5,  0 },
-      { 15, 12,  8,  2,  4,  9,  1,  7,  5, 11,  3, 14, 10,  0,  6, 13 }
-    },
+	/** Flag determines weather the initial permutation is applied */
+	public static final boolean DES_INITIAL_PERMUTATION = false;
 
-    { { 15,  1,  8, 14,  6, 11,  3,  4,  9,  7,  2, 13, 12,  0,  5, 10 },
-      {  3, 13,  4,  7, 15,  2,  8, 14, 12,  0,  1, 10,  6,  9, 11,  5 },
-      {  0, 14,  7, 11, 10,  4, 13,  1,  5,  8, 12,  6,  9,  3,  2, 15 },
-      { 13,  8, 10,  1,  3, 15,  4,  2, 11,  6,  7, 12,  0,  5, 14,  9 }
-    },
+	/** S-Boxes for DES */
+	public static final int[][][] S = new int[][][] {
+			{ { 14, 4, 13, 1, 2, 15, 11, 8, 3, 10, 6, 12, 5, 9, 0, 7 },
+					{ 0, 15, 7, 4, 14, 2, 13, 1, 10, 6, 12, 11, 9, 5, 3, 8 },
+					{ 4, 1, 14, 8, 13, 6, 2, 11, 15, 12, 9, 7, 3, 10, 5, 0 },
+					{ 15, 12, 8, 2, 4, 9, 1, 7, 5, 11, 3, 14, 10, 0, 6, 13 } },
 
-    { { 10,  0,  9, 14,  6,  3, 15,  5,  1, 13, 12,  7, 11,  4,  2,  8 },
[…]

### 2009-01-30 src/SpnDifferentialCryptanalysis.java

@@ -3,7 +3,10 @@ import java.util.BitSet;
 import java.util.List;
 import java.util.Random;
 
-
+/**
+ * Tools for Differential Cryptanalysis of SPN
+ * @author Sourav Sen Gupta
+ */
 public class SpnDifferentialCryptanalysis {
   
   /**

### 2009-01-30 src/SpnImpl.java

@@ -1,126 +1,149 @@
 import java.util.BitSet;
 
-
+/**
+ * Implementation of a simple SPN cipher
+ * 
+ * @author Slava Chernyak
+ */
 public class SpnImpl {
-  
-  public static int[] SBOX = new int[]{ 0xE, 0x4, 0xD, 0x1, 0x2, 0xF, 0xB, 0x8, 0x3, 0xA, 0x6, 0xC, 0x5, 0x9, 0x0, 0x7 };
-  public static int[] SBOX_INV = new int[]{ 0xE, 0x3, 0x4, 0x8, 0x1, 0xC, 0xA, 0xF, 0x7, 0xD, 0x9, 0x6, 0xB, 0x2, 0x0, 0x5}; 
-  
-  /** Note that this permutation is an involution */
-  public static int[] PERM = new int[]{ 1, 5, 9, 13, 2, 6, 10, 14, 3, 7, 11, 15, 4, 8, 12, 16};
-  
-  /**
-   * Perform S-BOX substitution on bit set s
-   * @param s a 4-bit bit set
-   * @return the result of the S-Box
-   */
-  public BitSet sBoxFunc(BitSet s, boolean inv) {
-    int idx = Util.toInteger(s, 4);
-    int result;
-    if (inv) {
-      result = SBOX_INV[idx];
-    } else {
-      result = SBOX[idx];     
-    }
-    return Util.toBitSet(result, 4);
-  }
-  
-  /**
-   * Perform the permutation of block s
-   * @param s 16 bit block to permute
-   * @return result of permutation
-   */
-  public BitSet Permute(BitSet s) {
-    BitSet fromPerm = new BitSet(16);
-    for (int i = 0; i < 16; i++) {
-      fromPerm.set(PERM[i] - 1, s.get(i));
-    }
-    return fromPerm;
-  }
-  
-  /**
-   * Performs all the s-box substitutions on block s
-   * @param s 16-bit block
-   * @return the result of the substitutions 
-   */
-  public BitSet sBoxes(BitSet s, boolean inv) {
-    BitSet result = new BitSet(16);
-    for (int i = 0; i < 4; i++) {
-      // compute each sbox
-      BitSet toS = new BitSet(4);
-      for (int j = 0; j < 4; j++) {
-        toS.set(j, s.get(i * 4 + j));        
-      }
-      BitSet fromS = sBoxFunc(toS, inv);
-      for (int j = 0; j < 4; j++) {
[…]

### 2009-01-30 src/SpnLinearCryptanalysis.java

@@ -3,141 +3,153 @@ import java.util.BitSet;
 import java.util.List;
 import java.util.Random;
 
-
+/**
+ * Tools for linear cryptanalysis of SPN
+ * 
+ * @author Slava Chernyak
+ */
 public class SpnLinearCryptanalysis {
-  
-  /**
-   * Generate random plain/ciphertext pairs via the SPN cipher with the given
-   * round keys
-   * @param pairs Number of pairs to generate
-   * @param roundKeys Round keys to use for the SPN cipher
-   * @return List of plain/cipher sets (each set is 16 bits long)
-   */
-  public List<BitSet[]> generateRandomPairs(int pairs, BitSet[] roundKeys) {
-    Random r = new Random();
-    SpnImpl spn = new SpnImpl();
-    List<BitSet[]> sets = new ArrayList<BitSet[]>();
-    for (int i = 0; i < pairs; i++) {
-      BitSet p = Util.toBitSet(r.nextInt(65536), 16);
-      BitSet c = spn.SpnEncBlock(p, roundKeys);
-      sets.add(new BitSet[]{p, c});
-    }
-    return sets;
-  }
-  
-  /**
-   * Backtrack across one s-box with a partial ciphertext and partial subkey guess
-   * @param partialc 4 bit partial of the ciphertext
-   * @param partialk 4 bit partial of the subkey
-   * @return
-   */
-  public BitSet backtrackPartialSubkeyGuess(BitSet partialc, BitSet partialk) {
-    partialc = Util.copyBitSet(partialc, 4);
-    partialc.xor(partialk);
-    SpnImpl spn = new SpnImpl();
-    BitSet U = spn.sBoxFunc(partialc, true);
-    return U;
-  }
-  
-  /**
-   * Mount the linear cryptanalysis attack on the SPN cipher to recover 8 bits
-   * of the final round key
-   */
-  public static void main(String[] args) {
-    System.out.println("Running linear cryptanalysis on SPN....");
-    // the expression to compute:
-    //  U_4,6 + U_4,8 + U_4,14 + U_4,16 + P_5 + P_7 + P_8 = 0
-    
-    // 0 - set the keys to be recovered - the partial 
-    // subkey bits that will be found are *2*4 in the last 
-    // round.
-    BitSet[] roundKeys = new BitSet[]{
-        Util.toBitSet(0x1234, 16),
[…]

### 2009-01-30 src/Util.java

@@ -1,68 +1,138 @@
 import java.util.BitSet;
 
+/**
+ * Static utilities for BitSet class
+ * 
+ * @author Slava Chernyak
+ * 
+ */
+public final class Util {
+	private Util() {
+	}
 
-public class Util {
-  
-  public static BitSet increment(BitSet set, int totalBits) {
-    boolean carry = true;
-    int place = totalBits - 1;
-    while (carry && place >= 0) {
-      carry = set.get(place);
-      set.set(place, ! set.get(place));
-      place--;
-    }
-    return set;
-  }
-  
-  public static boolean DirectSum(BitSet set, int totalBits) {
-    boolean acc = false;
-    for (int i = 0; i < totalBits; i++) {
-      acc ^= set.get(i);
-    }
-    return acc;
-  }
-  
-  public static int toInteger(BitSet set, int totalBits) {
-    int acc = 0;
-    for (int i = 0; i < totalBits; i++) {
-      acc += ((set.get(totalBits - i - 1)) ? 1 : 0) << i; 
-    }
-    return acc;
-  }
-  
-  public static BitSet toBitSet(int value, int totalBits) {
-    int idx = totalBits - 1;
-    BitSet result = new BitSet();
-    while (value > 0) {
-      result.set(idx--, value % 2 == 1);
-      value /= 2;
-    }
-    return result;
-  }
-  
-  public static boolean equalsBitSet(BitSet a, BitSet b, int totalBits) {
-    for(int i = 0; i < totalBits; i++) {
-      boolean ca = a.get(i);
-      boolean cb = b.get(i);
-      if (ca != cb) return false;
-    }
-    return true;
-  }
-  
-  public static BitSet copyBitSet(BitSet set, int totalBits) {
-    BitSet newset = new BitSet();
-    for (int i = 0; i < totalBits; i++) {
-      newset.set(i, set.get(i));
-    }
-    return newset;
-    
-  }
-  
-  public static BitSet concatenate(BitSet a, int totalBitsA, BitSet b, int totalBitsB) {
-    for(int i = 0; i < totalBitsB; i++) {
-      a.set(totalBitsA + i, b.get(i));
-    }
-    return a;
-  }
+	/**
+	 * Increment the numeric value represented by this bit set
+	 * 
+	 * @param set
+	 *          to increment
+	 * @param totalBits
+	 *          number of bits in the bit set
+	 * @return bit set with incremented value
+	 */
[…]

### 2009-01-30 trunk/src/LinearCryptanalysis.java
Readability

@@ -3,386 +3,394 @@ import java.util.BitSet;
 import java.util.List;
 import java.util.Random;
 
-
+/**
+ * Tools for linear cryptanalysis of DES
+ * 
+ * @author Slava Chernyak
+ */
 public class LinearCryptanalysis {
 
+	/**
+	 * Run any of the following commented out blocks
+	 * 
+	 * @param args
+	 */
+	public static void main(String[] args) {
+		DesLinearAttack4Round();
+		// PrintNSForAll();
+		// MatsuiEqn5Distribution();
+		// testFinalRound();
+	}
 
-  public static void main(String[] args) {
-	  DesLinearAttack4Round();
-	  //PrintNSForAll();
-	  //MatsuiEqn5Distribution();
-	  //testFinalRound();
-  }
-  
-  // Prints the Linear Approximation for all Sboxes
-  /**
- * @param args
- */
-  public static void PrintNSForAll() {
-    int max = 0;
-    int maxAlpha = -1;
-    int maxBeta = -1;
-    
-    for (int sbox = 0 ; sbox < 8; sbox++) {
-      System.out.println("SBOX: " + (sbox + 1));
-      
-      
-      BitSet alpha = new BitSet(6);
-      BitSet beta = new BitSet(4);
-      
-      for (int i = 1; i < 16; i++) {
-        System.out.print("\t" + i);
-      }
-      System.out.println();
-      
-      // Reset the max initial value for each SBox
-      max = 0;
-      
-      Util.increment(alpha, 6);
-      for (int i = 1; i < 64; i++) {
-        System.out.print(i + ":\t");
-        Util.increment(beta, 4);
-        for (int j = 1; j < 16; j++) {
-          int ns = MatsuiNS(sbox, alpha, beta);
-          int delta = ns - 32;
-          
-          if (Math.abs(delta) > max) {
-            max = Math.abs(delta);
-            maxAlpha = Util.toInteger(alpha, 6);
-            maxBeta = Util.toInteger(beta, 4);
-          }          
-          Util.increment(beta, 4);          
-          System.out.print(delta + "\t");          
-        }        
-        Util.increment(alpha, 6);        
-        System.out.println();        
-      }
-      System.out.println();
-      
[…]

### 2009-01-30 trunk/src/SpnDifferentialCryptanalysis.java
Readability some more...

@@ -5,15 +5,19 @@ import java.util.Random;
 
 /**
  * Tools for Differential Cryptanalysis of SPN
+ * 
  * @author Sourav Sen Gupta
  */
 public class SpnDifferentialCryptanalysis {
-  
+
   /**
    * Generate random plain/ciphertext pairs via the SPN cipher with the given
    * round keys
-   * @param pairs Number of pairs to generate
-   * @param roundKeys Round keys to use for the SPN cipher
+   * 
+   * @param pairs
+   *          Number of pairs to generate
+   * @param roundKeys
+   *          Round keys to use for the SPN cipher
    * @return List of plain/cipher sets (each set is 16 bits long)
    */
   public List<BitSet[]> generateChosenPairs(int pairs, BitSet[] roundKeys) {
@@ -21,31 +25,36 @@ public class SpnDifferentialCryptanalysis {
     SpnImpl spn = new SpnImpl();
     List<BitSet[]> sets = new ArrayList<BitSet[]>();
     for (int i = 0; i < pairs; i++) {
-    	// choose a random plaintext-ciphertext pair
-    	BitSet p1 = Util.toBitSet(r.nextInt(65536), 16);
-    	BitSet c1 = spn.SpnEncBlock(p1, roundKeys);
-    	
-    	// choose another pair such that dP = [0000 1011 0000 0000] is satisfied
-    	BitSet p2 = Util.copyBitSet(p1, 16);
-    	p2.xor(Util.toBitSet(2816, 16));
-    	BitSet c2 = spn.SpnEncBlock(p2, roundKeys);
-    	
-    	// discard the 'wrong pairs'
-    	// 'right pairs' should have dC = [0000 **** 0000 ****]
-    	BitSet dC = Util.copyBitSet(c1, 16);
-    	dC.xor(c2);
-    	
-    	if ((Util.toInteger(dC.get(0,4),4)==0) && (Util.toInteger(dC.get(8,12),4)==0)) {
-    		sets.add(new BitSet[]{c1, c2});
-    	}
+      // choose a random plaintext-ciphertext pair
+      BitSet p1 = Util.toBitSet(r.nextInt(65536), 16);
+      BitSet c1 = spn.SpnEncBlock(p1, roundKeys);
+
+      // choose another pair such that dP = [0000 1011 0000 0000] is satisfied
+      BitSet p2 = Util.copyBitSet(p1, 16);
[…]

### 2009-01-30 trunk/src/SpnImpl.java

@@ -7,143 +7,143 @@ import java.util.BitSet;
  */
 public class SpnImpl {
 
-	/** SPN S-Box */
-	public static int[] SBOX = new int[] { 0xE, 0x4, 0xD, 0x1, 0x2, 0xF, 0xB,
-			0x8, 0x3, 0xA, 0x6, 0xC, 0x5, 0x9, 0x0, 0x7 };
-	/** SPN Inverted S-Box */
-	public static int[] SBOX_INV = new int[] { 0xE, 0x3, 0x4, 0x8, 0x1, 0xC, 0xA,
-			0xF, 0x7, 0xD, 0x9, 0x6, 0xB, 0x2, 0x0, 0x5 };
+  /** SPN S-Box */
+  public static int[] SBOX = new int[] { 0xE, 0x4, 0xD, 0x1, 0x2, 0xF, 0xB,
+      0x8, 0x3, 0xA, 0x6, 0xC, 0x5, 0x9, 0x0, 0x7 };
+  /** SPN Inverted S-Box */
+  public static int[] SBOX_INV = new int[] { 0xE, 0x3, 0x4, 0x8, 0x1, 0xC, 0xA,
+      0xF, 0x7, 0xD, 0x9, 0x6, 0xB, 0x2, 0x0, 0x5 };
 
-	/** SPN Permutation - note that this permutation is an involution */
-	public static int[] PERM = new int[] { 1, 5, 9, 13, 2, 6, 10, 14, 3, 7, 11,
-			15, 4, 8, 12, 16 };
+  /** SPN Permutation - note that this permutation is an involution */
+  public static int[] PERM = new int[] { 1, 5, 9, 13, 2, 6, 10, 14, 3, 7, 11,
+      15, 4, 8, 12, 16 };
 
-	/**
-	 * Perform S-BOX substitution on bit set s
-	 * 
-	 * @param s
-	 *          a 4-bit bit set
-	 * @return the result of the S-Box
-	 */
-	public BitSet sBoxFunc(BitSet s, boolean inv) {
-		int idx = Util.toInteger(s, 4);
-		int result;
-		if (inv) {
-			result = SBOX_INV[idx];
-		} else {
-			result = SBOX[idx];
-		}
-		return Util.toBitSet(result, 4);
-	}
+  /**
+   * Perform S-BOX substitution on bit set s
+   * 
+   * @param s
+   *          a 4-bit bit set
+   * @return the result of the S-Box
+   */
[…]

### 2009-01-30 trunk/src/SpnLinearCryptanalysis.java

@@ -10,146 +10,146 @@ import java.util.Random;
  */
 public class SpnLinearCryptanalysis {
 
-	/**
-	 * Generate random plain/ciphertext pairs via the SPN cipher with the given
-	 * round keys
-	 * 
-	 * @param pairs
-	 *          Number of pairs to generate
-	 * @param roundKeys
-	 *          Round keys to use for the SPN cipher
-	 * @return List of plain/cipher sets (each set is 16 bits long)
-	 */
-	public List<BitSet[]> generateRandomPairs(int pairs, BitSet[] roundKeys) {
-		Random r = new Random();
-		SpnImpl spn = new SpnImpl();
-		List<BitSet[]> sets = new ArrayList<BitSet[]>();
-		for (int i = 0; i < pairs; i++) {
-			BitSet p = Util.toBitSet(r.nextInt(65536), 16);
-			BitSet c = spn.SpnEncBlock(p, roundKeys);
-			sets.add(new BitSet[] { p, c });
-		}
-		return sets;
-	}
+  /**
+   * Generate random plain/ciphertext pairs via the SPN cipher with the given
+   * round keys
+   * 
+   * @param pairs
+   *          Number of pairs to generate
+   * @param roundKeys
+   *          Round keys to use for the SPN cipher
+   * @return List of plain/cipher sets (each set is 16 bits long)
+   */
+  public List<BitSet[]> generateRandomPairs(int pairs, BitSet[] roundKeys) {
+    Random r = new Random();
+    SpnImpl spn = new SpnImpl();
+    List<BitSet[]> sets = new ArrayList<BitSet[]>();
+    for (int i = 0; i < pairs; i++) {
+      BitSet p = Util.toBitSet(r.nextInt(65536), 16);
+      BitSet c = spn.SpnEncBlock(p, roundKeys);
+      sets.add(new BitSet[] { p, c });
+    }
+    return sets;
+  }
 
-	/**
-	 * Backtrack across one s-box with a partial ciphertext and partial subkey
-	 * guess
-	 * 
-	 * @param partialc
-	 *          4 bit partial of the ciphertext
-	 * @param partialk
-	 *          4 bit partial of the subkey
-	 * @return
-	 */
-	public BitSet backtrackPartialSubkeyGuess(BitSet partialc, BitSet partialk) {
-		partialc = Util.copyBitSet(partialc, 4);
-		partialc.xor(partialk);
-		SpnImpl spn = new SpnImpl();
[…]

### 2009-01-30 src/LinearCryptanalysis.java
Readability some more...

@@ -10,387 +10,387 @@ import java.util.Random;
  */
 public class LinearCryptanalysis {
 
-	/**
-	 * Run any of the following commented out blocks
-	 * 
-	 * @param args
-	 */
-	public static void main(String[] args) {
-		DesLinearAttack4Round();
-		// PrintNSForAll();
-		// MatsuiEqn5Distribution();
-		// testFinalRound();
-	}
+  /**
+   * Run any of the following commented out blocks
+   * 
+   * @param args
+   */
+  public static void main(String[] args) {
+    DesLinearAttack4Round();
+    // PrintNSForAll();
+    // MatsuiEqn5Distribution();
+    // testFinalRound();
+  }
 
-	/**
-	 * Print the matsui NS values for all DES sboxes
-	 */
-	public static void PrintNSForAll() {
-		int max = 0;
-		int maxAlpha = -1;
-		int maxBeta = -1;
+  /**
+   * Print the matsui NS values for all DES sboxes
+   */
+  public static void PrintNSForAll() {
+    int max = 0;
+    int maxAlpha = -1;
+    int maxBeta = -1;
 
-		for (int sbox = 0; sbox < 8; sbox++) {
-			System.out.println("SBOX: " + (sbox + 1));
+    for (int sbox = 0; sbox < 8; sbox++) {
+      System.out.println("SBOX: " + (sbox + 1));
 
-			BitSet alpha = new BitSet(6);
-			BitSet beta = new BitSet(4);
+      BitSet alpha = new BitSet(6);
+      BitSet beta = new BitSet(4);
 
-			for (int i = 1; i < 16; i++) {
-				System.out.print("\t" + i);
-			}
-			System.out.println();
+      for (int i = 1; i < 16; i++) {
+        System.out.print("\t" + i);
+      }
+      System.out.println();
 
-			// Reset the max initial value for each SBox
-			max = 0;
+      // Reset the max initial value for each SBox
+      max = 0;
 
-			Util.increment(alpha, 6);
-			for (int i = 1; i < 64; i++) {
-				System.out.print(i + ":\t");
-				Util.increment(beta, 4);
-				for (int j = 1; j < 16; j++) {
-					int ns = MatsuiNS(sbox, alpha, beta);
-					int delta = ns - 32;
+      Util.increment(alpha, 6);
+      for (int i = 1; i < 64; i++) {
+        System.out.print(i + ":\t");
[…]

### 2009-01-30 src/Util.java

@@ -7,132 +7,132 @@ import java.util.BitSet;
  * 
  */
 public final class Util {
-	private Util() {
-	}
+  private Util() {
+  }
 
-	/**
-	 * Increment the numeric value represented by this bit set
-	 * 
-	 * @param set
-	 *          to increment
-	 * @param totalBits
-	 *          number of bits in the bit set
-	 * @return bit set with incremented value
-	 */
-	public static BitSet increment(BitSet set, int totalBits) {
-		boolean carry = true;
-		int place = totalBits - 1;
-		while (carry && place >= 0) {
-			carry = set.get(place);
-			set.set(place, !set.get(place));
-			place--;
-		}
-		return set;
-	}
+  /**
+   * Increment the numeric value represented by this bit set
+   * 
+   * @param set
+   *          to increment
+   * @param totalBits
+   *          number of bits in the bit set
+   * @return bit set with incremented value
+   */
+  public static BitSet increment(BitSet set, int totalBits) {
+    boolean carry = true;
+    int place = totalBits - 1;
+    while (carry && place >= 0) {
+      carry = set.get(place);
+      set.set(place, !set.get(place));
+      place--;
+    }
+    return set;
+  }
 
-	/**
-	 * Compute the direct sum (XOR) of all the bits in the set
-	 * 
-	 * @param set
-	 *          to compute direct sum on
-	 * @param totalBits
-	 *          number of bits in the bit set
-	 * @return value of direct sum
-	 */
-	public static boolean DirectSum(BitSet set, int totalBits) {
-		boolean acc = false;
-		for (int i = 0; i < totalBits; i++) {
-			acc ^= set.get(i);
-		}
-		return acc;
-	}
+  /**
+   * Compute the direct sum (XOR) of all the bits in the set
+   * 
+   * @param set
+   *          to compute direct sum on
+   * @param totalBits
+   *          number of bits in the bit set
+   * @return value of direct sum
+   */
+  public static boolean DirectSum(BitSet set, int totalBits) {
+    boolean acc = false;
+    for (int i = 0; i < totalBits; i++) {
+      acc ^= set.get(i);
+    }
+    return acc;
+  }
 
-	/**
-	 * Convert bit set to integer value (low numbered bits highiest order)
-	 * 
-	 * @param set
-	 *          to compute integer value of
-	 * @param totalBits
[…]

### 2009-01-30 src/DesImpl.java

@@ -7,399 +7,426 @@ import java.util.BitSet;
  * 
  */
 public class DesImpl {
-	/** Flag determines weather the initial permutation is applied */
-	public static final boolean DES_INITIAL_PERMUTATION = false;
+  /** Flag determines weather the initial permutation is applied */
+  public static final boolean DES_INITIAL_PERMUTATION = false;
 
-	/** S-Boxes for DES */
-	public static final int[][][] S = new int[][][] {
-			{ { 14, 4, 13, 1, 2, 15, 11, 8, 3, 10, 6, 12, 5, 9, 0, 7 },
-					{ 0, 15, 7, 4, 14, 2, 13, 1, 10, 6, 12, 11, 9, 5, 3, 8 },
-					{ 4, 1, 14, 8, 13, 6, 2, 11, 15, 12, 9, 7, 3, 10, 5, 0 },
-					{ 15, 12, 8, 2, 4, 9, 1, 7, 5, 11, 3, 14, 10, 0, 6, 13 } },
+  /** S-Boxes for DES */
+  public static final int[][][] S = new int[][][] {
+      { { 14, 4, 13, 1, 2, 15, 11, 8, 3, 10, 6, 12, 5, 9, 0, 7 },
+          { 0, 15, 7, 4, 14, 2, 13, 1, 10, 6, 12, 11, 9, 5, 3, 8 },
+          { 4, 1, 14, 8, 13, 6, 2, 11, 15, 12, 9, 7, 3, 10, 5, 0 },
+          { 15, 12, 8, 2, 4, 9, 1, 7, 5, 11, 3, 14, 10, 0, 6, 13 } },
 
-			{ { 15, 1, 8, 14, 6, 11, 3, 4, 9, 7, 2, 13, 12, 0, 5, 10 },
-					{ 3, 13, 4, 7, 15, 2, 8, 14, 12, 0, 1, 10, 6, 9, 11, 5 },
-					{ 0, 14, 7, 11, 10, 4, 13, 1, 5, 8, 12, 6, 9, 3, 2, 15 },
-					{ 13, 8, 10, 1, 3, 15, 4, 2, 11, 6, 7, 12, 0, 5, 14, 9 } },
+      { { 15, 1, 8, 14, 6, 11, 3, 4, 9, 7, 2, 13, 12, 0, 5, 10 },
+          { 3, 13, 4, 7, 15, 2, 8, 14, 12, 0, 1, 10, 6, 9, 11, 5 },
[…]

### 2009-01-30 trunk/talk/draft1.tex
Fixing more r21 stuff

@@
 \end{align*}
-The bias magnitude for this expression is for this expression is 
+The bias magnitude for this expression is
 \[ (12/64)^2 + (1 - 12/64)^2 - 1/2 \approx 0.30 \]
 This is the best expression for the 3-round DES.
+\end{frame}
+
+\begin{frame}
+We perform the attack on the 4-round DES similiarly to how we performed the attack on 4-round SPN, by "undoing" the last round.\\
+\vspace{5mm}
+\pause
+Note that in the last round, $C_L$ bits are not changed, only $C_R$ bits are affected. Note also that there is only one $C_R$ bit the 3-round equation.\\
+\vspace{5mm}
+\pause
+This bit, $C_R{17}$ is affected by the result of S-Box 1 in round 4.
+If we guess correctly for the 6 partial subkey bits $K_{4,1}$ to $K_{4,6}$, the 3-round equation will hold with high bias, otherwise it will likely hold with close to 0 bias.\\
+\vspace{5mm}
+\pause
+Thus, this attack will give us $6$ of the $56$ key bits.
+What this means in terms of time saved over exhaustive search is this: If it takes a year to find a key through exhaustive search of $2^{56}$ bits, after this attack it will only take $5$ days to exhaustively search the remaining $2^{50}$ bits.
+\end{frame}
+
+\begin{frame}
+How many pairs do we need? Probabalistic justification:\\
+\vspace{3mm}
+Matsui Lemma 5: Let $N$ be the number of given random plaintexts and $\epsilon$, the bias, be sufficiently small.
+Let $q^{(i)}$ be the probability that the following equation holds for a subkey candidate $K_n^{(i)}$ and a random variable X:
+\[ F_n(X,K_n)[l_1, l_2,\dots,l_d] = F_n(X, K_n^{(i)})[l_1, l_2, \dots, l_d].\]
+Then if $q^{(i)}$'s are independent, the success rate of the attack is
+\[ \int_{x=-2\sqrt{N}\epsilon}^\infty \left( \prod_{K_n^{(i)} \neq K_n} \int_{-x-4\sqrt{N}\epsilon q^{(i)}}^{x+4\sqrt{N}\epsilon (1 - q^{(i)})} \frac{1}{\sqrt{2\pi}} e^{-y^2/2} dy \right) \frac{1}{\sqrt{2\pi}} e^{-x^2/2} dx \]
+Where the product is taken over all subkey candidates except $K_n$.\\
+\vspace{3mm}
+Let $d = 1$ and $l_1 = 17$ in the above expression - this corresponds to our attack (distribution over guessing for one bit, $CR_{17}$ in the equation).
+The following table summarizes the sucess rates:
+\begin{center}
+\begin{tabular}{|c|r|r|r|r|}
+\hline
+N & 2$\epsilon^{-2}$ & $4\epsilon^{-2}$ & $8\epsilon^{-2}$ & $16\epsilon^{-2}$ \\
+\hline
+Success Rate & 0.486 & 0.785 & 0.967 & 0.999 \\
+\hline
+\end{tabular} 
+\end{center}
+\end{frame}
+
+\begin{frame}
+How many pairs do we need? Continued\\
+\vspace{3mm}
+Using the calculation from the previous slide, we see that for a 4-round DES, our equation holds with bias of about 1 in 4.
+Thus we will need about $16\times 4^2 = 256$ pairs for the attack to have near perfect chance of sucess.\\
+\vspace{3mm}
+For a full 16-round DES, the best approximation has bias $\sim 1.19\times 10^{-22}$, thus we would need on the order of $2^{47}$ plaintext-ciphertext pairs.
 \end{frame}

### 2009-01-30 talk/draft1.tex

@@
 
-\subsection{Differential Attack: 3-round DES}
+\subsection{Differential Attack: 4-round DES}
 \begin{frame}
-Differential Attack idea ...
+Differential Attack Idea
+
+Similar to the differential attack on SPN, here we care about the differential characteristics of the cipher. So, the basic idea is as follows
+
+\begin{itemize}
+\item{Construct the difference tables for the S-Boxes}
+\item{Find the most probable difference pairs}
+\item{Combine the pairs through all the rounds to get a higly probable differential characteristic until the penultimate round}
+\item{Exploit this differential characteristic by chosing a number of pairs of pliantexts with the desired difference and undo the last round by guessing the keybits}
+\item{If the initial choice of the input difference was smart, we will only need to guess a few keybits}
+\item{The guess approving the differential characteristic with highest probability is correct}
+\end{itemize}
 
@@
 \begin{frame}
-Differentials for SBoxes ...
+The chosen input difference:$\Delta P = 20 00 00 00 00 00 00 00$
 
+Round 1 Differential
+
+\[ \Delta P = 20 00 00 00 00 00 00 00 \rightarrow 20 00 00 00 00 00 00 00 = \Delta V_1 \]
+Probability: $p_1 = 1$ as Feistel gives output difference $0$ for input $0$
+
+Round 2 Differential
+
+\[ \Delta U_2 = \Delta V_1 = 20 00 00 00 00 00 00 00 \rightarrow \]
+
+this calculations and coding has not been completed yet ... will be done ...
 \end{frame}
@@
 \begin{frame}
-Differential characteristics for SBoxes ...
-
-\end{frame}
-
-\begin{frame}
-Piling up over 3 rounds ...
+Differentials for SBoxes
 

### 2009-01-30 trunk/src/DifferentialCryptanalysis.java

@@ -1,15 +1,24 @@
+import java.util.ArrayList;
 import java.util.BitSet;
+import java.util.List;
+import java.util.Random;
 
 
 public class DifferentialCryptanalysis {
   
+  public static void main(String[] args) {
+	  DesDifferentialAttack4Round();
+	  //PrintNSforAll();
+  }	
+	
+
   // Prints the Differential Tables for all Sboxes
   /**
- * @param args
- */
-public static void main(String[] args) {
-    
-    for (int sbox = 0 ; sbox < 8; sbox++) {
+  * @param args
+  */
+  public static void PrintNSforAll(){
+	
+	for (int sbox = 0 ; sbox < 8; sbox++) {
       System.out.println("SBOX: " + (sbox + 1));
       
       
@@ -80,4 +89,161 @@ public static void main(String[] args) {
     return count;
   }
 
+  
+  
+  
+  
+  /**
+   * Generate random plain/ciphertext pairs via the DES cipher with the given
+   * key and number of rounds
+   * @param pairs Number of pairs to generate
+   * @param key the DES key
+   * @param numRounds the number of DES rounds
+   * @return List of plain/cipher sets (each set is 16 bits long)
+   */
+  public static List<BitSet[]> generateRandomPairs(int pairs, BitSet key, int numRounds) {
+    Random r = new Random();
+    DesImpl des = new DesImpl();
+    List<BitSet[]> sets = new ArrayList<BitSet[]>();
+    for (int i = 0; i < pairs; i++) {
+      // four bloks of 16 bits each
+      BitSet p1 = Util.toBitSet(r.nextInt(65536), 16);
+      BitSet p2 = Util.toBitSet(r.nextInt(65536), 16);
+      BitSet p3 = Util.toBitSet(r.nextInt(65536), 16);
+      BitSet p4 = Util.toBitSet(r.nextInt(65536), 16);
+      BitSet p12 = Util.concatenate(p1, 16, p2, 16);
+      BitSet p34 = Util.concatenate(p3, 16, p4, 16);
+      
+      //first pair
+      BitSet p_one = Util.concatenate(p12, 32, p34, 32);
+      BitSet c_one = des.DesEncBlock(p_one, key, numRounds);
+      
+      //delta_p = 00 80 82 00 60 00 00 00
+      BitSet delta_p1 = Util.toBitSet(0x20000000,32);
[…]

### 2009-01-31 talk/draft1.tex
TODO fixes ... a bunch of typo corrections and formatting changes ...

@@
 \begin{block}{DES Function}
-\[ DES : (\mbox{Plaintext }(P), \mbox{ Key }(K))  \mapsto \mbox{Cipher }(C) \]
+DES : (Plaintext ($P$), Key ($K$)) $\mapsto$ Cipher ($C$)
 \end{block}
@@
 \end{column}
+
 \begin{column}{0.4\textwidth}
@@
 
+\vspace{0.2in}
 \begin{itemize}
 \item{Right half of the plaintext ($R$)}
-\item{Expansion function \\ $E : (0,1)^{32} \rightarrow (0,1)^{48}$}
+\item{Expansion function \\ $\qquad E : (0,1)^{32} \rightarrow (0,1)^{48}$}
 \item{Key for the round ($K$)}
-\item{Confusion function (S-Boxes) \\ $S_i: (0,1)^6 \rightarrow (0,1)^4$}
-\item{Diffusion function \\ $P : (0,1)^{32} \rightarrow (0,1)^{32}$}
+\item{Confusion function (S-Boxes) \\ $\qquad S_i: (0,1)^6 \rightarrow (0,1)^4$}
+\item{Diffusion function \\ $\qquad P : (0,1)^{32} \rightarrow (0,1)^{32}$}
 \end{itemize}
 \end{column}
+
 \begin{column}{0.55\textwidth}
@@
 Mathematically, the different pieces of DES are:
+
 \begin{itemize}
-\item{Initial Permutation $IP : (0,1)^{64} \rightarrow (0,1)^{64}$}
-\item{DES Unit Block functions $\sigma_{i+1} (L_i,R_i)$ for $i = 0,1,...,15$}
-\item{Transpositions $\tau : (L_{i+1},R_{i+1}) \mapsto (R_{i+1},L_{i+1})$ for $i = 0,1,...,14$}
-\item{Inverse Initial Permutation $IP^{-1} : (0,1)^{64} \rightarrow (0,1)^{64}$}
+\item{Initial Permutation \\ $\qquad IP : (0,1)^{64} \rightarrow (0,1)^{64}$}
+\item{DES Unit Block functions \\ $\qquad \sigma_{i+1} (L_i,R_i)$ for $i = 0,1,...,15$}
+\item{Transpositions \\ $\qquad \tau : (L_{i+1},R_{i+1}) \mapsto (R_{i+1},L_{i+1})$ for $i = 0,1,...,14$}
+\item{Inverse Initial Permutation \\ $\qquad IP^{-1} : (0,1)^{64} \rightarrow (0,1)^{64}$}
 \end{itemize}
@@
 \begin{block}{Algebraic Representation of DES}
-\begin{align*}
-C &= DES_K (P) = IP^{-1} \sigma_{16} \tau \cdots \tau \sigma_1 IP (P) \\
-P &= DES^{-1}_K (C) = IP^{-1} \sigma_{1} \tau \cdots \tau \sigma_{16} IP (C)
-\end{align*}
+$\qquad C = DES_K (P) = IP^{-1} \sigma_{16} \tau \cdots \tau \sigma_1 IP (P)$
+
+$\qquad P = DES^{-1}_K (C) = IP^{-1} \sigma_{1} \tau \cdots \tau \sigma_{16} IP (C)$
 \end{block}
 
-Note: The inverse DES comes from the fact that $\tau^2 = \sigma_i^2 = 1$
+\vspace{0.2in}
+{\small Note: The inverse DES comes from the fact that $\tau^2 = \sigma_i^2 = 1$}
 \end{frame}
@@
 \begin{frame}
-Linear Attack on DES idea:\\
+Linear Attack on DES idea:
+
 \begin{itemize}
-\item S-Boxes depend on a relatively small number of bits (6) which allows us to write down linear (or afine) expressions that approximate S-Boxes \\
-\item The effects of one round do not diffuse quickly over following rounds. Thus linear expresions (as above) that hold per-round can be combined across rounds.
+\item S-Boxes depend on a relatively small number of bits (6) which allows us to write down linear (or affine) expressions that approximate S-Boxes
+\item The effects of one round do not diffuse quickly over following rounds. Thus linear or affine expressions (as above) that hold per-round can be combined across rounds
 \end{itemize}
@@
 
-Vulnerability of DES:\\
+\vspace{0.2in}
+Vulnerability of DES:
+
 \begin{itemize}
-\item{S-Boxes show a strong tendency to produce certain differential pairs with high probability, rather than being random.}
-\item{The differences do not get diffused fast enough.}
-\item{Differentials are not affected by the round keys.}
+\item{S-Boxes show a strong tendency to produce certain differential pairs with high probability, rather than being random}
+\item{The differences do not get diffused fast enough through the permutations}
+\item{Differentials are not affected by the round keys as they get xor-ed out}
 \end{itemize}
@@
 \begin{itemize}
-\item SPN is a 16-bit block cipher (16-bit plaintext to 16-bit ciphertext) with 16-bit round keys.
+\item SPN is a 16-bit block cipher (16-bit plaintext to 16-bit ciphertext) with 16-bit round keys
 \item Each round consists of a substitution and a permutation (much like in DES)
-\item The substitution is the result of splitting the 16-bit input block to a round in to 4 4-bit sub-blocks.
-Each 4-bit sub-block is then mapped across a 4-bit to 4-bit S-box.
-We use the same S-box throughout.
-\item The permutation is applied to all 16-bits of the round following the substitution.
+\item The substitution is the result of splitting the 16-bit input block to a round in to 4$\times$4-bit sub-blocks.
+Each 4-bit sub-block is then mapped across a 4-bit to 4-bit S-Box.
+We use the same S-Box throughout
+\item The permutation is applied to all 16-bits of the round following the substitution
 \item Finally, key-mixing is achieved by XOR-ing the round key with every input block to a round as well as at the end of the last round (so that we can't simply ignore the last round)
-\item The SPN cipher we will look at will consist of 4 rounds.
+\item The SPN cipher we will look at will consist of 4 rounds
 \end{itemize}
@@
 \begin{frame}
-SPN S-Box and Permutation:\\
+SPN S-Box and Permutation
+
 \vspace{5mm}
@@
 \vspace{5mm}
+The S-Box provides the {\it confusion} function and the permutation applies the {\it confusion} operation in SPN, thus making it cryptographically similar to DES.
+\vspace{5mm}
 
@@
 \begin{frame}
-\large{Basic Definitions for the Linear attack:}\\
+Basic Definitions for the Linear attack:
+
 \vspace{5 mm}
-Linear cryptanalysis tries to take advantage of high probability occurrences of linear expressions involving plaintext bits, ciphertext bits and subkey bits. It is a known plaintext attack.\\
+Linear cryptanalysis tries to take advantage of high probability occurrences of linear expressions involving plaintext bits, ciphertext bits and subkey bits. It is a {\it known plaintext attack}.
+
 \vspace{5 mm}
@@
 \begin{frame}
-Linear and Affine approximation of S-Box:\\
+Linear and Affine approximation of S-Box:
+
 \vspace{5mm}
-How do we come up with the desired expression for the entire cipher? We start by looking at the only non-linear component, the S-Box.\\
+\begin{itemize}
+\item Question: How do we come up with the desired expression for the entire cipher? 
+\item Hint: We start by looking at the only non-linear component, the S-Box.
+\end{itemize}
+
 \vspace{5mm}
-To find the linear/afine approximation of the S-Box we simply consider every possible expression of the input bits $X_i$ and output bits $Y_j$. Thus the expression has the form
+To find the linear/affine approximation of the S-Box we simply consider every possible expression of the input bits $X_i$ and output bits $Y_j$. Thus the expression has the form
 \[ \bigoplus_{i \in U} X_i = \bigoplus_{j \in V} Y_j \]
-Where $U$ and $V$ range over all possible subsets of $\{1, 2, 3 ,4\}$. We then compare how often this expression coincides with the S-Box. \\
+Where $U$ and $V$ range over all possible subsets of $\{1, 2, 3 ,4\}$. We then compare how often this expression coincides with the S-Box.
+
 \vspace{5mm}
-Note that there are 16 possibilities for $U$ and $V$, hence 256 total possible expressions. (this is for a 4-bit S-box)
+Note that there are 16 possibilities for $U$ and $V$, hence 256 total possible expressions (for a 4-bit S-box).
 \end{frame}
@@
 \begin{frame}
-S-Box approximatin example:\\
+S-Box Approximation Example
+
 \vspace{5mm}
+\begin{columns}
+\begin{column}{0.7\textwidth}
 We can take the following expression as an example:
 \[ X_2 \oplus X_3 = Y_1 \oplus Y_3 \oplus Y_4 \]
-Applying all possible values for the input $X$ bits it turns out that the expression holds in 12 out of the 16 cases. Hence, this expression has a bias of $12/16 - 1/2 = 1/4$.\\
+Applying all possible values for the input $X$ bits it turns out that the expression holds in 12 out of the 16 cases. Hence, this expression has a bias of $12/16 - 1/2 = 1/4$.
+\end{column}
+
+\begin{column}{0.3\textwidth}
+\begin{figure}
+\includegraphics[width=0.7\textwidth]{spn_sbox_unkeyed.jpg}
+\end{figure}
+\end{column}
+\end{columns}
+
 \vspace{5mm}
@@
 \begin{frame}
-The number of agreements (minus 8) between the S-Box and every possible expression is summarized in the table below. Thus to get the bias, one must only divide by 16.\\
+The number of agreements (minus 8) between the S-Box and every possible expression is summarized in the table below. Thus to get the bias, one must only divide by 16.
+
 \vspace{5mm}
@@
 \begin{frame}
-What this means for 1 Round:\\
+\begin{columns}
+\begin{column}{0.4\textwidth}
+What this means for 1 Round:
+
 \vspace{5mm}
-Note, from the previous table, that the expression $X_1 \oplus X_3 \oplus X_4 = Y_2$ has a bias of $+1/4$. \\
+Note, from the previous table, that the expression \[ X_1 \oplus X_3 \oplus X_4 = Y_2 \] has a bias of $+1/4$.
+
 \vspace{5mm}
 Note also that $U_1 = P \oplus K_1$. 
-\vspace{5mm} We can now write down the following linear approximation across the 1st round of SPN:
+\end{column}
+\begin{column}{0.6\textwidth}
+\begin{figure}
+\includegraphics[height=0.45\textheight]{spn_single_round.jpg}
+\end{figure}
+\end{column}
+\end{columns}
+
+\vspace{5mm} 
+We can now write down the following linear approximation across the 1st round of SPN:
 \begin{eqnarray*}
@@
 \end{eqnarray*}
+
 This expression holds with probability of $3/4$ (bias of $+1/4$)
@@
 We can get expressions that hold with some non-$1/2$ probability for every round.
-But we must somehow combine them to write an expression relating the plaintext and ciphertext bits $\Rightarrow$ We use the Piling Up Principle:\\
+But we must somehow combine them to write an expression relating the plaintext and ciphertext bits.
+We use the Piling Up Principle for this purpose.
+
 \vspace{5mm}
-Note: \\
-$X_1 \oplus X_2 = 0$ is a linear expression equivalent to $X_1 = X_2$\\
-$X_1 \oplus X_2 = 1$ is an affine expression equivalent to $X_1 \neq X_2$\\
+Note:
+
+$\qquad X_1 \oplus X_2 = 0$ is a linear expression equivalent to $X_1 = X_2$\\
+$\qquad X_1 \oplus X_2 = 1$ is an affine expression equivalent to $X_1 \neq X_2$
+
 \vspace{5mm}
@@
 \begin{frame}
-Piling Up Principle Continued:\\
+Piling Up Principle (continued):
+
 \vspace{5mm}
-Assuming that $X_1$ and $X_2$ are independant:
+Assuming that $X_1$ and $X_2$ are independent, we get
 \[ Pr(X_1 = i, X_2 = j)  = \left\{ \begin{array}{ll}
@@
 			(1 - p_1)(1 - p_2) & \mbox{$i = 1, j = 1$}.\end{array} \right. \] 
-Now, note that $X_1 \oplus X_2 = 0 \Rightarrow X_1 = 1, X_2 =1$ OR $X_1 = 0, X_2 = 0$. Hence:
+Now, note that $X_1 \oplus X_2 = 0 \Rightarrow X_1 = 1, X_2 =1$ OR $X_1 = 0, X_2 = 0$. Hence,
 \begin{eqnarray*}
@@
 \begin{frame}
-Piling Up Principle Continued - The Piling Up Lemma:\\
+Piling Up Principle (continued) - The Piling Up Lemma:\\
 \vspace{5mm}
@@
 \vspace{5mm}
-Note that if some $\epsilon_k = 0$ (that is, some $X_k$ has no bias) then the total bias will be $0$.  
+Note that if some $\epsilon_k = 0$ (that is, some $X_k$ has no bias) then the total bias will be $0$. More about the probabilistic analysis later.
 \end{frame}
@@
 \begin{frame}
-Now, come back to Per-round approximations. We can write down 4 approximations (Let $S_{i,j}$ repesent the $j$-th S-Box in the $i$-th round):
-\begin{eqnarray*}
-S_{1,2}: X_1 \oplus X_3 \oplus X_4 & = & Y_2 \\
-S_{2,2}: X_2 = Y_2 \oplus Y_4 \\
-S_{3,2}: X_2 = Y_2 \oplus Y_4 \\
-S_{3,4}: X_2 = Y_2 \oplus Y_4 \\
-\end{eqnarray*}
+
+\begin{columns}
+\begin{column}{0.6\textwidth}
+Now, come back to per-round approximations. We can write down 4 approximations ($S_{ij}$ repesents the $j$-th S-Box in the $i$-th round):
+\begin{align*}
+& S_{12}: X_1 \oplus X_3 \oplus X_4 = Y_2 \\
+& S_{22}: X_2 = Y_2 \oplus Y_4 \\
+& S_{32}: X_2 = Y_2 \oplus Y_4 \\
+& S_{34}: X_2 = Y_2 \oplus Y_4 \\
+\end{align*}
 Each of these has a probability bias magnitude of $1/4$. We can use the Piling Up Principle to combine them into a single expression relating plaintext bits to cyphertext bits.
+\end{column}
+
+\begin{column}{0.4\textwidth}
+\begin{figure}
+\includegraphics[totalheight=0.8\textheight]{spn.jpg}
+\end{figure}
+\end{column}
+\end{columns}
+
 \end{frame}
@@
 \begin{frame}
-Consider the first 2 rounds:\\
-\vspace{5mm}
-\[  X_1 \oplus X_3 \oplus X_4 = Y_2 \Rightarrow \]
-\[ V_{1,6} = (P_5\oplus K_{1,5}) \oplus (P_7\oplus K_{1,7}) \oplus (P_8\oplus K_{1,8}) \]
+\begin{columns}
+\begin{column}{0.6\textwidth}
+Consider the first 2 rounds:
+\begin{align*}
+& X_1 \oplus X_3 \oplus X_4 = Y_2 \\ 
+\Rightarrow &(P_5\oplus K_{1,5}) \oplus (P_7\oplus K_{1,7}) \oplus (P_8\oplus K_{1,8}) =  V_{1,6}
+\end{align*}
+\[ X_2 = Y_2 \oplus Y_4 \: \Rightarrow \: (V_{1,6} \oplus K_{2,6}) =  V_{2,6} \oplus V_{2,8}\]
+
 \vspace{2mm}
-\[ X_2 = Y_2 \oplus Y_4 \Rightarrow \]
-\[ V_{2,6} \oplus V_{2,8} = V_{1,6} \oplus K_{2,6} \]
+Each of these has a bias of magnitude $1/4$ and we can combine to obtain:
+\[ V_{2,6} \oplus V_{2,8} \oplus P_5 \oplus P_7 \oplus P_8 \oplus K_{1,5} \oplus K_{1,7} \oplus K_{1,8} \oplus K_{2,6} = 0 \]
+By the Piling Up Principle this holds with bias
+\[ 2 \times 1/4 \times 1/4 = 1/8 \]
+\end{column}
+
+\begin{column}{0.4\textwidth}
+\begin{figure}
+\includegraphics[totalheight=0.8\textheight]{spn.jpg}
+\end{figure}
+\end{column}
+\end{columns}
+
 \vspace{2mm}
-Each of these has a bias of magnitude $1/4$. We can combine:
-\[  V_{2,6} \oplus V_{2,8} \oplus P_5 \oplus P_7 \oplus P_8 \oplus K_{1,5} \oplus K_{1,7} \oplus K_{1,8} \oplus K_{2,6} = 0 \]
-By the Piling Up Principle this expression holds with bias:
-\[ 2 * 1/4 * 1/4 = 1/8 \]
-Note the implicit assumption that S-Boxes are independant. More on this later.
+Note the implicit assumption that S-Boxes are independent. More on this later.
+
 \end{frame}
@@
 \vspace{3mm}
-We have an expression that links plaintext bits to input bits to the 4th round of SPN that holds with high bias ($1/32$). We can partially undo the last round by guessing the last key. \\
+We have an expression that links plaintext bits to input bits to the 4th round of SPN that holds with high bias ($1/32$). 
+\[ U_{4,6} \oplus U_{4,8} \oplus U_{4,14} \oplus U_{4, 16} \oplus P_5 \oplus P_7 \oplus P_8 = \Sigma K \]
+We can partially undo the last round by guessing the last key. \\
 \vspace{3mm}
@@
 \pause
-BUT To do this, we don't need to guess the entire key for the last round! Our expression only involves 4 key bits, output from 2 S-Boxes. Thus we only need to guess $2^8 = 256$ values.\\
+BUT To do this, we don't need to guess the entire key for the last round!
+Our expression only involves 4 fourth round input ($U_4$) bits, output from 2 S-Boxes of the third round.
+Thus we only need to guess $2^8 = 256$ values.\\
 \vspace{3mm}
@@
 
+\begin{frame}
+Actual attack on SPN ... Demo ...
+\end{frame}
+
 \subsection{Differential Attack on SPN}
@@
 Differential Cryptanalysis tries to exploit the high probability of certain occurrences of differential characteristics $(\Delta P \rightarrow \Delta C)$ in the cipher.
-This is a chosen ciphertext attack.
+This is a chosen plaintext attack.
 
-[Differential: $(\Delta P, \Delta C) = (P \oplus P', C \oplus C') $]
+\vspace{2mm}
+Differential: $(\Delta P, \Delta C) = (P \oplus P', C \oplus C')$
 
+\vspace{5mm}
 We have to analyze the cipher in pieces. The basic idea is
@@
 
+\vspace{2mm}
 Let us consider the S-Box of the SPN cipher as we constructed it:
@@
 
+\vspace{2mm}
 Algorithm
@@
 \item{Iterate through the first input $X = \{0000, ... ,1111 \}$}
-\item{This fixes the second input $X' = \{0100, ... ,1011\}$ (ordering does not matter)}
+\item{This fixes the second input $X' = X \oplus \Delta X = \{0100, ... ,1011\}$ (ordering does not matter)}
 \item{Obtain corresponding output difference $\Delta Y = Y \oplus Y'$}
@@
 
-Distribution of differences
+Note that, for $\Delta X = 1011$, $\Delta Y = 0010$ occurs $8$ times, out of the possible $16$ times. So, the pair $(1011,0010)$ has a probability of occurrence $8/16 = 1/2$.
 
-Note that, for $\Delta X = 1011$, $\Delta Y = 0010$ occurs $8$ times, out of the possible $16$ times.
-So, the difference pair $(1011,0010)$ has a probability of occurrence $\frac{8}{16} = \frac{1}{2}$.
-So, we can tabulate the complete data for the S-Box to check this probabilities.
+\vspace{2mm}
+We can tabulate the complete data for the S-Box to check this probabilities.
 \end{frame}
@@
 \item{But this table clearly shows bias towards certain pairs}
-\item{Just divide the entries by $2^4 = 16$ to get the probabilities, and exploit the scenario of highest probability of occurrence}
+\item{Just divide the entries by $2^4 = 16$ to get the probabilities, and exploit the scenario of highest probability of occurrence ($Pr(2|B) = 8/16 = 1/2$ here)}
 \end{itemize}
@@
 \begin{frame}
-Why can not we make an ideal S-Box?
+A few properties of the S-Box Distribution Table
 
+\vspace{5mm}
 Notice that in the difference table
@@
 
+\vspace{5mm}
 As an ideal S-Box expects to give out no information of $\Delta Y$ given $\Delta X$, we wish it could have all entries $1$, i.e,
-\[ Pr(\Delta Y | \Delta X) = \frac{1}{16} \quad \forall \Delta X  \]
-But, from our discussion about the distribution table, this is infeasible.
+\[ Pr(\Delta Y | \Delta X) = \frac{1}{16} \quad \forall \: \Delta X  \]
+But, from our discussion above, this is infeasible, and we hope to exploit this.
 \end{frame}
@@
 
-Let us consider the S-Box of the SPN cipher, with the round keys this time:
+\vspace{5mm}
+Let us consider the S-Box of the SPN cipher, with the round keys this time.
 \begin{columns}
@@
 
+\vspace{5mm}
 Note that for each pair of inputs $(W,W')$, the actual inputs to the S-Box are $(X,X') = (W \oplus K, W' \oplus K)$. Hence, the input difference for the keyed S-Box is 
@@
 
-Thus, the keyed S-Box has the same difference distribution table as the unkeyed S-Box.
+\vspace{2mm}
+Thus, the keyed S-Box has the same difference distribution table.
 \end{frame}
@@
 \begin{frame}
-Constructing a Differential Characteristic
 
@@
 \begin{column}{0.6\textwidth}
+Constructing a Differential Characteristic
 
+\vspace{5mm}
 The diagram beside illustrates the tracing of the non-zero bits of the input difference through the SPN structure.
 
+\vspace{2mm}
 Input Diff: $\Delta P = 0B00$ \\
 
+\vspace{2mm}
 Propagating Difference Pairs
@@
 
+\vspace{2mm}
 Differential Characteristic
@@
 
+\vspace{2mm}
 Notation
-$\Delta U_i$: Input difference to $i$-th round
-$\Delta V_i$: Output difference from $i$-th round
-$K_i$: Round Key for $i$-th round
-$X_{i,j}$: $j$-th bit of $X_i$
+\begin{itemize}
+\item{$\Delta U_i$: Input difference to $i$-th round}
+\item{$\Delta V_i$: Output difference from $i$-th round}
+\item{$K_i$: Round Key for $i$-th round}
+\item{$X_{i,j}$: $j$-th bit of $X_i$}
+\end{itemize}
 
+\vspace{2mm}
 Here, we have the following in the differential characteristic
 \begin{itemize}
-\item{Round 1: $\Delta P = \Delta U_1 = 0B00 \rightarrow 0040 = \Delta V_1$ [$Pr = \frac{8}{16}$]}
-\item{Round 2: $\Delta V_1 = \Delta U_2 = 0040 \rightarrow 0220 = \Delta V_2$ [$Pr = \frac{6}{16}$]}
-\item{Round 3: $\Delta V_2 = \Delta U_3 = 0220 \rightarrow 0606 = \Delta V_3$ [$Pr = \left(\frac{6}{16}\right)^2$]}
+\item{Round 1: $\Delta P = \Delta U_1 = 0B00 \rightarrow 0040 = \Delta V_1$  [$Pr = 8/16$]}
+\item{Round 2: $\Delta V_1 = \Delta U_2 = 0040 \rightarrow 0220 = \Delta V_2$  [$Pr = 6/16$]}
+\item{Round 3: $\Delta V_2 = \Delta U_3 = 0220 \rightarrow 0606 = \Delta V_3$  [$Pr = \left(6/16\right)^2$]}
 \item{Round 4: $\Delta V_3 = \Delta U_4 = 0606$}
@@
 
+\vspace{2mm}
 Hence, assuming the S-Boxes to be independent over the rounds, we obtain
 \[ \Delta P = 0B00 \rightarrow 0606 = \Delta U_4 \]
-with probability $Pr(\Delta U_4 = 0606 | \Delta P = 0B00) = \frac{8}{16} \times \frac{6}{16} \times \left(\frac{6}{16}\right)^2 = \frac{27}{1024}$.
+with probability $Pr(\Delta U_4 | \Delta P) = 8/16 \times 6/16 \times \left(6/16\right)^2 = 27/1024$.
 
-Note: The probability of this characteristic $Pr = \frac{27}{2^{10}} \gg \frac{1}{2^{16}}$
+\vspace{2mm}
+Note: The probability of this characteristic $Pr = 27/1024 \gg 1/2^{16}$
 
-Now, lets see if we can exploit this situation to extract the key bits.
 \end{frame}
@@
 \begin{itemize}
-\item{We have a differential characteristic that links plaintext difference $\Delta P$ to input difference $\Delta U_4$ of the 4th round output of SPN and holds with high probability ($\frac{27}{1024}$)}
+\item{We have a differential characteristic that links plaintext difference $\Delta P$ to input difference $\Delta U_4$ of the 4th round output of SPN and holds with high probability ($27/1024$)}
 \item{We can partially undo the last round by guessing the last key $K_5$}
-\item{If we guess correctly, the differential characteristic will hold with high probability, close to $\frac{27}{1024}$. If we guess wrong, it will hold with low probability}
+\item{If we guess correctly, the differential characteristic will hold with high probability, close to $27/1024$. If we guess wrong, it will hold with low probability}
 \item{But in this case, we don't need to guess the entire key $K_5$ for the last round!}
@@
 
+\vspace{5mm}
 Input Diff: $\Delta P = 0B00$
 
+\vspace{2mm}
 Characteristic
@@
 
-Probability $Pr(\Delta U_4 = 0606 | \Delta P = 0B00) = \frac{27}{1024}$
+\vspace{2mm}
+Probability $Pr(\Delta U_4 = 0606 | \Delta P = 0B00) = 27/1024$
 
+\vspace{5mm}
 Partial Subkey guess only for
@@
 
-\subsection{Actual Attacks}
 \begin{frame}
-\begin{beamercolorbox}[ht=2.5ex,dp=1.125ex,center,rounded=true,shadow=true]{author in head/foot}
-Let's get our hands dirty!
-\end{beamercolorbox}
-\end{frame}
-
-\begin{frame}
-Demo of attacks on SPN ... Runtime comparisons with Brute Force ...
-
+Actual attack on SPN ... Demo ...
 \end{frame}
@@
 \begin{frame}
+The plan for the DES Attacks:
 \begin{itemize}
 \item{Structurally and cryptographically, DES is similar to SPN.}
-\item{We will adapt the linear attack on SPN to a linear attack on a 3-round DES algorithm and mount this attack.}
+\item{We will adapt the linear attack on SPN to a linear attack on a 4-round DES algorithm and mount this attack.}
 \item{We will also show how to extend this attack to a full 16-round DES algorithm and show it is faster than an exhaustive search.}
@@
 We can use the same technique as before.
-For a given S-Box $a \in \{1, 2, \dots 8\}, 1 \leq \alpha \leq 63, 1 \leq \beta \leq 15$ we define $NS_{a}(\alpha, \beta)$ as the number of times (out of 64) for S-Box $a$ that for all the input patterns masked by $\alpha$ the output pattern masked by $\beta$ agrees with the value of S-Box $a$.
+For a given S-Box $a \in \{1, 2, \dots 8\}$, $1 \leq \alpha \leq 63$, $1 \leq \beta \leq 15$, we define $NS_{a}(\alpha, \beta)$ as the number of times (out of 64) for S-Box $a$ that for all the input patterns masked by $\alpha$ the output pattern masked by $\beta$ agrees with the value of S-Box $a$.
 \[ NS_{a}(\alpha, \beta) = | \{ X | 0 \leq X < 64, ( \bigoplus_{i=0}^5 X[i]\cdot\alpha[i]) = (\bigoplus_{j=0}^3 S_a(X)[j] \cdot \beta[j])\} | \]
@@
 \begin{frame}
-Let $P_L$ and $P_R$ be the left and right plaintext bits. Let $C_L$ and $C_R$ be the left and right ciphertext bits. Let $K_i$ be the round key for round $i$. \\
+Let $P_L$ and $P_R$ be the left and right plaintext bits.
+Let $C_L$ and $C_R$ be the left and right ciphertext bits.
+Let $K_i$ be the round key for round $i$ and $U_{4L}$ and $U_{4R}$ the left and right halves of the input to round 4.\\
 \vspace{5mm}
@@
 \begin{align*}
- P_L{25} & \oplus P_L{14} \oplus P_L{8} \oplus P_L{3} \oplus \\
-         & C_L{25} \oplus C_L{14} \oplus C_L{8} \oplus C_L{3} \oplus \\
-			& P_R{17} \oplus C_R{17} = K_1{26} \oplus K_3{26} \\
+ P_L[25] & \oplus P_L[14] \oplus P_L[8] \oplus P_L[3] \oplus \\
+         & U_{4L}[25] \oplus U_{4L}[14] \oplus U_{4L}[8] \oplus U_{4L}[3] \oplus \\
+			& P_R[17] \oplus U_{4R}[17] = K_1[26] \oplus K_3[26] \\
 \end{align*}
@@
 \pause
-Note that in the last round, $C_L$ bits are not changed, only $C_R$ bits are affected. Note also that there is only one $C_R$ bit the 3-round equation.\\
+Note that in the last round, $U_{4L}$ bits are not changed, only $U_{4R}$ bits are affected. Note also that there is only one $U_R$ bit the 3-round equation.
+\begin{align*} 
+P_L[25] \oplus P_L[14] \oplus P_L[8] \oplus & P_L[3] \oplus U_{4L}[25] \oplus U_{4L}[14] \oplus U_{4L}[8] \\ & \oplus U_{4L}[3] \oplus P_R[17] \oplus U_{4R}[17] = K_1[26] \oplus K_3[26]
+\end{align*}
 \vspace{5mm}
 \pause
-This bit, $C_R{17}$ is affected by the result of S-Box 1 in round 4.
+This bit, $U_{4R}[17]$ is affected by the result of S-Box 1 in round 4.
 If we guess correctly for the 6 partial subkey bits $K_{4,1}$ to $K_{4,6}$, the 3-round equation will hold with high bias, otherwise it will likely hold with close to 0 bias.\\
@@
 \begin{frame}
-How many pairs do we need? Probabalistic justification:\\
+How many pairs do we need? (Probabilistic justification)
+
 \vspace{3mm}
@@
 \begin{frame}
-How many pairs do we need? Continued\\
+How many pairs do we need? (Continued)
+
 \vspace{3mm}
@@
 
+\begin{frame}
+The actual Attack on 4 round DES ... Demo ...
+\end{frame}
+
 \subsection{Differential Attack: 4-round DES}
@@
 
+\vspace{5mm}
 Similar to the differential attack on SPN, here we care about the differential characteristics of the cipher. So, the basic idea is as follows
@@
 \item{Find the most probable difference pairs}
-\item{Combine the pairs through all the rounds to get a higly probable differential characteristic until the penultimate round}
-\item{Exploit this differential characteristic by chosing a number of pairs of pliantexts with the desired difference and undo the last round by guessing the keybits}
+\item{Combine the pairs through all the rounds to get a highly probable differential characteristic until the penultimate round}
+\item{Exploit this differential characteristic by chosing a number of pairs of plaintexts with the desired difference and undo the last round by guessing the keybits}
 \item{If the initial choice of the input difference was smart, we will only need to guess a few keybits}
@@
 \begin{frame}
-Demo of the attack ...
-
+The actual Attack on 4 round DES ... Demo ...
 \end{frame}
@@
 What if that is false ? ... Discussion ...
-
-\end{frame}
-
-\subsection{Number of Pairs Required}
-\begin{frame}
-Probabilistic possibility of the attacks ... Pairs of data points needed ...
-
-\end{frame}
-
-\begin{frame}
-Generalized attack idea on any given system ... Discussion ...
 

### 2009-01-31 src/DifferentialCryptanalysis.java

@@ -208,9 +208,10 @@ public class DifferentialCryptanalysis {
 		    	BitSet delta_U4R = Util.copyBitSet(C1R_prev, 32);
 		        delta_U4R.xor(C2R_prev);
 		        
-		        BitSet check = Util.toBitSet(0x00008002,32);
+		        BitSet check = Util.toBitSet(0x00008000,32);
 		    	BitSet bits = delta_U4R.get(0,32);
-		        boolean lhs = Util.equalsBitSet(bits, check, 32);
+		    	System.out.print(Util.toInteger(bits, 32));
+		        boolean lhs = delta_U4R.get(16); //Util.equalsBitSet(bits, check, 32);
 	            
 	            if (lhs == true) counts[i]++;	     
 		    }

### 2009-01-31 trunk/talk/draft1.tex
History portion done ... Overlays upto Chapter 2 done ... Photos modified ... Random typos and formating fixed ...

@@
 \begin{frame}
+\textit{\huge $\mathcal O$}nce upon a time ... \pause there was a secure standard encryption algorithm called DES (Data Encryption Standard) ...
+
+\vspace{2mm}
+\pause
+\begin{itemize}[<+->]
+\item{1973: NBS publishes a request for a standard encryption algorithm}
+\item{1975: DES is published in the Federal Register for comment}
+\item{1976: DES is approved as a standard}
+\item{1977: DES is published as a FIPS standard FIPS PUB 46}
+\item{1992: Biham and Shamir report the {\it differential cryptanalysis}}
+\item{1993: DES is reaffirmed for the third time as FIPS 46-2}
+\item{1994: The first {\it linear cryptanalysis} of DES is performed by Matsui}
+\item{1997: The DESCHALL Project breaks a DES message in public}
+\item{1998: The EFF's Deep Crack {\it brute forces} a DES key in 56 hours}
+\item{1999: Deep Crack and distributed.net break DES in 22 hours}
+\item{1999: FIPS 46-3 reaffirms DES, preferring the use of Triple DES}
+\item{2001: The Advanced Encryption Standard (AES) is published in FIPS 197}
+\item{2002: The AES standard becomes effective}
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\tableofcontents
+\end{frame}
+
+\begin{frame}
 \begin{beamercolorbox}[ht=2.5ex,dp=1.125ex,center,rounded=true,shadow=true]{author in head/foot}
@@
 
-\subsection{Data Encrytption Standard}
+\subsection{Data Encryption Standard}
 \begin{frame}
@@
 \end{columns}
+
+{\tiny Designers: H Feistel, W Tuchman, D Coppersmith, A Konheim, E Grossman, B Notz, L Smith and B Tuckerman from IBM}
 \end{frame}
@@
 Mathematically, the different pieces of DES are:
+
 \begin{itemize}
-\item{Initial Permutation $IP : (0,1)^{64} \rightarrow (0,1)^{64}$}
-\item{DES Unit Block functions $\sigma_{i+1} (L_i,R_i)$ for $i = 0,1,...,15$}
-\item{Transpositions $\tau : (L_{i+1},R_{i+1}) \mapsto (R_{i+1},L_{i+1})$ for $i = 0,1,...,14$}
-\item{Inverse Initial Permutation $IP^{-1} : (0,1)^{64} \rightarrow (0,1)^{64}$}
+\item{Initial Permutation \\ $\qquad IP : (0,1)^{64} \rightarrow (0,1)^{64}$}
+\item{DES Unit Block functions \\ $\qquad \sigma_{i+1} (L_i,R_i)$ for $i = 0,1,...,15$}
+\item{Transposition \\ $\qquad \tau : (L_{i+1},R_{i+1}) \mapsto (R_{i+1},L_{i+1})$ for $i = 0,1,...,14$}
+\item{Inverse Initial Permutation \\ $\qquad IP^{-1} : (0,1)^{64} \rightarrow (0,1)^{64}$}
 \end{itemize}
 
+\pause
 \begin{block}{Algebraic Representation of DES}
-\begin{align*}
-C &= DES_K (P) = IP^{-1} \sigma_{16} \tau \cdots \tau \sigma_1 IP (P) \\
-P &= DES^{-1}_K (C) = IP^{-1} \sigma_{1} \tau \cdots \tau \sigma_{16} IP (C)
-\end{align*}
+$\qquad C \:  = \:  DES_K (P) \: = \: IP^{-1} \: \sigma_{16} \:  \tau \: \cdots \:  \tau \:  \sigma_1 \: IP \: (\: P \: )$
+
+$\qquad P \: = \: DES^{-1}_K (C) \: = \: IP^{-1} \: \sigma_{1} \: \tau \: \cdots \: \tau \: \sigma_{16} \: IP \: (\: C\: )$
 \end{block}
 
-Note: The inverse DES comes from the fact that $\tau^2 = \sigma_i^2 = 1$
+\pause
+\vspace{0.2in}
+{\small Note: The inverse DES comes from the fact that $\tau^2 = \sigma_i^2 = 1$}
 \end{frame}
@@
 \begin{frame}
-Linear Attack on DES idea:\\
-\begin{itemize}
-\item S-Boxes depend on a relatively small number of bits (6) which allows us to write down linear (or afine) expressions that approximate S-Boxes \\
-\item The effects of one round do not diffuse quickly over following rounds. Thus linear expresions (as above) that hold per-round can be combined across rounds.
+Linear Attack on DES idea:
+
+\begin{itemize}[<+->]
+\item S-Boxes depend on a relatively small number of bits (6) which allows us to write down linear (or affine) expressions that approximate S-Boxes
+\item The effects of one round do not diffuse quickly over following rounds. Thus linear or affine expressions (as above) that hold per-round can be combined across rounds
 \end{itemize}
@@
 \begin{frame}
-Specifically, if $P_{i}$ are plaintext bits, $C_{i}$ are cyphertext bits, and $K_{i}$ are subkey bits, then we wish to find an expression of the form
+Specifically, if $P_{i}$ are plaintext bits, $C_{i}$ are ciphertext bits, and $K_{i}$ are subkey bits, then we wish to find an expression of the form
 \[ P_{i_1} \oplus P_{i_2} \cdots P_{i_j} \oplus C_{i_1} \oplus C_{i_2} \cdots C_{i_k} = K_{i_1} \oplus K_{i_2} \oplus K_{i_m} \]
-Such that this expression has a high \textit{or} low probability of occurence.\\
+\pause such that this expression has a high \textit{or} low probability of occurrence.
+
 \vspace{5mm}
-Consider: No such obvious expression should exist, otherwise the cipher is trivially weak.
-If we were to randomly select bits for the above expression, it would hold exactly $1/2$ the time.
-If we find an expression such as above that displays a high \textit{bias}, that is, it holds much more or less freqently than $1/2$ the time, we can exploit this.
+\pause Consider: No such obvious expression should exist, otherwise the cipher is trivially weak. If we were to randomly select bits for the above expression, it would hold exactly $1/2$ the time.
+
+\vspace{5mm}
+\pause If we find an expression such as above that displays a high \textit{bias}, that is, it holds much more or less frequently than $1/2$ the time, we can exploit this.
 \end{frame}
@@
 
-Vulnerability of DES:\\
-\begin{itemize}
-\item{S-Boxes show a strong tendency to produce certain differential pairs with high probability, rather than being random.}
-\item{The differences do not get diffused fast enough.}
-\item{Differentials are not affected by the round keys.}
+\vspace{0.2in}
+\pause Vulnerability of DES:
+
+\begin{itemize}[<+->]
+\item{S-Boxes show a strong tendency to produce certain differential pairs with high probability, rather than being random}
+\item{The differences do not get diffused fast enough through the permutations}
+\item{Differentials are not affected by the round keys as they get XOR-ed out}
 \end{itemize}
@@
 \begin{frame}
-In an $k$-round DES, we can trace the path of a certain input difference to get an output difference with a high probability. This tracing path is known as a {\it differential characteristic}.
+In an $k$-round DES, we can trace the path of a certain input difference to get an output difference with a high probability. \pause This tracing path is known as a {\it differential characteristic}.
 
@@
 
-In an ideal situation, one will expect the probability
+\pause In an ideal situation, one will expect the probability
 \[ P \left(\Delta C | \Delta P \right) = \frac{1}{2^n} \]
-for an $n$-bit cipher system. Differential attacks seek to exploit a scenario where
+for an $n$-bit cipher system. \pause Differential attacks seek to exploit a scenario where
 \[ P \left(\Delta C | \Delta P \right) = p_D \gg \frac{1}{2^n} \]
 
-This is essentially a chosen-plaintext attack as we want the specific input difference to occur for every pair.
+\pause Note: This is essentially a chosen-plaintext attack as we want the specific input difference to occur for every pair.
 \end{frame}
@@
 \end{figure}
-Let us simplify DES by
-\begin{itemize}
-\item{Reducing the bitlength from $64$ to $16$}
+\pause Let us simplify DES by
+\begin{itemize}[<+->]
+\item{Reducing the code length from $64$ to $16$}
 \item{Removing expansion function $E(R)$}
@@
 \begin{column}{0.4\textwidth}
+\pause
 \begin{figure}
@@
 \begin{center}
-{\tiny Disclaimer: This simplification does not affect the discussion of the techniques of Linear/Differential Cryptanalysis.}
+\pause {\tiny Disclaimer: This simplification does not affect the discussion of the techniques of Linear/Differential Cryptanalysis.}
 \end{center}
@@
 \begin{frame}
-We define a simple Substitution-Permutation Network Cipher (SPN) which has structural similarity to DES.
+We construct a simple Substitution-Permutation Network Cipher (SPN) which has structural similarity to DES.
 
-\begin{itemize}
-\item SPN is a 16-bit block cipher (16-bit plaintext to 16-bit ciphertext) with 16-bit round keys.
+\begin{itemize}[<+->]
+\item SPN is a 16-bit block cipher (16-bit plaintext to 16-bit ciphertext) with 16-bit round keys
 \item Each round consists of a substitution and a permutation (much like in DES)
-\item The substitution is the result of splitting the 16-bit input block to a round in to 4 4-bit sub-blocks.
-Each 4-bit sub-block is then mapped across a 4-bit to 4-bit S-box.
-We use the same S-box throughout.
-\item The permutation is applied to all 16-bits of the round following the substitution.
+\item The substitution is the result of splitting the 16-bit input block to a round in to 4$\times$4-bit sub-blocks.
+Each 4-bit sub-block is then mapped across a 4-bit to 4-bit S-Box.
+We use the same S-Box throughout
+\item The permutation is applied to all 16-bits of the round following the substitution
 \item Finally, key-mixing is achieved by XOR-ing the round key with every input block to a round as well as at the end of the last round (so that we can't simply ignore the last round)
-\item The SPN cipher we will look at will consist of 4 rounds.
+\item The SPN cipher we will look at will consist of 4 rounds
 \end{itemize}
@@
 \vspace{5mm}
+\pause The S-Box provides the {\it confusion} function and the permutation applies the {\it confusion} operation in SPN, thus making it cryptographically similar to DES.
+\vspace{5mm}
 
@@
 \begin{frame}
-\large{Basic Definitions for the Linear attack:}\\
+Basic Definitions for the Linear attack
+
 \vspace{5 mm}
-Linear cryptanalysis tries to take advantage of high probability occurrences of linear expressions involving plaintext bits, ciphertext bits and subkey bits. It is a known plaintext attack.\\
+\pause Linear cryptanalysis tries to take advantage of high probability occurrences of linear expressions involving plaintext bits, ciphertext bits and subkey bits.
+It is a {\it known plaintext attack}.
+
 \vspace{5 mm}
-Define $P_i$ where $i = 1, 2, \dots 16$ as the $i$-th plaintext bit.\\
-Define $C_i$ where $i = 1, 2, \dots 16$ as the $i$-th cyphertext bit.\\
-Define $U_{j,i}$ where $j = 1, 2, 3, 4$ and $i = 1, 2, \dots 16$ be the $i$-th input bit to the $j$-th round of SPN.\\
-Define $V_{j,i}$ where $j = 1, 2, 3, 4$ and $i = 1, 2, \dots 16$ be the $i$-th output bit of the $j$-th round of SPN.\\
+\pause Define 
+\begin{itemize}[<+->]
+\item{$P_i$ where $i = 1, 2, \dots 16$ as the $i$-th plaintext bit}
+\item{$C_i$ where $i = 1, 2, \dots 16$ as the $i$-th ciphertext bit}
+\item{$U_{j,i}$ where $j = 1, 2, 3, 4$ and $i = 1, 2, \dots 16$ be the $i$-th input bit to the $j$-th round of SPN}
+\item{$V_{j,i}$ where $j = 1, 2, 3, 4$ and $i = 1, 2, \dots 16$ be the $i$-th output bit of the $j$-th round of SPN}
+\end{itemize}
 \end{frame}
@@
 \begin{frame}
-Linear and Affine approximation of S-Box:\\
+Linear and Affine approximation of S-Box
+
 \vspace{5mm}
-How do we come up with the desired expression for the entire cipher? We start by looking at the only non-linear component, the S-Box.\\
+\begin{itemize}[<+->]
+\item Question: How do we come up with the desired expression for the entire cipher? 
+\item Hint: We start by looking at the only non-linear component, the S-Box.
+\end{itemize}
+
 \vspace{5mm}
-To find the linear/afine approximation of the S-Box we simply consider every possible expression of the input bits $X_i$ and output bits $Y_j$. Thus the expression has the form
+\pause To find the linear/affine approximation of the S-Box we simply consider every possible expression of the input bits $X_i$ and output bits $Y_j$. \pause Thus the expression has the form
 \[ \bigoplus_{i \in U} X_i = \bigoplus_{j \in V} Y_j \]
-Where $U$ and $V$ range over all possible subsets of $\{1, 2, 3 ,4\}$. We then compare how often this expression coincides with the S-Box. \\
+where $U$ and $V$ range over all possible subsets of $\{1, 2, 3 ,4\}$. \pause We then compare how often this expression coincides with the S-Box.
+
 \vspace{5mm}
-Note that there are 16 possibilities for $U$ and $V$, hence 256 total possible expressions. (this is for a 4-bit S-box)
+\pause Note that there are 16 possibilities for $U$ and $V$, hence 256 total possible expressions (for a 4-bit S-box).
 \end{frame}
@@
 \begin{frame}
-S-Box approximatin example:\\
+S-Box Approximation Example
+
 \vspace{5mm}
+\begin{columns}
+\begin{column}{0.3\textwidth}
+\begin{figure}
+\includegraphics[width=0.7\textwidth]{spn_sbox_unkeyed.jpg}
+\end{figure}
+\end{column}
+
+\begin{column}{0.7\textwidth}
 We can take the following expression as an example:
 \[ X_2 \oplus X_3 = Y_1 \oplus Y_3 \oplus Y_4 \]
-Applying all possible values for the input $X$ bits it turns out that the expression holds in 12 out of the 16 cases. Hence, this expression has a bias of $12/16 - 1/2 = 1/4$.\\
+\pause Applying all possible values for the input $X$ bits it turns out that the expression holds in 12 out of the 16 cases. \pause Hence, this expression has a bias of $12/16 - 1/2 = 1/4$.
+\end{column}
+\end{columns}
+
 \vspace{5mm}
-Similarly the following expression
+\pause Similarly the following expression
 \[ X_3 \oplus X_4 = Y_1 \oplus Y_4 \]
-Has a bias of $-3/8$, which is higher in magnitude than the previous one. It is the magnitude of the bias that is important.
+Has a bias of $-3/8$, which is higher in magnitude than the previous one. \pause It is the magnitude of the bias that is important.
 \end{frame}
@@
 \begin{frame}
-What this means for 1 Round:\\
+\begin{columns}
+\begin{column}{0.4\textwidth}
+What this means for 1 Round:
+
 \vspace{5mm}
-Note, from the previous table, that the expression $X_1 \oplus X_3 \oplus X_4 = Y_2$ has a bias of $+1/4$. \\
+Note, from the previous table, that the expression \[ X_1 \oplus X_3 \oplus X_4 = Y_2 \] has a bias of $+1/4$.
+
 \vspace{5mm}
-Note also that $U_1 = P \oplus K_1$. 
-\vspace{5mm} We can now write down the following linear approximation across the 1st round of SPN:
+\onslide<2->{Note also that $U_1 = P \oplus K_1$.}
+\end{column}
+\begin{column}{0.6\textwidth}
+\begin{figure}
+\includegraphics[height=0.45\textheight]{spn_single_round.jpg}
+\end{figure}
+\end{column}
+\end{columns}
+
+\vspace{5mm} 
+\onslide<3->{We can now write down the following linear approximation across the 1st round of SPN:
 \begin{eqnarray*}
-V_{1,6} & = & U_{1,5} \oplus U_{1,7} \oplus U_{1,8} \mbox{  $\leftarrow$ S-Box approximation above}\\
+V_{1,6} & = & U_{1,5} \oplus U_{1,7} \oplus U_{1,8} \quad \mbox{  $\leftarrow$ S-Box $S_{12}$ approximation above}\\
 		  & = & (P_5\oplus K_{1,5}) \oplus (P_7\oplus K_{1,7}) \oplus (P_8\oplus K_{1,8})
 \end{eqnarray*}
-This expression holds with probability of $3/4$ (bias of $+1/4$)
+
+\pause This expression holds with probability of $3/4$ (bias of $+1/4$)}
 \end{frame}
@@
 We can get expressions that hold with some non-$1/2$ probability for every round.
-But we must somehow combine them to write an expression relating the plaintext and ciphertext bits $\Rightarrow$ We use the Piling Up Principle:\\
+But we must somehow combine them to write an expression relating the plaintext and ciphertext bits.
+\pause We use the Piling Up Principle for this purpose.
+
 \vspace{5mm}
-Note: \\
-$X_1 \oplus X_2 = 0$ is a linear expression equivalent to $X_1 = X_2$\\
-$X_1 \oplus X_2 = 1$ is an affine expression equivalent to $X_1 \neq X_2$\\
+\pause Note:
+
+$\qquad X_1 \oplus X_2 = 0$ is a linear expression equivalent to $X_1 = X_2$\\
+$\qquad X_1 \oplus X_2 = 1$ is an affine expression equivalent to $X_1 \neq X_2$
+
 \vspace{5mm}
-Assume the following probability distribution
+\pause Assume the following probability distribution
 \[ Pr(X_1 = i)  = \left\{ \begin{array}{ll}
@@
 			(1 - p_1)(1 - p_2) & \mbox{$i = 1, j = 1$}.\end{array} \right. \] 
-Now, note that $X_1 \oplus X_2 = 0 \Rightarrow X_1 = 1, X_2 =1$ OR $X_1 = 0, X_2 = 0$. Hence:
+\pause Now, note that $X_1 \oplus X_2 = 0 \Rightarrow X_1 = 1, X_2 =1$ OR $X_1 = 0, X_2 = 0$. \pause Hence,
 \begin{eqnarray*}
@@
 \end{eqnarray*}
-If we write $p_k = 1/2 + \epsilon_k$ we have
+\pause If we write $p_k = 1/2 + \epsilon_k$ we have
 \[ Pr(X_1 \oplus X_2 = 0) = 1/2 + 2\epsilon_1 \epsilon_2 \]
-Hence the bias of $X_1 \oplus X_2 = 0$ is 
+\pause Hence the bias of $X_1 \oplus X_2 = 0$ is 
 \[ 2 \epsilon_1 \epsilon_2 \]
@@
 \begin{frame}
-Piling Up Principle Continued - The Piling Up Lemma:\\
+Piling Up Principle (continued) - The Piling Up Lemma:\\
 \vspace{5mm}
-We extend the idea on the previous slide to the general case: For $n$ independant random binary variables, $X_1, X_2, \dots, X_n$, 
+We extend the idea on the previous slide to the general case \pause : For $n$ independent random binary variables, $X_1, X_2, \dots, X_n$, 
 \[ Pr(X_1 \oplus X_2 \cdots \oplus X_n = 0) = 1/2 + 2^{n-1} \prod_{i=1}^n \epsilon_i \]
-In terms of bias,
+\pause In terms of bias,
 \[ \epsilon_{1,2,\dots,n} = 2^{n-1} \prod_{i = 1}^n \epsilon_i \]
@@
 \vspace{5mm}
-Note that if some $\epsilon_k = 0$ (that is, some $X_k$ has no bias) then the total bias will be $0$.  
+\pause Note that if some $\epsilon_k = 0$ (that is, some $X_k$ has no bias) then total bias will be $0$. More about the probabilistic analysis later.
 \end{frame}
@@
 \begin{frame}
-Now, come back to Per-round approximations. We can write down 4 approximations (Let $S_{i,j}$ repesent the $j$-th S-Box in the $i$-th round):
-\begin{eqnarray*}
-S_{1,2}: X_1 \oplus X_3 \oplus X_4 & = & Y_2 \\
-S_{2,2}: X_2 = Y_2 \oplus Y_4 \\
-S_{3,2}: X_2 = Y_2 \oplus Y_4 \\
-S_{3,4}: X_2 = Y_2 \oplus Y_4 \\
-\end{eqnarray*}
-Each of these has a probability bias magnitude of $1/4$. We can use the Piling Up Principle to combine them into a single expression relating plaintext bits to cyphertext bits.
+
+\begin{columns}
+\begin{column}{0.6\textwidth}
+Now, come back to per-round approximations. We can write down 4 approximations ($S_{ij}$ represents the $j$-th S-Box in the $i$-th round):
+\begin{align*}
+& S_{12}: X_1 \oplus X_3 \oplus X_4 = Y_2 \\
+& S_{22}: X_2 = Y_2 \oplus Y_4 \\
+& S_{32}: X_2 = Y_2 \oplus Y_4 \\
+& S_{34}: X_2 = Y_2 \oplus Y_4 \\
+\end{align*}
+\onslide<2->{Each of these has a probability bias magnitude of $1/4$. We can use the Piling Up Principle to combine them into a single expression relating plaintext bits to ciphertext bits.}
+\end{column}
+
+\begin{column}{0.4\textwidth}
+\begin{figure}
+\includegraphics[totalheight=0.8\textheight]{spn.jpg}
+\end{figure}
+\end{column}
+\end{columns}
+
 \end{frame}
@@
 \begin{frame}
-Consider the first 2 rounds:\\
-\vspace{5mm}
-\[  X_1 \oplus X_3 \oplus X_4 = Y_2 \Rightarrow \]
-\[ V_{1,6} = (P_5\oplus K_{1,5}) \oplus (P_7\oplus K_{1,7}) \oplus (P_8\oplus K_{1,8}) \]
+\begin{columns}
+\begin{column}{0.6\textwidth}
+Consider the first 2 rounds:
+\begin{align*}
+& X_1 \oplus X_3 \oplus X_4 = Y_2 \\ 
+\Rightarrow &(P_5\oplus K_{1,5}) \oplus (P_7\oplus K_{1,7}) \oplus (P_8\oplus K_{1,8}) =  V_{1,6}
+\end{align*}
+\[ X_2 = Y_2 \oplus Y_4 \: \Rightarrow \: (V_{1,6} \oplus K_{2,6}) =  V_{2,6} \oplus V_{2,8}\]
+
 \vspace{2mm}
-\[ X_2 = Y_2 \oplus Y_4 \Rightarrow \]
-\[ V_{2,6} \oplus V_{2,8} = V_{1,6} \oplus K_{2,6} \]
+\onslide<2->{Each of these has a bias of magnitude $1/4$ and we can combine to obtain:
+\[ V_{2,6} \oplus V_{2,8} \oplus P_5 \oplus P_7 \oplus P_8 \oplus K_{1,5} \oplus K_{1,7} \oplus K_{1,8} \oplus K_{2,6} = 0 \]}
+
 \vspace{2mm}
-Each of these has a bias of magnitude $1/4$. We can combine:
-\[  V_{2,6} \oplus V_{2,8} \oplus P_5 \oplus P_7 \oplus P_8 \oplus K_{1,5} \oplus K_{1,7} \oplus K_{1,8} \oplus K_{2,6} = 0 \]
-By the Piling Up Principle this expression holds with bias:
-\[ 2 * 1/4 * 1/4 = 1/8 \]
-Note the implicit assumption that S-Boxes are independant. More on this later.
+\onslide<3->{By the Piling Up Principle this holds with bias
+\[ 2 \times 1/4 \times 1/4 = 1/8 \]}
+\end{column}
+
+\begin{column}{0.4\textwidth}
+\begin{figure}
+\includegraphics[totalheight=0.8\textheight]{spn.jpg}
+\end{figure}
+\end{column}
+\end{columns}
+
+\vspace{2mm}
+\onslide<4->{Note the implicit assumption that S-Boxes are independent. More on this later.}
+
 \end{frame}
@@
 \[ U_{4,6} \oplus U_{4,8} \oplus U_{4,14} \oplus U_{4, 16} \oplus P_5 \oplus P_7 \oplus P_8 = \Sigma K \]
-Where $\Sigma K$ is the sum over some key bits. Note that Since the key is fixed, $\Sigma K = 0$ or $1$ and thus we can ignore it since we only care about the bias.\\
+\pause Where $\Sigma K$ is the sum over some key bits. Note that Since the key is fixed, $\Sigma K = 0$ or $1$ and thus we can ignore it since we only care about the bias.\\
 \vspace{5mm}
-The magnitude of the bias of the above expression, by the Piling Up Principle, is $1/32$. \\
+\pause The magnitude of the bias of the above expression, by the Piling Up Principle, is $1/32$. \\
 \vspace{5mm}
-Next we show how we can extract key bits using this information.
+\pause Next we show how we can extract key bits using this information.
 \end{frame}
@@
 \vspace{3mm}
-We have an expression that links plaintext bits to input bits to the 4th round of SPN that holds with high bias ($1/32$). We can partially undo the last round by guessing the last key. \\
+We have an expression that links plaintext bits to input bits to the 4th round of SPN that holds with high bias ($1/32$). 
+\[ U_{4,6} \oplus U_{4,8} \oplus U_{4,14} \oplus U_{4, 16} \oplus P_5 \oplus P_7 \oplus P_8 = \Sigma K \]
+\pause We can partially undo the last round by guessing the last key. \\
 \vspace{3mm}
 \pause
-If we guess correctly, the equation will hold with high bias. If we guess wrong, the equation will probabily hold with probability close to 1/2, that is, a bias close to 0.\\
+If we guess correctly, the equation will hold with high bias. If we guess wrong, the equation will probably hold with probability close to 1/2, that is, a bias close to 0.\\
 \vspace{3mm}
 \pause
-BUT To do this, we don't need to guess the entire key for the last round! Our expression only involves 4 key bits, output from 2 S-Boxes. Thus we only need to guess $2^8 = 256$ values.\\
+BUT To do this, we don't need to guess the entire key for the last round!
+\pause Our expression only involves 4 fourth round input ($U_4$) bits, output from 2 S-Boxes of the third round.
+Thus we only need to guess $2^8 = 256$ values.\\
 \vspace{3mm}
@@
 \begin{frame}
-Probabalistic Justification:\\
+Probabilistic justification for number of pairs:\\
 \vspace{5mm}
-In general, the number of plaintext-ciphertext pairs needed is related inversely-quadratically to the bias. That is
+\pause In general, the number of plaintext-ciphertext pairs needed is related inversely-quadratically to the bias. That is
 \[ N_{pairs} \approx 1 / \epsilon^2 \]
-For the SPN cipher approximation with bias $1/32$ we need about $1000$ pairs. To perform the attack with near full probability of success.
+\pause For the SPN cipher approximation with bias $1/32$ we need about $1000$ pairs to perform the attack with near full probability of success.
+\end{frame}
+
+\begin{frame}
+Actual attack on SPN ... Demo ...
 \end{frame}
@@
 \begin{frame}
-Differential Cryptanalysis tries to exploit the high probability of certain occurrences of differential characteristics $(\Delta P \rightarrow \Delta C)$ in the cipher.
-This is a chosen ciphertext attack.
+Differential: $(\Delta P, \Delta C) = (P \oplus P', C \oplus C')$
 
-[Differential: $(\Delta P, \Delta C) = (P \oplus P', C \oplus C') $]
+\vspace{5mm}
+\pause Differential Cryptanalysis tries to exploit the high probability of certain occurrences of differential characteristics $(\Delta P \rightarrow \Delta C)$ in the cipher.
 
-We have to analyze the cipher in pieces. The basic idea is
-\begin{itemize}
+\vspace{2mm}
+\pause Note: This is a chosen plaintext attack.
+
+\vspace{5mm}
+\pause We have to analyze the cipher in pieces. \pause The basic idea is
+\begin{itemize}[<+->]
 \item{Analyze the difference pattern and their probabilities of the S-Box}
@@
 
-Algorithm
-\begin{enumerate}
+\vspace{2mm}
+\pause Algorithm
+\begin{enumerate}[<+->]
 \item{Fix a certain input difference $\Delta X = 0100$, say}
 \item{Iterate through the first input $X = \{0000, ... ,1111 \}$}
-\item{This fixes the second input $X' = \{0100, ... ,1011\}$ (ordering does not matter)}
+\item{This fixes the second input $X' = X \oplus \Delta X = \{0100, ... ,1011\}$ (ordering does not matter)}
 \item{Obtain corresponding output difference $\Delta Y = Y \oplus Y'$}
@@
 
-Distribution of differences
+\pause Note that, for $\Delta X = 1011$, $\Delta Y = 0010$ occurs $8$ times, out of the possible $16$ times. So, the pair $(1011,0010)$ has a probability of occurrence $8/16 = 1/2$.
 
-Note that, for $\Delta X = 1011$, $\Delta Y = 0010$ occurs $8$ times, out of the possible $16$ times.
-So, the difference pair $(1011,0010)$ has a probability of occurrence $\frac{8}{16} = \frac{1}{2}$.
-So, we can tabulate the complete data for the S-Box to check this probabilities.
+\vspace{2mm}
+\pause We can tabulate the complete data for the S-Box to check this probabilities.
 \end{frame}
@@
 
-Note that
-\begin{itemize}
+\pause Note that
+\begin{itemize}[<+->]
 \item{In an ideal S-Box, we would like all entries to be $1$}
 \item{But this table clearly shows bias towards certain pairs}
-\item{Just divide the entries by $2^4 = 16$ to get the probabilities, and exploit the scenario of highest probability of occurrence}
+\item{Just divide the entries by $2^4 = 16$ to get the probabilities, and exploit the scenario of highest probability of occurrence ($Pr(2|B) = 8/16 = 1/2$ here)}
 \end{itemize}
@@
 \begin{frame}
-Why can not we make an ideal S-Box?
+A few properties of the S-Box Distribution Table
 
-Notice that in the difference table
-\begin{itemize}
+\vspace{5mm}
+\pause Notice that in the difference table
+\begin{itemize}[<+->]
 \item{The $(0,0)$ entry is $16$, just because identical inputs ($\Delta X = 0 \Rightarrow X = X'$) should produce identical outputs ($Y = Y' \Rightarrow \Delta Y = 0$)}
@@
 
-As an ideal S-Box expects to give out no information of $\Delta Y$ given $\Delta X$, we wish it could have all entries $1$, i.e,
-\[ Pr(\Delta Y | \Delta X) = \frac{1}{16} \quad \forall \Delta X  \]
-But, from our discussion about the distribution table, this is infeasible.
+\vspace{5mm}
+\pause As an ideal S-Box expects to give out no information of $\Delta Y$ given $\Delta X$, we wish it could have all entries $1$, i.e,
+\[ Pr(\Delta Y | \Delta X) = \frac{1}{16} \quad \forall \: \Delta X  \]
+\pause But, from our discussion above, this is infeasible, and we hope to exploit this.
 \end{frame}
@@
 
-Let us consider the S-Box of the SPN cipher, with the round keys this time:
+\vspace{5mm}
+\pause Let us consider the S-Box of the SPN cipher, with the round keys this time.
+\pause
 \begin{columns}
@@
 
-Note that for each pair of inputs $(W,W')$, the actual inputs to the S-Box are $(X,X') = (W \oplus K, W' \oplus K)$. Hence, the input difference for the keyed S-Box is 
+\vspace{5mm}
+\pause Note that for each pair of inputs $(W,W')$, the actual inputs to the S-Box are $(X,X') = (W \oplus K, W' \oplus K)$. Hence, the input difference for the keyed S-Box is 
 \[ \Delta W = W \oplus W' = (X \oplus K) \oplus (X' \oplus K) = X \oplus X' = \Delta X \]
-and hence $\Delta W \rightarrow \Delta Y$ corresponds identically to $\Delta X \rightarrow \Delta Y$.
+\pause and hence $\Delta W \rightarrow \Delta Y$ corresponds identically to $\Delta X \rightarrow \Delta Y$.
 
-Thus, the keyed S-Box has the same difference distribution table as the unkeyed S-Box.
+\vspace{2mm}
+\pause Thus, the keyed S-Box has the same difference distribution table.
 \end{frame}
@@
 \begin{frame}
+
+\begin{columns}
+\begin{column}{0.4\textwidth}
+\begin{figure}
+\includegraphics[totalheight=0.8\textheight]{spn_diff_char.jpg}
+\end{figure}
+\end{column}
+
+\begin{column}{0.6\textwidth}
 Constructing a Differential Characteristic
 
-\begin{columns}
-\begin{column}{0.6\textwidth}
+\vspace{5mm}
+\pause The diagram beside illustrates the tracing of the non-zero bits of the input difference through the SPN structure.
 
-The diagram beside illustrates the tracing of the non-zero bits of the input difference through the SPN structure.
+\vspace{2mm}
+Input Difference: $\Delta P = 0B00$
 
-Input Diff: $\Delta P = 0B00$ \\
-
-Propagating Difference Pairs
-\begin{itemize}
+\vspace{2mm}
+\pause Propagating Difference Pairs
+\begin{itemize}[<+->]
 \item{$S_{12}$: $1011 \rightarrow 0010$ [$Pr = \frac{8}{16}$]}
@@
 
-Differential Characteristic
+\vspace{2mm}
+\pause Differential Characteristic
 \[ 0B00 \rightarrow 0040 \rightarrow 0220 \rightarrow 0606 \]
-\end{column}
-\begin{column}{0.4\textwidth}
-\begin{figure}
-\includegraphics[totalheight=0.8\textheight]{spn_diff_char.jpg}
-\end{figure}
 \end{column}
@@
 
-Notation
-$\Delta U_i$: Input difference to $i$-th round
-$\Delta V_i$: Output difference from $i$-th round
-$K_i$: Round Key for $i$-th round
-$X_{i,j}$: $j$-th bit of $X_i$
+\vspace{2mm}
+\pause Notation
+\begin{itemize}
+\item{$\Delta U_i$: Input difference to $i$-th round}
+\item{$\Delta V_i$: Output difference from $i$-th round}
+\item{$K_i$: Round Key for $i$-th round}
+\item{$X_{i,j}$: $j$-th bit of $X_i$}
+\end{itemize}
 
-Here, we have the following in the differential characteristic
-\begin{itemize}
-\item{Round 1: $\Delta P = \Delta U_1 = 0B00 \rightarrow 0040 = \Delta V_1$ [$Pr = \frac{8}{16}$]}
-\item{Round 2: $\Delta V_1 = \Delta U_2 = 0040 \rightarrow 0220 = \Delta V_2$ [$Pr = \frac{6}{16}$]}
-\item{Round 3: $\Delta V_2 = \Delta U_3 = 0220 \rightarrow 0606 = \Delta V_3$ [$Pr = \left(\frac{6}{16}\right)^2$]}
+\vspace{2mm}
+\pause Here, we have the following in the differential characteristic
+\begin{itemize}[<+->]
+\item{Round 1: $\Delta P = \Delta U_1 = 0B00 \rightarrow 0040 = \Delta V_1$  [$Pr = 8/16$]}
+\item{Round 2: $\Delta V_1 = \Delta U_2 = 0040 \rightarrow 0220 = \Delta V_2$  [$Pr = 6/16$]}
+\item{Round 3: $\Delta V_2 = \Delta U_3 = 0220 \rightarrow 0606 = \Delta V_3$  [$Pr = \left(6/16\right)^2$]}
 \item{Round 4: $\Delta V_3 = \Delta U_4 = 0606$}
@@
 
-Hence, assuming the S-Boxes to be independent over the rounds, we obtain
+\vspace{2mm}
+\pause Hence, assuming the S-Boxes to be independent over the rounds, we obtain
 \[ \Delta P = 0B00 \rightarrow 0606 = \Delta U_4 \]
-with probability $Pr(\Delta U_4 = 0606 | \Delta P = 0B00) = \frac{8}{16} \times \frac{6}{16} \times \left(\frac{6}{16}\right)^2 = \frac{27}{1024}$.
+with probability $Pr(\Delta U_4 | \Delta P) = 8/16 \times 6/16 \times \left(6/16\right)^2 = 27/1024$.
 
-Note: The probability of this characteristic $Pr = \frac{27}{2^{10}} \gg \frac{1}{2^{16}}$
+\vspace{2mm}
+\pause Note: The probability of this characteristic $Pr = 27/1024 \gg 1/2^{16}$
 
-Now, lets see if we can exploit this situation to extract the key bits.
 \end{frame}
@@
 Attack Idea
-\begin{itemize}
-\item{We have a differential characteristic that links plaintext difference $\Delta P$ to input difference $\Delta U_4$ of the 4th round output of SPN and holds with high probability ($\frac{27}{1024}$)}
+\begin{itemize}[<+->]
+\item{We have a differential characteristic that links plaintext difference $\Delta P$ to input difference $\Delta U_4$ of the 4th round output of SPN and holds with high probability ($27/1024$)}
 \item{We can partially undo the last round by guessing the last key $K_5$}
-\item{If we guess correctly, the differential characteristic will hold with high probability, close to $\frac{27}{1024}$. If we guess wrong, it will hold with low probability}
+\item{If we guess correctly, the differential characteristic will hold with high probability, close to $27/1024$. If we guess wrong, it will hold with low probability}
 \item{But in this case, we don't need to guess the entire key $K_5$ for the last round!}
@@
 \begin{columns}
-\begin{column}{0.6\textwidth}
-Differential Characteristic flow
-
-Input Diff: $\Delta P = 0B00$
-
-Characteristic
-\[ \Delta P = 0B00 \rightarrow 0606 = \Delta U_4 \]
-
-Probability $Pr(\Delta U_4 = 0606 | \Delta P = 0B00) = \frac{27}{1024}$
-
-Partial Subkey guess only for
-\[ [K_{5,5} ... K_{5,8}, K_{5,13} ... K_{5,16}] \]
-\end{column}
-
 \begin{column}{0.4\textwidth}
@@
 \end{column}
+
+\begin{column}{0.6\textwidth}
+Differential Characteristic flow
+
+\vspace{5mm}
+\pause Input Difference: $\Delta P = 0B00$
+
+\vspace{2mm}
+\pause Characteristic
+\[ \Delta P = 0B00 \rightarrow 0606 = \Delta U_4 \]
+
+\vspace{2mm}
+Probability $Pr(\Delta U_4 = 0606 | \Delta P = 0B00) = 27/1024$
+
+\vspace{5mm}
+\pause Partial Subkey guess only for
+\[ [K_{5,5} ... K_{5,8}, K_{5,13} ... K_{5,16}] \]
+\end{column}
 \end{columns}
@@
 We can use the same technique as before.
-For a given S-Box $a \in \{1, 2, \dots 8\}, 1 \leq \alpha \leq 63, 1 \leq \beta \leq 15$ we define $NS_{a}(\alpha, \beta)$ as the number of times (out of 64) for S-Box $a$ that for all the input patterns masked by $\alpha$ the output pattern masked by $\beta$ agrees with the value of S-Box $a$.
+For a given S-Box $a \in \{1, 2, \dots 8\}$, $1 \leq \alpha \leq 63$, $1 \leq \beta \leq 15$, we define $NS_{a}(\alpha, \beta)$ as the number of times (out of 64) for S-Box $a$ that for all the input patterns masked by $\alpha$ the output pattern masked by $\beta$ agrees with the value of S-Box $a$.
 \[ NS_{a}(\alpha, \beta) = | \{ X | 0 \leq X < 64, ( \bigoplus_{i=0}^5 X[i]\cdot\alpha[i]) = (\bigoplus_{j=0}^3 S_a(X)[j] \cdot \beta[j])\} | \]
-This looks more complicated, but is infact the same expression as for SPN, except it takes into account all possible S-Boxes.
+This looks more complicated, but is in fact the same expression as for SPN, except it takes into account all possible S-Boxes.
 \end{frame}
@@
 \begin{frame}
-We perform the attack on the 4-round DES similiarly to how we performed the attack on 4-round SPN, by "undoing" the last round.\\
+We perform the attack on the 4-round DES similarly to how we performed the attack on 4-round SPN, by ``undoing'' the last round.\\
 \vspace{5mm}
 \pause
-Note that in the last round, $C_L$ bits are not changed, only $C_R$ bits are affected. Note also that there is only one $C_R$ bit the 3-round equation.\\
+Note that in the last round, $U_{4L}$ bits are not changed, only $U_{4R}$ bits are affected. Note also that there is only one $U_R$ bit the 3-round equation.
+\begin{align*} 
+P_L[25] \oplus P_L[14] \oplus P_L[8] \oplus & P_L[3] \oplus U_{4L}[25] \oplus U_{4L}[14] \oplus U_{4L}[8] \\ & \oplus U_{4L}[3] \oplus P_R[17] \oplus U_{4R}[17] = K_1[26] \oplus K_3[26]
+\end{align*}
 \vspace{5mm}
 \pause
-This bit, $C_R{17}$ is affected by the result of S-Box 1 in round 4.
+This bit, $U_{4R}[17]$ is affected by the result of S-Box 1 in round 4.
 If we guess correctly for the 6 partial subkey bits $K_{4,1}$ to $K_{4,6}$, the 3-round equation will hold with high bias, otherwise it will likely hold with close to 0 bias.\\
@@
 Let $d = 1$ and $l_1 = 17$ in the above expression - this corresponds to our attack (distribution over guessing for one bit, $CR_{17}$ in the equation).
-The following table summarizes the sucess rates:
+The following table summarizes the success rates:
 \begin{center}
@@
 \begin{frame}
-How many pairs do we need? Continued\\
+How many pairs do we need? (Continued)
+
 \vspace{3mm}
 Using the calculation from the previous slide, we see that for a 4-round DES, our equation holds with bias of about 1 in 4.
-Thus we will need about $16\times 4^2 = 256$ pairs for the attack to have near perfect chance of sucess.\\
+Thus we will need about $16\times 4^2 = 256$ pairs for the attack to have near perfect chance of success.\\
 \vspace{3mm}
@@
 
-\subsection{Differential Attack: 3-round DES}
 \begin{frame}
-Differential Attack idea ...
+The actual Attack on 4 round DES ... Demo ...
+\end{frame}
+
+\subsection{Differential Attack: 4-round DES}
+\begin{frame}
+Differential Attack Idea
+
+\vspace{5mm}
+Similar to the differential attack on SPN, here we care about the differential characteristics of the cipher. So, the basic idea is as follows
+
+\begin{itemize}
+\item{Construct the difference tables for the S-Boxes}
+\item{Find the most probable difference pairs}
+\item{Combine the pairs through all the rounds to get a highly probable differential characteristic until the penultimate round}
+\item{Exploit this differential characteristic by choosing a number of pairs of plaintexts with the desired difference and undo the last round by guessing the keybits}
+\item{If the initial choice of the input difference was smart, we will only need to guess a few keybits}
+\item{The guess approving the differential characteristic with highest probability is correct}
+\end{itemize}
 
@@
 \begin{frame}
-Differentials for SBoxes ...
+The chosen input difference:$\Delta P = 20 00 00 00 00 00 00 00$
+
+Round 1 Differential
+
+\[ \Delta P = 20 00 00 00 00 00 00 00 \rightarrow 20 00 00 00 00 00 00 00 = \Delta V_1 \]
+Probability: $p_1 = 1$ as Feistel gives output difference $0$ for input $0$
+
+Round 2 Differential
+
+\[ \Delta U_2 = \Delta V_1 = 20 00 00 00 00 00 00 00 \rightarrow \]
+
+this calculations and coding has not been completed yet ... will be done ...
+\end{frame}
+
+\begin{frame}
+Differentials for SBoxes
 
@@
 \begin{frame}
-Differential characteristics for SBoxes ...
-
-\end{frame}
-
-\begin{frame}
-Piling up over 3 rounds ...
-
-\end{frame}
-
-\begin{frame}
-Demo of the attack ...
-
+The actual Attack on 4 round DES ... Demo ...
 \end{frame}
@@
 \begin{frame}
-Note that throughout this discussion we have assumed that the S-Boxes were independant. This allowed us to use the Piling Up Principle to combine linear approximations of S-Boxes across rounds. \\
+Note that throughout this discussion we have assumed that the S-Boxes were independent. This allowed us to use the Piling Up Principle to combine linear approximations of S-Boxes across rounds. \\
 \vspace{3mm}

### 2009-01-31 talk/draft1.tex
Slide 18: Permutation provides [confusion] -> diffusion

@@
 \item{DES Unit Block functions \\ $\qquad \sigma_{i+1} (L_i,R_i)$ for $i = 0,1,...,15$}
-\item{Transpositions \\ $\qquad \tau : (L_{i+1},R_{i+1}) \mapsto (R_{i+1},L_{i+1})$ for $i = 0,1,...,14$}
+\item{Transposition \\ $\qquad \tau : (L_{i+1},R_{i+1}) \mapsto (R_{i+1},L_{i+1})$ for $i = 0,1,...,14$}
 \item{Inverse Initial Permutation \\ $\qquad IP^{-1} : (0,1)^{64} \rightarrow (0,1)^{64}$}
@@
 
+\pause
 \begin{block}{Algebraic Representation of DES}
-$\qquad C = DES_K (P) = IP^{-1} \sigma_{16} \tau \cdots \tau \sigma_1 IP (P)$
-
-$\qquad P = DES^{-1}_K (C) = IP^{-1} \sigma_{1} \tau \cdots \tau \sigma_{16} IP (C)$
+$\qquad C \:  = \:  DES_K (P) \: = \: IP^{-1} \: \sigma_{16} \:  \tau \: \cdots \:  \tau \:  \sigma_1 \: IP \: (\: P \: )$
+
+$\qquad P \: = \: DES^{-1}_K (C) \: = \: IP^{-1} \: \sigma_{1} \: \tau \: \cdots \: \tau \: \sigma_{16} \: IP \: (\: C\: )$
 \end{block}
 
+\pause
 \vspace{0.2in}
@@
 
-\begin{itemize}
+\begin{itemize}[<+->]
 \item S-Boxes depend on a relatively small number of bits (6) which allows us to write down linear (or affine) expressions that approximate S-Boxes
@@
 \begin{frame}
-Specifically, if $P_{i}$ are plaintext bits, $C_{i}$ are cyphertext bits, and $K_{i}$ are subkey bits, then we wish to find an expression of the form
+Specifically, if $P_{i}$ are plaintext bits, $C_{i}$ are ciphertext bits, and $K_{i}$ are subkey bits, then we wish to find an expression of the form
 \[ P_{i_1} \oplus P_{i_2} \cdots P_{i_j} \oplus C_{i_1} \oplus C_{i_2} \cdots C_{i_k} = K_{i_1} \oplus K_{i_2} \oplus K_{i_m} \]
-Such that this expression has a high \textit{or} low probability of occurence.\\
-\vspace{5mm}
-Consider: No such obvious expression should exist, otherwise the cipher is trivially weak.
-If we were to randomly select bits for the above expression, it would hold exactly $1/2$ the time.
-If we find an expression such as above that displays a high \textit{bias}, that is, it holds much more or less freqently than $1/2$ the time, we can exploit this.
+\pause such that this expression has a high \textit{or} low probability of occurrence.
+
+\vspace{5mm}
+\pause Consider: No such obvious expression should exist, otherwise the cipher is trivially weak. If we were to randomly select bits for the above expression, it would hold exactly $1/2$ the time.
+
+\vspace{5mm}
+\pause If we find an expression such as above that displays a high \textit{bias}, that is, it holds much more or less frequently than $1/2$ the time, we can exploit this.
 \end{frame}
@@
 \vspace{0.2in}
-Vulnerability of DES:
-
-\begin{itemize}
+\pause Vulnerability of DES:
+
+\begin{itemize}[<+->]
 \item{S-Boxes show a strong tendency to produce certain differential pairs with high probability, rather than being random}
 \item{The differences do not get diffused fast enough through the permutations}
-\item{Differentials are not affected by the round keys as they get xor-ed out}
-\end{itemize}
-\end{frame}
-
-\begin{frame}
-In an $k$-round DES, we can trace the path of a certain input difference to get an output difference with a high probability. This tracing path is known as a {\it differential characteristic}.
+\item{Differentials are not affected by the round keys as they get XOR-ed out}
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+In an $k$-round DES, we can trace the path of a certain input difference to get an output difference with a high probability. \pause This tracing path is known as a {\it differential characteristic}.
 
@@
 \begin{frame}
-We define a simple Substitution-Permutation Network Cipher (SPN) which has structural similarity to DES.
-
-\begin{itemize}
+We construct a simple Substitution-Permutation Network Cipher (SPN) which has structural similarity to DES.
+
+\begin{itemize}[<+->]
 \item SPN is a 16-bit block cipher (16-bit plaintext to 16-bit ciphertext) with 16-bit round keys
@@
 \vspace{5mm}
-The S-Box provides the {\it confusion} function and the permutation applies the {\it confusion} operation in SPN, thus making it cryptographically similar to DES.
+\pause The S-Box provides the {\it confusion} function and the permutation applies the {\it diffusion} operation in SPN, thus making it cryptographically similar to DES.
 \vspace{5mm}
@@
 \begin{frame}
-Basic Definitions for the Linear attack:
+Basic Definitions for the Linear attack
 
 \vspace{5 mm}
-Linear cryptanalysis tries to take advantage of high probability occurrences of linear expressions involving plaintext bits, ciphertext bits and subkey bits. It is a {\it known plaintext attack}.
+\pause Linear cryptanalysis tries to take advantage of high probability occurrences of linear expressions involving plaintext bits, ciphertext bits and subkey bits.
+It is a {\it known plaintext attack}.
 
 \vspace{5 mm}
-Define $P_i$ where $i = 1, 2, \dots 16$ as the $i$-th plaintext bit.\\
-Define $C_i$ where $i = 1, 2, \dots 16$ as the $i$-th cyphertext bit.\\
-Define $U_{j,i}$ where $j = 1, 2, 3, 4$ and $i = 1, 2, \dots 16$ be the $i$-th input bit to the $j$-th round of SPN.\\
-Define $V_{j,i}$ where $j = 1, 2, 3, 4$ and $i = 1, 2, \dots 16$ be the $i$-th output bit of the $j$-th round of SPN.\\
-\end{frame}
-
-\begin{frame}
-Linear and Affine approximation of S-Box:
-
-\vspace{5mm}
-\begin{itemize}
+\pause Define 
+\begin{itemize}[<+->]
+\item{$P_i$ where $i = 1, 2, \dots 16$ as the $i$-th plaintext bit}
+\item{$C_i$ where $i = 1, 2, \dots 16$ as the $i$-th ciphertext bit}
+\item{$U_{j,i}$ where $j = 1, 2, 3, 4$ and $i = 1, 2, \dots 16$ be the $i$-th input bit to the $j$-th round of SPN}
+\item{$V_{j,i}$ where $j = 1, 2, 3, 4$ and $i = 1, 2, \dots 16$ be the $i$-th output bit of the $j$-th round of SPN}
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+Linear and Affine approximation of S-Box
+
+\vspace{5mm}
+\begin{itemize}[<+->]
 \item Question: How do we come up with the desired expression for the entire cipher? 
@@
 \vspace{5mm}
-To find the linear/affine approximation of the S-Box we simply consider every possible expression of the input bits $X_i$ and output bits $Y_j$. Thus the expression has the form
+\pause To find the linear/affine approximation of the S-Box we simply consider every possible expression of the input bits $X_i$ and output bits $Y_j$. \pause Thus the expression has the form
 \[ \bigoplus_{i \in U} X_i = \bigoplus_{j \in V} Y_j \]
-Where $U$ and $V$ range over all possible subsets of $\{1, 2, 3 ,4\}$. We then compare how often this expression coincides with the S-Box.
-
-\vspace{5mm}
-Note that there are 16 possibilities for $U$ and $V$, hence 256 total possible expressions (for a 4-bit S-box).
+where $U$ and $V$ range over all possible subsets of $\{1, 2, 3 ,4\}$. \pause We then compare how often this expression coincides with the S-Box.
+
+\vspace{5mm}
+\pause Note that there are 16 possibilities for $U$ and $V$, hence 256 total possible expressions (for a 4-bit S-box).
 \end{frame}
@@
 \begin{columns}
+\begin{column}{0.3\textwidth}
+\begin{figure}
+\includegraphics[width=0.7\textwidth]{spn_sbox_unkeyed.jpg}
+\end{figure}
+\end{column}
+
 \begin{column}{0.7\textwidth}
@@
 \[ X_2 \oplus X_3 = Y_1 \oplus Y_3 \oplus Y_4 \]
-Applying all possible values for the input $X$ bits it turns out that the expression holds in 12 out of the 16 cases. Hence, this expression has a bias of $12/16 - 1/2 = 1/4$.
-\end{column}
-
-\begin{column}{0.3\textwidth}
-\begin{figure}
-\includegraphics[width=0.7\textwidth]{spn_sbox_unkeyed.jpg}
-\end{figure}
-\end{column}
-\end{columns}
-
-\vspace{5mm}
-Similarly the following expression
+\pause Applying all possible values for the input $X$ bits it turns out that the expression holds in 12 out of the 16 cases. \pause Hence, this expression has a bias of $12/16 - 1/2 = 1/4$.
+\end{column}
+\end{columns}
+
+\vspace{5mm}
+\pause Similarly the following expression
 \[ X_3 \oplus X_4 = Y_1 \oplus Y_4 \]
-Has a bias of $-3/8$, which is higher in magnitude than the previous one. It is the magnitude of the bias that is important.
+Has a bias of $-3/8$, which is higher in magnitude than the previous one. \pause It is the magnitude of the bias that is important.
 \end{frame}
@@
 \vspace{5mm}
-Note also that $U_1 = P \oplus K_1$. 
+\onslide<2->{Note also that $U_1 = P \oplus K_1$.}
 \end{column}
@@
 \vspace{5mm} 
-We can now write down the following linear approximation across the 1st round of SPN:
+\onslide<3->{We can now write down the following linear approximation across the 1st round of SPN:
 \begin{eqnarray*}
-V_{1,6} & = & U_{1,5} \oplus U_{1,7} \oplus U_{1,8} \mbox{  $\leftarrow$ S-Box approximation above}\\
+V_{1,6} & = & U_{1,5} \oplus U_{1,7} \oplus U_{1,8} \quad \mbox{  $\leftarrow$ S-Box $S_{12}$ approximation above}\\
 		  & = & (P_5\oplus K_{1,5}) \oplus (P_7\oplus K_{1,7}) \oplus (P_8\oplus K_{1,8})
@@
 
-This expression holds with probability of $3/4$ (bias of $+1/4$)
+\pause This expression holds with probability of $3/4$ (bias of $+1/4$)}
 \end{frame}
@@
 But we must somehow combine them to write an expression relating the plaintext and ciphertext bits.
-We use the Piling Up Principle for this purpose.
-
-\vspace{5mm}
-Note:
+\pause We use the Piling Up Principle for this purpose.
+
+\vspace{5mm}
+\pause Note:
 
@@
 \vspace{5mm}
-Assume the following probability distribution
+\pause Assume the following probability distribution
 \[ Pr(X_1 = i)  = \left\{ \begin{array}{ll}
@@
 			(1 - p_1)(1 - p_2) & \mbox{$i = 1, j = 1$}.\end{array} \right. \] 
-Now, note that $X_1 \oplus X_2 = 0 \Rightarrow X_1 = 1, X_2 =1$ OR $X_1 = 0, X_2 = 0$. Hence,
+\pause Now, note that $X_1 \oplus X_2 = 0 \Rightarrow X_1 = 1, X_2 =1$ OR $X_1 = 0, X_2 = 0$. \pause Hence,
 \begin{eqnarray*}
@@
 \vspace{5mm}
-We extend the idea on the previous slide to the general case: For $n$ independant random binary variables, $X_1, X_2, \dots, X_n$, 
+We extend the idea on the previous slide to the general case \pause : For $n$ independent random binary variables, $X_1, X_2, \dots, X_n$, 
 \[ Pr(X_1 \oplus X_2 \cdots \oplus X_n = 0) = 1/2 + 2^{n-1} \prod_{i=1}^n \epsilon_i \]
-In terms of bias,
+\pause In terms of bias,
 \[ \epsilon_{1,2,\dots,n} = 2^{n-1} \prod_{i = 1}^n \epsilon_i \]
@@
 \vspace{5mm}
-Note that if some $\epsilon_k = 0$ (that is, some $X_k$ has no bias) then the total bias will be $0$. More about the probabilistic analysis later.
+\pause Note that if some $\epsilon_k = 0$ (that is, some $X_k$ has no bias) then total bias will be $0$. More about the probabilistic analysis later.
 \end{frame}
@@
 \begin{column}{0.6\textwidth}
-Now, come back to per-round approximations. We can write down 4 approximations ($S_{ij}$ repesents the $j$-th S-Box in the $i$-th round):
+Now, come back to per-round approximations. We can write down 4 approximations ($S_{ij}$ represents the $j$-th S-Box in the $i$-th round):
 \begin{align*}
@@
 \end{align*}
-Each of these has a probability bias magnitude of $1/4$. We can use the Piling Up Principle to combine them into a single expression relating plaintext bits to cyphertext bits.
+\onslide<2->{Each of these has a probability bias magnitude of $1/4$. We can use the Piling Up Principle to combine them into a single expression relating plaintext bits to ciphertext bits.}
 \end{column}
@@
 \vspace{2mm}
-Each of these has a bias of magnitude $1/4$ and we can combine to obtain:
-\[ V_{2,6} \oplus V_{2,8} \oplus P_5 \oplus P_7 \oplus P_8 \oplus K_{1,5} \oplus K_{1,7} \oplus K_{1,8} \oplus K_{2,6} = 0 \]
-By the Piling Up Principle this holds with bias
-\[ 2 \times 1/4 \times 1/4 = 1/8 \]
+\onslide<2->{Each of these has a bias of magnitude $1/4$ and we can combine to obtain:
+\[ V_{2,6} \oplus V_{2,8} \oplus P_5 \oplus P_7 \oplus P_8 \oplus K_{1,5} \oplus K_{1,7} \oplus K_{1,8} \oplus K_{2,6} = 0 \]}
+
+\vspace{2mm}
+\onslide<3->{By the Piling Up Principle this holds with bias
+\[ 2 \times 1/4 \times 1/4 = 1/8 \]}
 \end{column}
@@
 \vspace{2mm}
-Note the implicit assumption that S-Boxes are independent. More on this later.
+\onslide<4->{Note the implicit assumption that S-Boxes are independent. More on this later.}
 
@@
 \[ U_{4,6} \oplus U_{4,8} \oplus U_{4,14} \oplus U_{4, 16} \oplus P_5 \oplus P_7 \oplus P_8 = \Sigma K \]
-Where $\Sigma K$ is the sum over some key bits. Note that Since the key is fixed, $\Sigma K = 0$ or $1$ and thus we can ignore it since we only care about the bias.\\
-\vspace{5mm}
-The magnitude of the bias of the above expression, by the Piling Up Principle, is $1/32$. \\
-\vspace{5mm}
-Next we show how we can extract key bits using this information.
+\pause Where $\Sigma K$ is the sum over some key bits. Note that Since the key is fixed, $\Sigma K = 0$ or $1$ and thus we can ignore it since we only care about the bias.\\
+\vspace{5mm}
+\pause The magnitude of the bias of the above expression, by the Piling Up Principle, is $1/32$. \\
+\vspace{5mm}
+\pause Next we show how we can extract key bits using this information.
 \end{frame}
@@
 \[ U_{4,6} \oplus U_{4,8} \oplus U_{4,14} \oplus U_{4, 16} \oplus P_5 \oplus P_7 \oplus P_8 = \Sigma K \]
-We can partially undo the last round by guessing the last key. \\
+\pause We can partially undo the last round by guessing the last key. \\
 \vspace{3mm}
 \pause
-If we guess correctly, the equation will hold with high bias. If we guess wrong, the equation will probabily hold with probability close to 1/2, that is, a bias close to 0.\\
+If we guess correctly, the equation will hold with high bias. If we guess wrong, the equation will probably hold with probability close to 1/2, that is, a bias close to 0.\\
 \vspace{3mm}
@@
 BUT To do this, we don't need to guess the entire key for the last round!
-Our expression only involves 4 fourth round input ($U_4$) bits, output from 2 S-Boxes of the third round.
+\pause Our expression only involves 4 fourth round input ($U_4$) bits, output from 2 S-Boxes of the third round.
 Thus we only need to guess $2^8 = 256$ values.\\
@@
 \begin{frame}
-Probabalistic Justification:\\
-\vspace{5mm}
-In general, the number of plaintext-ciphertext pairs needed is related inversely-quadratically to the bias. That is
+Probabilistic justification for number of pairs:\\
+\vspace{5mm}
+\pause In general, the number of plaintext-ciphertext pairs needed is related inversely-quadratically to the bias. That is
 \[ N_{pairs} \approx 1 / \epsilon^2 \]
-For the SPN cipher approximation with bias $1/32$ we need about $1000$ pairs. To perform the attack with near full probability of success.
+\pause For the SPN cipher approximation with bias $1/32$ we need about $1000$ pairs to perform the attack with near full probability of success.
 \end{frame}
@@
 \begin{frame}
-Differential Cryptanalysis tries to exploit the high probability of certain occurrences of differential characteristics $(\Delta P \rightarrow \Delta C)$ in the cipher.
-This is a chosen plaintext attack.
-
-\vspace{2mm}
 Differential: $(\Delta P, \Delta C) = (P \oplus P', C \oplus C')$
@@
 \vspace{5mm}
-We have to analyze the cipher in pieces. The basic idea is
-\begin{itemize}
+\pause Differential Cryptanalysis tries to exploit the high probability of certain occurrences of differential characteristics $(\Delta P \rightarrow \Delta C)$ in the cipher.
+
+\vspace{2mm}
+\pause Note: This is a chosen plaintext attack.
+
+\vspace{5mm}
+\pause We have to analyze the cipher in pieces. \pause The basic idea is
+\begin{itemize}[<+->]
 \item{Analyze the difference pattern and their probabilities of the S-Box}
@@
 \vspace{2mm}
-Algorithm
-\begin{enumerate}
+\pause Algorithm
+\begin{enumerate}[<+->]
 \item{Fix a certain input difference $\Delta X = 0100$, say}
@@
 
-Note that, for $\Delta X = 1011$, $\Delta Y = 0010$ occurs $8$ times, out of the possible $16$ times. So, the pair $(1011,0010)$ has a probability of occurrence $8/16 = 1/2$.
-
-\vspace{2mm}
-We can tabulate the complete data for the S-Box to check this probabilities.
+\pause Note that, for $\Delta X = 1011$, $\Delta Y = 0010$ occurs $8$ times, out of the possible $16$ times. So, the pair $(1011,0010)$ has a probability of occurrence $8/16 = 1/2$.
+
+\vspace{2mm}
+\pause We can tabulate the complete data for the S-Box to check this probabilities.
 \end{frame}
@@
 
-Note that
-\begin{itemize}
+\pause Note that
+\begin{itemize}[<+->]
 \item{In an ideal S-Box, we would like all entries to be $1$}
@@
 \vspace{5mm}
-Notice that in the difference table
-\begin{itemize}
+\pause Notice that in the difference table
+\begin{itemize}[<+->]
 \item{The $(0,0)$ entry is $16$, just because identical inputs ($\Delta X = 0 \Rightarrow X = X'$) should produce identical outputs ($Y = Y' \Rightarrow \Delta Y = 0$)}
@@
 \vspace{5mm}
-As an ideal S-Box expects to give out no information of $\Delta Y$ given $\Delta X$, we wish it could have all entries $1$, i.e,
+\pause As an ideal S-Box expects to give out no information of $\Delta Y$ given $\Delta X$, we wish it could have all entries $1$, i.e,
 \[ Pr(\Delta Y | \Delta X) = \frac{1}{16} \quad \forall \: \Delta X  \]
-But, from our discussion above, this is infeasible, and we hope to exploit this.
+\pause But, from our discussion above, this is infeasible, and we hope to exploit this.
 \end{frame}
@@
 \vspace{5mm}
-Let us consider the S-Box of the SPN cipher, with the round keys this time.
+\pause Let us consider the S-Box of the SPN cipher, with the round keys this time.
+\pause
 \begin{columns}
@@
 \vspace{5mm}
-Note that for each pair of inputs $(W,W')$, the actual inputs to the S-Box are $(X,X') = (W \oplus K, W' \oplus K)$. Hence, the input difference for the keyed S-Box is 
+\pause Note that for each pair of inputs $(W,W')$, the actual inputs to the S-Box are $(X,X') = (W \oplus K, W' \oplus K)$. Hence, the input difference for the keyed S-Box is 
 \[ \Delta W = W \oplus W' = (X \oplus K) \oplus (X' \oplus K) = X \oplus X' = \Delta X \]
-and hence $\Delta W \rightarrow \Delta Y$ corresponds identically to $\Delta X \rightarrow \Delta Y$.
-
-\vspace{2mm}
-Thus, the keyed S-Box has the same difference distribution table.
-\end{frame}
-
-\begin{frame}
-
-\begin{columns}
+\pause and hence $\Delta W \rightarrow \Delta Y$ corresponds identically to $\Delta X \rightarrow \Delta Y$.
+
+\vspace{2mm}
+\pause Thus, the keyed S-Box has the same difference distribution table.
+\end{frame}
+
+\begin{frame}
+
+\begin{columns}
+\begin{column}{0.4\textwidth}
+\begin{figure}
+\includegraphics[totalheight=0.8\textheight]{spn_diff_char.jpg}
+\end{figure}
+\end{column}
+
 \begin{column}{0.6\textwidth}
@@
 \vspace{5mm}
-The diagram beside illustrates the tracing of the non-zero bits of the input difference through the SPN structure.
-
-\vspace{2mm}
-Input Diff: $\Delta P = 0B00$ \\
-
-\vspace{2mm}
-Propagating Difference Pairs
-\begin{itemize}
+\pause The diagram beside illustrates the tracing of the non-zero bits of the input difference through the SPN structure.
+
+\vspace{2mm}
+Input Difference: $\Delta P = 0B00$
+
+\vspace{2mm}
+\pause Propagating Difference Pairs
+\begin{itemize}[<+->]
 \item{$S_{12}$: $1011 \rightarrow 0010$ [$Pr = \frac{8}{16}$]}
@@
 \vspace{2mm}
-Differential Characteristic
+\pause Differential Characteristic
 \[ 0B00 \rightarrow 0040 \rightarrow 0220 \rightarrow 0606 \]
 \end{column}
-\begin{column}{0.4\textwidth}
-\begin{figure}
-\includegraphics[totalheight=0.8\textheight]{spn_diff_char.jpg}
-\end{figure}
-\end{column}
 \end{columns}
@@
 \vspace{2mm}
-Notation
+\pause Notation
 \begin{itemize}
@@
 \vspace{2mm}
-Here, we have the following in the differential characteristic
-\begin{itemize}
+\pause Here, we have the following in the differential characteristic
+\begin{itemize}[<+->]
 \item{Round 1: $\Delta P = \Delta U_1 = 0B00 \rightarrow 0040 = \Delta V_1$  [$Pr = 8/16$]}
@@
 \vspace{2mm}
-Hence, assuming the S-Boxes to be independent over the rounds, we obtain
+\pause Hence, assuming the S-Boxes to be independent over the rounds, we obtain
 \[ \Delta P = 0B00 \rightarrow 0606 = \Delta U_4 \]
@@
 \vspace{2mm}
-Note: The probability of this characteristic $Pr = 27/1024 \gg 1/2^{16}$
+\pause Note: The probability of this characteristic $Pr = 27/1024 \gg 1/2^{16}$
 
@@
 Attack Idea
-\begin{itemize}
+\begin{itemize}[<+->]
 \item{We have a differential characteristic that links plaintext difference $\Delta P$ to input difference $\Delta U_4$ of the 4th round output of SPN and holds with high probability ($27/1024$)}
@@
 \begin{columns}
+\begin{column}{0.4\textwidth}
+\begin{figure}
+\includegraphics[totalheight=0.8\textheight]{spn_diff_char.jpg}
+\end{figure}
+\end{column}
+
 \begin{column}{0.6\textwidth}
@@
 \vspace{5mm}
-Input Diff: $\Delta P = 0B00$
-
-\vspace{2mm}
-Characteristic
+\pause Input Difference: $\Delta P = 0B00$
+
+\vspace{2mm}
+\pause Characteristic
 \[ \Delta P = 0B00 \rightarrow 0606 = \Delta U_4 \]
@@
 \vspace{5mm}
-Partial Subkey guess only for
+\pause Partial Subkey guess only for
 \[ [K_{5,5} ... K_{5,8}, K_{5,13} ... K_{5,16}] \]
-\end{column}
-
-\begin{column}{0.4\textwidth}
-\begin{figure}
-\includegraphics[totalheight=0.8\textheight]{spn_diff_char.jpg}
-\end{figure}
 \end{column}
@@
 \[ NS_{a}(\alpha, \beta) = | \{ X | 0 \leq X < 64, ( \bigoplus_{i=0}^5 X[i]\cdot\alpha[i]) = (\bigoplus_{j=0}^3 S_a(X)[j] \cdot \beta[j])\} | \]
-This looks more complicated, but is infact the same expression as for SPN, except it takes into account all possible S-Boxes.
+This looks more complicated, but is in fact the same expression as for SPN, except it takes into account all possible S-Boxes.
 \end{frame}
@@
 \vspace{5mm}
-With 6-bit S-Boxes things are not as easy as with 4-bit S-Boxes.
-\begin{figure}
-\includegraphics[totalheight=0.5\textheight]{NS_S5.PNG}
-\caption{Bias for S-Box 5 }
+\begin{figure}
+\includegraphics[totalheight=0.7\textheight]{NS_S5.PNG}
+\caption{$NS_5(\alpha, \beta) - 32$}
 \end{figure}
@@
 \begin{frame}
-We perform the attack on the 4-round DES similiarly to how we performed the attack on 4-round SPN, by "undoing" the last round.\\
+We perform the attack on the 4-round DES similarly to how we performed the attack on 4-round SPN, by ``undoing'' the last round.\\
 \vspace{5mm}
@@
 Using the calculation from the previous slide, we see that for a 4-round DES, our equation holds with bias of about 1 in 4.
-Thus we will need about $16\times 4^2 = 256$ pairs for the attack to have near perfect chance of sucess.\\
+Thus we will need about $16\times 4^2 = 256$ pairs for the attack to have near perfect chance of success.\\
 \vspace{3mm}
@@
 \item{Combine the pairs through all the rounds to get a highly probable differential characteristic until the penultimate round}
-\item{Exploit this differential characteristic by chosing a number of pairs of plaintexts with the desired difference and undo the last round by guessing the keybits}
+\item{Exploit this differential characteristic by choosing a number of pairs of plaintexts with the desired difference and undo the last round by guessing the keybits}
 \item{If the initial choice of the input difference was smart, we will only need to guess a few keybits}

### 2009-02-01 trunk/src/DifferentialCryptanalysis.java
DifferentialCryptanalysis on 4-round DES works ... YAHOO !!!

@@ -3,11 +3,16 @@ import java.util.BitSet;
 import java.util.List;
 import java.util.Random;
 
-
+/**
+ * Tools for Differential Cryptanalysis of DES
+ * 
+ * @author Sourav Sen Gupta
+ */
 public class DifferentialCryptanalysis {
   
   public static void main(String[] args) {
 	  DesDifferentialAttack4Round();
+	  //DesDifferentialAttack3Round();
 	  //PrintNSforAll();
   }	
 	
@@ -118,26 +123,38 @@ public class DifferentialCryptanalysis {
       BitSet p_one = Util.concatenate(p12, 32, p34, 32);
       BitSet c_one = des.DesEncBlock(p_one, key, numRounds);
       
-      //delta_p = 00 80 82 00 60 00 00 00
-      BitSet delta_p1 = Util.toBitSet(0x20000000,32);
-      BitSet delta_p2 = Util.toBitSet(0x00000000,32);
-      BitSet delta_p = Util.concatenate(delta_p1, 32, delta_p2, 32);
-            
+      //delta_p = 20 00 00 00 00 00 00 00
+      BitSet delta_pL1 = Util.toBitSet(0x2000, 16);
+      BitSet delta_pL2 = Util.toBitSet(0x0000, 16);
+      BitSet delta_pR1 = Util.toBitSet(0x0000, 16);
+      BitSet delta_pR2 = Util.toBitSet(0x0000, 16);
+      BitSet delta_pL = Util.concatenate(delta_pL1, 16, delta_pL2, 16);
+      BitSet delta_pR = Util.concatenate(delta_pR1, 16, delta_pR2, 16);
+      
+      BitSet delta_p = Util.concatenate(delta_pL, 32, delta_pR, 32);
+      
       //second pair
       BitSet p_two = Util.copyBitSet(p_one, 64);
       p_two.xor(delta_p);
       BitSet c_two = des.DesEncBlock(p_two, key, numRounds);
-            
+      
+      BitSet delta_CR = c_one.get(32, 64);
+      delta_CR.xor(c_two.get(32,64));
+           
       sets.add(new BitSet[]{c_one, c_two});
+      
     }
     return sets;
   }
   
  
   
-  public static void DesDifferentialAttack4Round() {
-	    // do analysis for 4-round DES
-	    int numRounds = 4;
+  /**
+ * 
+ */
+public static void DesDifferentialAttack3Round() {
[…]

### 2009-02-01 trunk/talk/draft1.tex
Differential DES slides done ... Fixed some typos ... Overlays done upto Chapter 3 ...

@@
 \vspace{5mm}
-\pause The S-Box provides the {\it confusion} function and the permutation applies the {\it confusion} operation in SPN, thus making it cryptographically similar to DES.
+\pause The S-Box provides the {\it confusion} function and the permutation applies the {\it diffusion} operation in SPN, thus making it cryptographically similar to DES.
 \vspace{5mm}
@@
 
+\begin{frame}
+\begin{beamercolorbox}[ht=2.5ex,dp=1.125ex,center,rounded=true,shadow=true]{author in head/foot}
+Linear Cryptanalysis of 4 Round DES
+\end{beamercolorbox}
+\end{frame}
+
 \subsection{Linear Attack: 4-round DES}
@@
 The plan for the DES Attacks:
-\begin{itemize}
+\begin{itemize}[<+->]
 \item{Structurally and cryptographically, DES is similar to SPN.}
@@
 \begin{frame}
-The first step, as before, is to come up with a linear approximation for the S-Boxes used by DES.\\
-\vspace{5mm}
-Note that the S-Boxes here are 6-bit to 4-bit maps.
-We can use the same technique as before.
-For a given S-Box $a \in \{1, 2, \dots 8\}$, $1 \leq \alpha \leq 63$, $1 \leq \beta \leq 15$, we define $NS_{a}(\alpha, \beta)$ as the number of times (out of 64) for S-Box $a$ that for all the input patterns masked by $\alpha$ the output pattern masked by $\beta$ agrees with the value of S-Box $a$.
+The first step, as before, is to come up with a linear approximation for the S-Boxes used by DES.
+
+\vspace{5mm}
+\pause Note that the S-Boxes here are 6-bit to 4-bit maps. We can use the same technique as before.
+
+\vspace{5mm}
+\pause For a given S-Box $a \in \{1, 2, \dots 8\}$, $1 \leq \alpha \leq 63$, $1 \leq \beta \leq 15$, we define $NS_{a}(\alpha, \beta)$ as the number of times (out of 64) for S-Box $a$ that for all the input patterns masked by $\alpha$ the output pattern masked by $\beta$ agrees with the value of S-Box $a$.
 \[ NS_{a}(\alpha, \beta) = | \{ X | 0 \leq X < 64, ( \bigoplus_{i=0}^5 X[i]\cdot\alpha[i]) = (\bigoplus_{j=0}^3 S_a(X)[j] \cdot \beta[j])\} | \]
-This looks more complicated, but is in fact the same expression as for SPN, except it takes into account all possible S-Boxes.
-\end{frame}
-
-\begin{frame}
-Using $NS$ to find a linear approximation for the S-Boxes.\\
-\vspace{5mm}
-With 6-bit S-Boxes things are not as easy as with 4-bit S-Boxes.
-\begin{figure}
-\includegraphics[totalheight=0.5\textheight]{NS_S5.PNG}
-\caption{Bias for S-Box 5 }
+
+\vspace{3mm}
+\pause This looks more complicated, but is in fact the same expression as for SPN, except it takes into account all possible S-Boxes.
+\end{frame}
+
+\begin{frame}
+Using $NS$ to find a linear approximation for the S-Boxes.
+\vspace{2mm}
+\begin{figure}
+\includegraphics[totalheight=0.7\textheight]{NS_S5.PNG}
+\caption{$NS_5(\alpha, \beta) - 32$}
 \end{figure}
@@
 \begin{frame}
-From previous result, the S-Box approximation with highest bias for S-Box 5. Let $X$ be the input bits to the S-Box and let $Y$ be the output bits. Then:
+So, we noticed the S-Box approximation with highest bias for S-Box 5.
+
+\vspace{5mm}
+\pause Letting $X$ be the input bits to S-Box 5 and $Y$ be the output bits,
 \[ X_{2} = Y_{1} \oplus Y_{2} \oplus Y_{3} \oplus Y_{4} \]
-Holds with bias of absolute value $\sim 0.31$\\
-\vspace{5mm}
-Tracing through the Feistel function to the right input bits $R$, round key bits $K$, and Feistel function output bits $F$ we get
+holds with bias of absolute value $\sim 0.31$
+
+\vspace{5mm}
+\pause Tracing through the Feistel function to the right input bits $R$, round key bits $K$, and Feistel function output bits $F$ we get
 \[ R_{17} \oplus F_{25} \oplus F_{14} \oplus F_{8} \oplus F_{3} = K_{26} \]
@@
 Let $C_L$ and $C_R$ be the left and right ciphertext bits.
-Let $K_i$ be the round key for round $i$ and $U_{4L}$ and $U_{4R}$ the left and right halves of the input to round 4.\\
-\vspace{5mm}
-Using the S-Box 5 approximation and the Piling Up Principle we can combine the approximations to get
+Let $K_i$ be the round key for round $i$ and $U_{4L}$ and $U_{4R}$ the left and right halves of the input to round 4.
+
+\vspace{5mm}
+\pause Using the S-Box 5 approximation and the Piling Up Principle we can combine the approximations to get
 \begin{align*}
@@
 \end{align*}
-The bias magnitude for this expression is
+
+\pause The bias magnitude for this expression is
 \[ (12/64)^2 + (1 - 12/64)^2 - 1/2 \approx 0.30 \]
-This is the best expression for the 3-round DES.
+
+\pause This is the best expression for the 3-round DES.
 \end{frame}
@@
 \pause
-Note that in the last round, $U_{4L}$ bits are not changed, only $U_{4R}$ bits are affected. Note also that there is only one $U_R$ bit the 3-round equation.
+Note that in the last round, $U_{4L}$ bits are not changed, only $U_{4R}$ bits are affected. Note also that there is only one $U_{4R}$ bit the 3-round equation.
 \begin{align*} 
@@
 \pause
-This bit, $U_{4R}[17]$ is affected by the result of S-Box 1 in round 4.
+This bit, $U_{4R}[17]$, is affected by the result of S-Box 1 in round 4.
 If we guess correctly for the 6 partial subkey bits $K_{4,1}$ to $K_{4,6}$, the 3-round equation will hold with high bias, otherwise it will likely hold with close to 0 bias.\\
@@
 Thus, this attack will give us $6$ of the $56$ key bits.
-What this means in terms of time saved over exhaustive search is this: If it takes a year to find a key through exhaustive search of $2^{56}$ bits, after this attack it will only take $5$ days to exhaustively search the remaining $2^{50}$ bits.
+\pause What this means in terms of time saved over exhaustive search is this: If it takes a year to find a key through exhaustive search of $2^{56}$ bits, after this attack it will only take $5$ days to exhaustively search the remaining $2^{50}$ bits.
 \end{frame}
@@
 \vspace{3mm}
-Matsui Lemma 5: Let $N$ be the number of given random plaintexts and $\epsilon$, the bias, be sufficiently small.
+\pause Matsui Lemma 5: Let $N$ be the number of given random plaintexts and $\epsilon$, the bias, be sufficiently small.
 Let $q^{(i)}$ be the probability that the following equation holds for a subkey candidate $K_n^{(i)}$ and a random variable X:
 \[ F_n(X,K_n)[l_1, l_2,\dots,l_d] = F_n(X, K_n^{(i)})[l_1, l_2, \dots, l_d].\]
-Then if $q^{(i)}$'s are independent, the success rate of the attack is
+\pause Then if $q^{(i)}$'s are independent, the success rate of the attack is
 \[ \int_{x=-2\sqrt{N}\epsilon}^\infty \left( \prod_{K_n^{(i)} \neq K_n} \int_{-x-4\sqrt{N}\epsilon q^{(i)}}^{x+4\sqrt{N}\epsilon (1 - q^{(i)})} \frac{1}{\sqrt{2\pi}} e^{-y^2/2} dy \right) \frac{1}{\sqrt{2\pi}} e^{-x^2/2} dx \]
-Where the product is taken over all subkey candidates except $K_n$.\\
+where the product is taken over all subkey candidates except $K_n$.
+
 \vspace{3mm}
-Let $d = 1$ and $l_1 = 17$ in the above expression - this corresponds to our attack (distribution over guessing for one bit, $CR_{17}$ in the equation).
+\pause Let $d = 1$ and $l_1 = 17$ in the above expression - this corresponds to our attack (distribution over guessing for one bit, $U_{4R}[17]$ in the equation).
 The following table summarizes the success rates:
@@
 Using the calculation from the previous slide, we see that for a 4-round DES, our equation holds with bias of about 1 in 4.
-Thus we will need about $16\times 4^2 = 256$ pairs for the attack to have near perfect chance of success.\\
+\pause Thus we will need about $16\times 4^2 = 256$ pairs for the attack to have near perfect chance of success.
+
 \vspace{3mm}
-For a full 16-round DES, the best approximation has bias $\sim 1.19\times 10^{-22}$, thus we would need on the order of $2^{47}$ plaintext-ciphertext pairs.
+\pause For a full 16-round DES, the best approximation has bias $\sim 1.19\times 10^{-22}$, thus we would need on the order of $2^{47}$ plaintext-ciphertext pairs.
 \end{frame}
@@
 \begin{frame}
+\begin{beamercolorbox}[ht=2.5ex,dp=1.125ex,center,rounded=true,shadow=true]{author in head/foot}
+Differential Cryptanalysis of 4 Round DES
+\end{beamercolorbox}
+\end{frame}
+
+\begin{frame}
 Differential Attack Idea
@@
 \vspace{5mm}
-Similar to the differential attack on SPN, here we care about the differential characteristics of the cipher. So, the basic idea is as follows
-
-\begin{itemize}
+Similar to the differential attack on SPN, here we care about the differential characteristics of the cipher. \pause So, the basic idea is as follows
+
+\begin{itemize}[<+->]
 \item{Construct the difference tables for the S-Boxes}
@@
 \begin{frame}
-The chosen input difference:$\Delta P = 20 00 00 00 00 00 00 00$
-
-Round 1 Differential
-
-\[ \Delta P = 20 00 00 00 00 00 00 00 \rightarrow 20 00 00 00 00 00 00 00 = \Delta V_1 \]
-Probability: $p_1 = 1$ as Feistel gives output difference $0$ for input $0$
-
-Round 2 Differential
-
-\[ \Delta U_2 = \Delta V_1 = 20 00 00 00 00 00 00 00 \rightarrow \]
-
-this calculations and coding has not been completed yet ... will be done ...
-\end{frame}
-
-\begin{frame}
-Differentials for SBoxes
+\begin{columns}
+\begin{column}{0.4\textwidth}
+\begin{figure}
+\includegraphics[totalheight=0.8\textheight]{des_diff_char.jpg}
+\end{figure}
+\end{column}
+
+\begin{column}{0.6\textwidth}
+Differential Characteristic of DES
+
+\vspace{5mm}
+Input Difference: $\Delta P = 20\:00\:00\:00\:00\:00\:00\:00$
+
+\vspace{5mm}
+\pause Propagating Difference Pairs
+\begin{itemize}[<+->]
+\item{Round 1: $S_i$ : $00 \rightarrow 00$}
+\item{Round 2: $S_1$ : $04 \rightarrow 05$}
+\item{Round 3: $S_4$ : $01 \rightarrow 05$ \\ $\qquad \qquad \: S_5$ : $10 \rightarrow 07$ \\ $\qquad \qquad \: S_8$ : $04 \rightarrow 07$}
+\end{itemize}
+
+\vspace{5mm}
+\pause Differential Characteristic
+\[ 20\:00\:00\:00\:00\:00\:00\:00 \rightarrow A0\:06\:18\:A0\:00\:00\:80\:02 \]
+\end{column}
+\end{columns}
+\end{frame}
+
+\begin{frame}
+Probabilistic Analysis
+
+\vspace{5mm}
+Constructing the Difference Distribution tables for the S-Boxes and analyzing the probability distribution for differential propagation through SBoxes:
+
+\vspace{2mm}
+\pause
+\begin{itemize}[<+->]
+\item{Round 1: $S_i$ : $00 \rightarrow 00$ with $Pr = 1$}
+\item{Round 2: $S_1$ : $04 \rightarrow 05$ with $Pr = 10/64$}
+\item{Round 3: $S_4$ : $01 \rightarrow 05$  with $Pr = 16/64$ \\ $\qquad \qquad \: S_5$ : $10 \rightarrow 07$ with $Pr = 12/64$ \\ $\qquad \qquad \: S_8$ : $04 \rightarrow 07$ with $Pr = 12/64$}
+\end{itemize}
+
+\vspace{3mm}
+\pause Differential Characteristic: $20\:00\:00\:00\:00\:00\:00\:00 \rightarrow A0\:06\:18\:A0\:00\:00\:80\:02$ holds with probability
+\[ p_D \: = \: (10/64) \times (16/64) \times (12/64) \times (12/64) \: = \: 45/32768 \approx 0.001373 \]
+
+\vspace{2mm}
+\pause Note: $p_D$ is much higher than the random distribution probability $1/2^{64}$ and we hope to exploit this situation to break DES.
+
+\end{frame}
+
+\begin{frame}
+\begin{columns}
+\begin{column}{0.4\textwidth}
+\begin{figure}
+\includegraphics[totalheight=0.8\textheight]{des_diff_char.jpg}
+\end{figure}
+\end{column}
+
+\begin{column}{0.6\textwidth}
+Differential Characteristic of DES
+\[ 20\:00\:00\:00\:00\:00\:00\:00 \rightarrow A0\:06\:18\:A0\:00\:00\:80\:02 \]
+
+\pause Note: The right half of the round 3 characteristic output is $00\:00\:80\:02$.
+
+\vspace{2mm}
+\pause Thus, the right halves of the ciphertexts at this round differ by 2 bits, and these are related to S-Box $S_1$ in round 4.
+
+\vspace{2mm}
+\pause Hence, we just need to guess the {\it target partial subkey} $K_{4,1}, ..., K_{4,6}$ for $S_1$, i.e. the first 6 bits of $K_4$.
+
+\vspace{2mm}
+\pause Thereafter, we can ``undo'' round 4 using this guess and  the maximum probability of satisfying the 3 round differential characteristic will indicate a correct guess.
+\end{column}
+\end{columns}
+\end{frame}
+
+\begin{frame}
+How many pairs do we need? (Probabilistic justification)
 
@@
 
-\subsection{Extension to full DES}
-\begin{frame}
-Extension of the same ideas to 6 / 8 / 16 round DES ... Probabilistic explanation ...
-
-\end{frame}
-
 \section{Conclusion}
@@
 \begin{beamercolorbox}[ht=2.5ex,dp=1.125ex,center,rounded=true,shadow=true]{author in head/foot}
-Did we miss anything important?
+Concluding Remarks
 \end{beamercolorbox}

### 2009-02-01 trunk/talk/draft1.tex
Some fixes to linear slides, new figures, and a bit more conclusion.

@@
 \begin{figure}
-\includegraphics[width=0.7\textwidth]{spn_sbox_unkeyed.jpg}
+\includegraphics[width=0.7\textwidth]{spn_sbox_X23_Y134.PNG}
 \end{figure}
@@
 \begin{figure}
-\includegraphics[height=0.45\textheight]{spn_single_round.jpg}
+\includegraphics[height=0.45\textheight]{spn_round_expand.PNG}
 \end{figure}
@@
 \begin{frame}
-Actual attack on SPN ... Demo ...
+DEMO: Attack on 4-round SPN
 \end{frame}
@@
 \begin{frame}
-The actual Attack on 4 round DES ... Demo ...
+DEMO: Attack on 4-round DES
 \end{frame}
@@
 \begin{frame}
-Links between Linear and Differential Attacks ...
-
+Linear Resistance:\\
+\vspace{5mm}
+Thm: A function $F: GF(2)^p \rightarrow GF(2)^q$is perfect nonlinear (ie bias is 0) if and only if it is \textit{bent}. \\
+\vspace{3mm}
+Thm: Bent functions exist only for $p \geq 2q$ and $p$ even. 
+\vspace{5mm}
 \end{frame}
@@
 \begin{frame}
-If you are building a cryptosystem ... To Do / Not To Do ...
-
-\end{frame}
-
-\begin{frame}
-Questions ... Doubts ... Coding ... Misc ... Discussion ...
-
+The problems of DES originated in two aspects:
+\begin{itemize}
+\item{6 bit S-Boxes are too easy to identify with linear or affine functions with high bias}
+\item{Diffusion between rounds is just not fast enough}
+\end{itemize}
+AES, which replaced DES, provides an 8-Bit S-Box in the SubBytes step and higher diffusion through the ShiftRows and MixColumns steps. 
+\end{frame}
+
+\begin{frame}
+References:\\
+\vspace{5mm}
+Matsui\\
+Beham, Shamir\\
+Heys
+Chabaud, Vaudenay\\
+Manferdelli
+\end{frame}
+
+\begin{frame}
+\begin{center}
+The End.\\
+Questions?
+\end{center}
 \end{frame}

### 2009-02-01 src/DifferentialCryptanalysis.java
Formatting fixes

@@ -9,78 +9,80 @@ import java.util.Random;
  * @author Sourav Sen Gupta
  */
 public class DifferentialCryptanalysis {
-  
-  public static void main(String[] args) {
-	  DesDifferentialAttack4Round();
-	  //DesDifferentialAttack3Round();
-	  //PrintNSforAll();
-  }	
-	
 
-  // Prints the Differential Tables for all Sboxes
+  public static void main(String[] args) {
+    DesDifferentialAttack4Round();
+    // DesDifferentialAttack3Round();
+    // PrintNSforAll();
+  }
+
   /**
-  * @param args
-  */
-  public static void PrintNSforAll(){
-	
-	for (int sbox = 0 ; sbox < 8; sbox++) {
+   * Prints the Differential Tables for all Sboxes
+   */
+  public static void PrintNSforAll() {
+
+    for (int sbox = 0; sbox < 8; sbox++) {
       System.out.println("SBOX: " + (sbox + 1));
-      
-      
+
       BitSet alpha = new BitSet(6);
       alpha.set(0, 5, false);
       BitSet beta = new BitSet(4);
       beta.set(0, 3, false);
-      
+
       for (int i = 0; i < 16; i++) {
         System.out.print("\t" + i);
       }
       System.out.println();
-      
+
       for (int i = 0; i < 64; i++) {
         System.out.print(i + ":\t");
         for (int j = 0; j < 16; j++) {
           int ns = BihamShamirNS(sbox, alpha, beta);
-          
-          Util.increment(beta, 4);          
-          System.out.print(ns + "\t");          
-        }        
-        Util.increment(alpha, 6);        
-        System.out.println();        
+
+          Util.increment(beta, 4);
+          System.out.print(ns + "\t");
+        }
+        Util.increment(alpha, 6);
+        System.out.println();
       }
       System.out.println();
       System.out.println();
-      
-    }    
-        
+
+    }
+
   }
-  
-  
+
   /**
-   * Defines the NS function in lines of Biham Shamir "Differential Cyptanalysis of DES like Cryptosystems"
-   * @param sbox identifies the S-Box (1 to 8)
-   * @param alpha represents a value between 0 and 63 selecting bits of the input difference
[…]

### 2009-02-01 src/LinearCryptanalysis.java

@@ -1,3 +1,4 @@
+import java.io.IOException;
 import java.util.ArrayList;
 import java.util.BitSet;
 import java.util.List;
@@ -16,9 +17,9 @@ public class LinearCryptanalysis {
    * @param args
    */
   public static void main(String[] args) {
-    DesLinearAttack4Round();
+    //DesLinearAttack4Round();
     // PrintNSForAll();
-    // MatsuiEqn5Distribution();
+    MatsuiEqn5Distribution();
     // testFinalRound();
   }
 
@@ -182,8 +183,8 @@ public class LinearCryptanalysis {
     BitSet Fguess = des.PermuteCInv(des.Feistel(CR, Kguess));
 
     System.out.println("F, and F guess");
-    printBitSet(F, 32, 32);
-    printBitSet(Fguess, 32, 32);
+    Util.printBitSet(F, 32);
+    Util.printBitSet(Fguess, 32);
 
     // now a slightly more involved test - we encrypt
     // a block with 4 rounds, then invert the last round
@@ -225,21 +226,21 @@ public class LinearCryptanalysis {
     BitSet C4R = C4.get(32, 64);
 
     System.out.println("Ls");
-    printBitSet(C3L, 32, 32);
-    printBitSet(C4L, 32, 32);
+    Util.printBitSet(C3L, 32);
+    Util.printBitSet(C4L, 32);
     System.out.println("Rs");
-    printBitSet(C3R, 32, 32);
-    printBitSet(C4R, 32, 32);
+    Util.printBitSet(C3R, 32);
+    Util.printBitSet(C4R, 32);
 
     // undo the last round
     C4L.xor(des.Feistel(C4R, K4));
     // undone L of 4th round will be R of 3-round DES
     System.out.println("Ls");
-    printBitSet(C3L, 32, 32);
-    printBitSet(C4R, 32, 32);
+    Util.printBitSet(C3L, 32);
+    Util.printBitSet(C4R, 32);
     System.out.println("Rs");
-    printBitSet(C3R, 32, 32);
-    printBitSet(C4L, 32, 32);
+    Util.printBitSet(C3R, 32);
+    Util.printBitSet(C4L, 32);
   }
 
   /**
@@ -260,14 +261,12 @@ public class LinearCryptanalysis {
     DesImpl des = new DesImpl();
     List<BitSet[]> sets = new ArrayList<BitSet[]>();
     for (int i = 0; i < pairs; i++) {
[…]

### 2009-02-01 trunk/src/Util.java
Formatting fixes

@@ -135,4 +135,42 @@ public final class Util {
     }
     return a;
   }
+
+  /**
+   * Concatenate an array of bit sets of equal length
+   * 
+   * @param sets
+   *          Array of sets to concatenate
+   * @param totalBitsEach
+   *          number of bits in each bit set to concatenate
+   * @return A single bit set with concatenated bits
+   */
+  public static BitSet concatenate(BitSet[] sets, int totalBitsEach) {
+    BitSet result = new BitSet(sets.length * totalBitsEach);
+    for (int i = 0; i < sets.length; i++) {
+      for (int j = 0; j < totalBitsEach; j++) {
+        result.set((i * totalBitsEach) + j, sets[i].get(j));
+      }
+    }
+    return result;
+  }
+  
+  /**
+   * Print a bit set to the screen
+   * @param set
+   * @param totalBits
+   */
+  public static void printBitSet(BitSet set, int totalBits) {
+    int ptr = 0;
+    while (ptr < totalBits) {
+      int val = Util.toInteger(set.get(ptr, ptr + 4), 4);
+      System.out.print(Integer.toHexString(val).toUpperCase());
+      //System.out.print(set.get(ptr) ? "1" : "0");
+//      if (++ptr % lineLength == 0) {
+//        System.out.println();
+//      }
+      ptr += 4;
+    }
+    System.out.println();
+  }
 }

### 2009-02-02 trunk/talk/draft1.tex
Slide fixes + more conclusion

@@
 \item{$V_{j,i}$ where $j = 1, 2, 3, 4$ and $i = 1, 2, \dots 16$ be the $i$-th output bit of the $j$-th round of SPN}
+\item{$K_{j,i}$ where $j = 1, 2, 3, 4$ and $i = 1, 2, \dots 16$ be the $i$-th bit of the round key for the $j$-th round of SPN}
 \end{itemize}
@@
 \[ Pr(X_1 = i)  = \left\{ \begin{array}{ll}
-         p_1 & \mbox{$i = 0$};\\
+         p_1 & \mbox{$i = 0$}\\
          (1 - p_1) & \mbox{$i = 1$}.\end{array} \right. \] 
@@
 \[ U_{4,6} \oplus U_{4,8} \oplus U_{4,14} \oplus U_{4, 16} \oplus P_5 \oplus P_7 \oplus P_8 = \Sigma K \]
-\pause Where $\Sigma K$ is the sum over some key bits. Note that Since the key is fixed, $\Sigma K = 0$ or $1$ and thus we can ignore it since we only care about the bias.\\
+\pause Where $\Sigma K$ is the sum over some key bits. Note that since the key is fixed, $\Sigma K = 0$ or $1$ and thus we can ignore it since we only care about the bias.\\
 \vspace{5mm}
@@
  P_L[25] & \oplus P_L[14] \oplus P_L[8] \oplus P_L[3] \oplus \\
-         & U_{4L}[25] \oplus U_{4L}[14] \oplus U_{4L}[8] \oplus U_{4L}[3] \oplus \\
-			& P_R[17] \oplus U_{4R}[17] = K_1[26] \oplus K_3[26] \\
+         & U_{4R}[25] \oplus U_{4R}[14] \oplus U_{4R}[8] \oplus U_{4R}[3] \oplus \\
+			& P_R[17] \oplus U_{4L}[17] = K_1[26] \oplus K_3[26] \\
 \end{align*}
@@
 \pause
-Note that in the last round, $U_{4L}$ bits are not changed, only $U_{4R}$ bits are affected. Note also that there is only one $U_{4R}$ bit the 3-round equation.
+Note that in the last round, $U_{4R}$ bits are not changed, only $U_{4L}$ bits are affected. Note also that there is only one $U_{4L}$ bit the 3-round equation.
 \begin{align*} 
-P_L[25] \oplus P_L[14] \oplus P_L[8] \oplus & P_L[3] \oplus U_{4L}[25] \oplus U_{4L}[14] \oplus U_{4L}[8] \\ & \oplus U_{4L}[3] \oplus P_R[17] \oplus U_{4R}[17] = K_1[26] \oplus K_3[26]
+P_L[25] \oplus P_L[14] \oplus P_L[8] \oplus & P_L[3] \oplus U_{4R}[25] \oplus U_{4R}[14] \oplus U_{4R}[8] \\ & \oplus U_{4R}[3] \oplus P_R[17] \oplus U_{4L}[17] = K_1[26] \oplus K_3[26]
 \end{align*}
@@
 \pause
-This bit, $U_{4R}[17]$, is affected by the result of S-Box 1 in round 4.
+This bit, $U_{4L}[17]$, is affected by the result of S-Box 1 in round 4.
 If we guess correctly for the 6 partial subkey bits $K_{4,1}$ to $K_{4,6}$, the 3-round equation will hold with high bias, otherwise it will likely hold with close to 0 bias.\\
@@
 
-\begin{frame}
-What if that is false ? ... Discussion ...
-
-\end{frame}
-
 \subsection{Links between the Attacks}
@@
 Linear Resistance:\\
-\vspace{5mm}
-Thm: A function $F: GF(2)^p \rightarrow GF(2)^q$is perfect nonlinear (ie bias is 0) if and only if it is \textit{bent}. \\
-\vspace{3mm}
-Thm: Bent functions exist only for $p \geq 2q$ and $p$ even. 
-\vspace{5mm}
-\end{frame}
-
-\begin{frame}
-Mathematically and while implementing ...
-
+\vspace{3mm}
+Let $F: GF(2)^p \rightarrow GF(2)^q$.\\
+\vspace{3mm}
+Defn: A function F is bent if its Walsh transform $\mathcal{W}\{F\}$ is $\pm 2^{p/2}$.\\
+\vspace{3mm}
+Thm: A function F is perfect nonlinear (ie bias is 0) if and only if it is \textit{bent}. \\
+\vspace{3mm}
+Thm: Bent functions exist only for $p \geq 2q$ and $p$ even. \\
+\vspace{5mm}
+To sumarize: When $p\geq 2q$ and $p$ even, differential-resistant is equivalent to linear resistant, and to vectorial Bentness.
 \end{frame}
@@
 \vspace{5mm}
-Matsui\\
-Beham, Shamir\\
-Heys
-Chabaud, Vaudenay\\
-Manferdelli
+M. Matsui, "Linear Cryptanalysis Method for DES Cipher", \textit{Advances in Cryptology - EUROCRYPT '93 (Lecture Notes in Computer Science no. 765),} Springer-Verlag, pp. 38\\
+\vspace{2mm}
+E Biham And A. Shamir, "Differential Cryptanalysis of DES-like Cryptosystems", \textit{Journal of Cryptology}, vol 4, no 1, pp. 3-72, 1991.\\
+\vspace{2mm}
+H. Heys, "A Tutorial on Linear and Differential Cryptanalysis"\\
+\vspace{2mm}
+F.
+Chabaud And S.
+Vaudenay, "Links Between Differential and Linear Cryptanalysis", \textit{Advances in Cryptology - EUROCRYPT '94 (Lexture Notes in Computer Science no.
+950)}, Springer-Verlag, pp.
+356-365, 1995.\\
+\vspace{2mm}
+J. Manferdelli, UW Cryptography Lecture Notes
 \end{frame}

### 2009-02-02 trunk/talk/draft1.tex
A couple of small changes to DES Linear slides

@@
 \begin{frame}
+\begin{columns}
+
+\begin{column}{0.35\textwidth}
+\begin{figure}
+\includegraphics[totalheight=0.8\textheight]{DES_4.PNG}
+\end{figure}
+\end{column}
+
+\begin{column}{0.65\textwidth}
+\small
 We perform the attack on the 4-round DES similarly to how we performed the attack on 4-round SPN, by ``undoing'' the last round.\\
-\vspace{5mm}
+\vspace{3mm}
 \pause
-Note that in the last round, $U_{4R}$ bits are not changed, only $U_{4L}$ bits are affected. Note also that there is only one $U_{4L}$ bit the 3-round equation.
+Note that in the last round, $U_{4R}$ bits are not changed, only $U_{4L}$ bits are affected. Note also that there is only one $U_{4L}$ bit the 3-round equation.\\
 \begin{align*} 
-P_L[25] \oplus P_L[14] \oplus P_L[8] \oplus & P_L[3] \oplus U_{4R}[25] \oplus U_{4R}[14] \oplus U_{4R}[8] \\ & \oplus U_{4R}[3] \oplus P_R[17] \oplus U_{4L}[17] = K_1[26] \oplus K_3[26]
-\end{align*}
-\vspace{5mm}
+& P_L[25]    \oplus P_L[14]    \oplus P_L[8]    \oplus P_L[3] \oplus  \\
+& U_{4R}[25] \oplus U_{4R}[14] \oplus U_{4R}[8] \oplus U_{4R}[3]  \oplus \\ 
+& P_R[17]    \oplus U_{4L}[17] = K_1[26] \oplus K_3[26] \\
+\end{align*}\\
+\vspace{3mm}
 \pause
@@
 If we guess correctly for the 6 partial subkey bits $K_{4,1}$ to $K_{4,6}$, the 3-round equation will hold with high bias, otherwise it will likely hold with close to 0 bias.\\
-\vspace{5mm}
+\vspace{3mm}
 \pause
-Thus, this attack will give us $6$ of the $56$ key bits.
-\pause What this means in terms of time saved over exhaustive search is this: If it takes a year to find a key through exhaustive search of $2^{56}$ bits, after this attack it will only take $5$ days to exhaustively search the remaining $2^{50}$ bits.
+Thus, this attack will give us $6$ of the $56$ key bits. If it takes a year to find a key through exhaustive search of $2^{56}$ bits, it only takes $5$ days to exhaustively search $2^{50}$ bits.
+\end{column}
+
+\end{columns}
 \end{frame}

### 2009-02-02 src/DifferentialCryptanalysis.java
Info printing at the beginning of each attack.

@@ -1,3 +1,4 @@
+import java.io.IOException;
 import java.util.ArrayList;
 import java.util.BitSet;
 import java.util.List;
@@ -274,24 +275,36 @@ public class DifferentialCryptanalysis {
     int numRounds = 4;
 
     // set the key
-    BitSet key1 = Util.toBitSet(0x1234, 16);
-    BitSet key2 = Util.toBitSet(0x8743, 16);
-    BitSet key3 = Util.toBitSet(0xFAC3, 16);
-    BitSet key4 = Util.toBitSet(0xECAB, 16);
-    BitSet key12 = Util.concatenate(key1, 16, key2, 16);
-    BitSet key34 = Util.concatenate(key3, 16, key4, 16);
-    BitSet key = Util.concatenate(key12, 32, key34, 32);
+    BitSet key = Util.concatenate(new BitSet[] { Util.toBitSet(0x1234, 16),
+        Util.toBitSet(0x8743, 16), Util.toBitSet(0xFAC3, 16),
+        Util.toBitSet(0xECAB, 16) }, 16);
 
     // print the actual key bits
     DesImpl des = new DesImpl();
     BitSet K4 = des.KeySchedule(key, 4);
     BitSet K4_S1 = K4.get(0, 6);
 
-    System.out.println("actual subkey is "
-        + Integer.toHexString(Util.toInteger(K4_S1, 6)));
-
     // generate the pairs
-    int numPairs = 1000;
+    int numPairs = 10000;
+    
+    // print info 
+    System.out.println("4-round DES Differential Cryptanalysis.");
+    System.out.println("Using " + numPairs + " pairs.");
+    System.out.println("****************");
+    System.out.println("Using Key:");
+    Util.printBitSet(key, 64);
+    System.out.println("****************");
+    System.out.println("Real partial subkey:");
+    System.out.println("\tK4[1-6]:\t" +
+        Integer.toHexString(Util.toInteger(K4_S1, 6)).toUpperCase());
+    System.out.println("****************");
+    System.out.println("Press Enter to run differential attack...");
+    try {
+      System.in.read();
+    } catch(IOException exn) {
+      
+    }
+    
     System.out.print("Generating " + numPairs + " pairs...");
     List<BitSet[]> pairs = generateRandomPairs(numPairs, key, numRounds);
     System.out.println("done");

### 2009-02-02 src/LinearCryptanalysis.java

@@ -17,9 +17,9 @@ public class LinearCryptanalysis {
    * @param args
    */
   public static void main(String[] args) {
-    //DesLinearAttack4Round();
+    DesLinearAttack4Round();
     // PrintNSForAll();
-    MatsuiEqn5Distribution();
+    // MatsuiEqn5Distribution();
     // testFinalRound();
   }
 
@@ -289,16 +289,22 @@ public class LinearCryptanalysis {
     DesImpl des = new DesImpl();
     BitSet K4 = des.KeySchedule(key, 4);
 
-    // print info about the real key
+    
+    int numPairs = 1000;
+    
+    // print info 
+    System.out.println("4-round DES Linear Cryptanalysis.");
+    System.out.println("Using " + numPairs + " pairs.");
     System.out.println("****************");
     System.out.println("Using Key:");
     Util.printBitSet(key, 64);
     System.out.println("****************");
+    System.out.println("Real partial subkey:");
     BitSet K4_S1 = K4.get(0, 6);
-    System.out.println("Actual Partial Subkey:" +
+    System.out.println("\tK4[1-6]:\t" +
         Integer.toHexString(Util.toInteger(K4_S1, 6)).toUpperCase());
     System.out.println("****************");
-    System.out.println("Press Enter to start...");
+    System.out.println("Press Enter to run linear attack...");
     try {
       System.in.read();
     } catch(IOException exn) {
@@ -306,7 +312,6 @@ public class LinearCryptanalysis {
     }
 
     // generate the pairs
-    int numPairs = 1000;
     System.out.print("Generating " + numPairs + " pairs...");
     List<BitSet[]> pairs = generateRandomPairs(numPairs, key, numRounds);
     System.out.println("done");

### 2009-02-02 src/SpnDifferentialCryptanalysis.java

@@ -1,3 +1,4 @@
+import java.io.IOException;
 import java.util.ArrayList;
 import java.util.BitSet;
 import java.util.List;
@@ -83,6 +84,25 @@ public class SpnDifferentialCryptanalysis {
 
     // 1 - generate 10000 pairs
     int pairct = 10000;
+    
+    // print out the real key bits and some info
+    System.out.println("4-Round SPN Differential Cryptanalysis.");
+    System.out.println("Using " + pairct + " pairs.");
+    BitSet realPskA = roundKeys[4].get(4, 8);
+    BitSet realPskB = roundKeys[4].get(12, 16);
+    System.out.println("Real partial subkey values:");
+    System.out.println("\tK4[4-8]:\t" + 
+        Integer.toHexString(Util.toInteger(realPskA, 4)));
+    System.out.println("\tK4[12-16]:\t" + 
+        Integer.toHexString(Util.toInteger(realPskB, 4)));
+    System.out.println("**********************");
+    System.out.println("Press Enter to run differential attack...");
+    try {
+      System.in.read();
+    } catch (IOException exn) {
+    
+    }
+      
 
     SpnDifferentialCryptanalysis slc = new SpnDifferentialCryptanalysis();
     List<BitSet[]> pairs = slc.generateChosenPairs(pairct, roundKeys);

### 2009-02-02 src/SpnLinearCryptanalysis.java

@@ -1,3 +1,4 @@
+import java.io.IOException;
 import java.util.ArrayList;
 import java.util.BitSet;
 import java.util.List;
@@ -68,6 +69,24 @@ public class SpnLinearCryptanalysis {
 
     // 1 - generate 10000 pairs
     int pairct = 10000;
+    
+    // print out the real key bits and some info
+    System.out.println("4-Round SPN Linear Cryptanalysis.");
+    System.out.println("Using " + pairct + " pairs.");
+    BitSet realPskA = roundKeys[4].get(4, 8);
+    BitSet realPskB = roundKeys[4].get(12, 16);
+    System.out.println("Real partial subkey values:");
+    System.out.println("\tK4[4-8]:\t" + 
+        Integer.toHexString(Util.toInteger(realPskA, 4)));
+    System.out.println("\tK4[12-16]:\t" + 
+        Integer.toHexString(Util.toInteger(realPskB, 4)));
+    System.out.println("**********************");
+    System.out.println("Press Enter to run linear attack...");
+    try {
+      System.in.read();
+    } catch (IOException exn) {
+      
+    }
 
     SpnLinearCryptanalysis slc = new SpnLinearCryptanalysis();
     List<BitSet[]> pairs = slc.generateRandomPairs(pairct, roundKeys);

### 2009-02-02 talk/draft1.tex
DifferentialCryptanalysis.java fixed ... Diff Slides modified ...

@@
 \begin{frame}
-The first step, as before, is to come up with a linear approximation for the S-Boxes used by DES.\\
-\vspace{5mm}
-Note that the S-Boxes here are 6-bit to 4-bit maps.
-We can use the same technique as before.
-For a given S-Box $a \in \{1, 2, \dots 8\}$, $1 \leq \alpha \leq 63$, $1 \leq \beta \leq 15$, we define $NS_{a}(\alpha, \beta)$ as the number of times (out of 64) for S-Box $a$ that for all the input patterns masked by $\alpha$ the output pattern masked by $\beta$ agrees with the value of S-Box $a$.
+The first step, as before, is to come up with a linear approximation for the S-Boxes used by DES.
+
+\vspace{5mm}
+\pause Note that the S-Boxes here are 6-bit to 4-bit maps. We can use the same technique as before.
+
+\vspace{5mm}
+\pause For a given S-Box $a \in \{1, 2, \dots 8\}$, $1 \leq \alpha \leq 63$, $1 \leq \beta \leq 15$, we define $NS_{a}(\alpha, \beta)$ as the number of times (out of 64) for S-Box $a$ that for all the input patterns masked by $\alpha$ the output pattern masked by $\beta$ agrees with the value of S-Box $a$.
 \[ NS_{a}(\alpha, \beta) = | \{ X | 0 \leq X < 64, ( \bigoplus_{i=0}^5 X[i]\cdot\alpha[i]) = (\bigoplus_{j=0}^3 S_a(X)[j] \cdot \beta[j])\} | \]
-This looks more complicated, but is in fact the same expression as for SPN, except it takes into account all possible S-Boxes.
-\end{frame}
-
-\begin{frame}
-Using $NS$ to find a linear approximation for the S-Boxes.\\
-\vspace{5mm}
+
+\vspace{3mm}
+\pause This looks more complicated, but is in fact the same expression as for SPN, except it takes into account all possible S-Boxes.
+\end{frame}
+
+\begin{frame}
+Using $NS$ to find a linear approximation for the S-Boxes.
+\vspace{2mm}
 \begin{figure}
@@
 Let $C_L$ and $C_R$ be the left and right ciphertext bits.
-Let $K_i$ be the round key for round $i$ and $U_{4L}$ and $U_{4R}$ the left and right halves of the input to round 4.\\
-\vspace{5mm}
-Using the S-Box 5 approximation and the Piling Up Principle we can combine the approximations to get
+Let $K_i$ be the round key for round $i$ and $U_{4L}$ and $U_{4R}$ the left and right halves of the input to round 4.
+
+\vspace{5mm}
+\pause Using the S-Box 5 approximation and the Piling Up Principle we can combine the approximations to get
 \begin{align*}
  P_L[25] & \oplus P_L[14] \oplus P_L[8] \oplus P_L[3] \oplus \\
-         & U_{4L}[25] \oplus U_{4L}[14] \oplus U_{4L}[8] \oplus U_{4L}[3] \oplus \\
-			& P_R[17] \oplus U_{4R}[17] = K_1[26] \oplus K_3[26] \\
+         & U_{4R}[25] \oplus U_{4R}[14] \oplus U_{4R}[8] \oplus U_{4R}[3] \oplus \\
+			& P_R[17] \oplus U_{4L}[17] = K_1[26] \oplus K_3[26] \\
 \end{align*}
-The bias magnitude for this expression is
+
+\pause The bias magnitude for this expression is
 \[ (12/64)^2 + (1 - 12/64)^2 - 1/2 \approx 0.30 \]
-This is the best expression for the 3-round DES.
-\end{frame}
-
-\begin{frame}
+
+\pause This is the best expression for the 3-round DES.
+\end{frame}
+
+\begin{frame}
+\begin{columns}
+
+\begin{column}{0.35\textwidth}
+\begin{figure}
+\includegraphics[totalheight=0.8\textheight]{DES_4.PNG}
+\end{figure}
+\end{column}
+
+\begin{column}{0.65\textwidth}
+\small
 We perform the attack on the 4-round DES similarly to how we performed the attack on 4-round SPN, by ``undoing'' the last round.\\
-\vspace{5mm}
-\pause
-Note that in the last round, $U_{4L}$ bits are not changed, only $U_{4R}$ bits are affected. Note also that there is only one $U_R$ bit the 3-round equation.
+\vspace{3mm}
+\pause
+Note that in the last round, $U_{4R}$ bits are not changed, only $U_{4L}$ bits are affected. Note also that there is only one $U_{4L}$ bit the 3-round equation.\\
 \begin{align*} 
-P_L[25] \oplus P_L[14] \oplus P_L[8] \oplus & P_L[3] \oplus U_{4L}[25] \oplus U_{4L}[14] \oplus U_{4L}[8] \\ & \oplus U_{4L}[3] \oplus P_R[17] \oplus U_{4R}[17] = K_1[26] \oplus K_3[26]
-\end{align*}
-\vspace{5mm}
-\pause
-This bit, $U_{4R}[17]$ is affected by the result of S-Box 1 in round 4.
+& P_L[25]    \oplus P_L[14]    \oplus P_L[8]    \oplus P_L[3] \oplus  \\
+& U_{4R}[25] \oplus U_{4R}[14] \oplus U_{4R}[8] \oplus U_{4R}[3]  \oplus \\ 
+& P_R[17]    \oplus U_{4L}[17] = K_1[26] \oplus K_3[26] \\
+\end{align*}\\
+\vspace{3mm}
+\pause
+This bit, $U_{4L}[17]$, is affected by the result of S-Box 1 in round 4.
 If we guess correctly for the 6 partial subkey bits $K_{4,1}$ to $K_{4,6}$, the 3-round equation will hold with high bias, otherwise it will likely hold with close to 0 bias.\\
-\vspace{5mm}
-\pause
-Thus, this attack will give us $6$ of the $56$ key bits.
-What this means in terms of time saved over exhaustive search is this: If it takes a year to find a key through exhaustive search of $2^{56}$ bits, after this attack it will only take $5$ days to exhaustively search the remaining $2^{50}$ bits.
+\vspace{3mm}
+\pause
+Thus, this attack will give us $6$ of the $56$ key bits. If it takes a year to find a key through exhaustive search of $2^{56}$ bits, it only takes $5$ days to exhaustively search $2^{50}$ bits.
+\end{column}
+
+\end{columns}
 \end{frame}
@@
 \vspace{3mm}
-Matsui Lemma 5: Let $N$ be the number of given random plaintexts and $\epsilon$, the bias, be sufficiently small.
+\pause Matsui Lemma 5: Let $N$ be the number of given random plaintexts and $\epsilon$, the bias, be sufficiently small.
 Let $q^{(i)}$ be the probability that the following equation holds for a subkey candidate $K_n^{(i)}$ and a random variable X:
 \[ F_n(X,K_n)[l_1, l_2,\dots,l_d] = F_n(X, K_n^{(i)})[l_1, l_2, \dots, l_d].\]
-Then if $q^{(i)}$'s are independent, the success rate of the attack is
+\pause Then if $q^{(i)}$'s are independent, the success rate of the attack is
 \[ \int_{x=-2\sqrt{N}\epsilon}^\infty \left( \prod_{K_n^{(i)} \neq K_n} \int_{-x-4\sqrt{N}\epsilon q^{(i)}}^{x+4\sqrt{N}\epsilon (1 - q^{(i)})} \frac{1}{\sqrt{2\pi}} e^{-y^2/2} dy \right) \frac{1}{\sqrt{2\pi}} e^{-x^2/2} dx \]
-Where the product is taken over all subkey candidates except $K_n$.\\
-\vspace{3mm}
-Let $d = 1$ and $l_1 = 17$ in the above expression - this corresponds to our attack (distribution over guessing for one bit, $CR_{17}$ in the equation).
+where the product is taken over all subkey candidates except $K_n$.
+
+\vspace{3mm}
+\pause Let $d = 1$ and $l_1 = 17$ in the above expression - this corresponds to our attack (distribution over guessing for one bit, $U_{4R}[17]$ in the equation).
 The following table summarizes the success rates:
@@
 Using the calculation from the previous slide, we see that for a 4-round DES, our equation holds with bias of about 1 in 4.
-Thus we will need about $16\times 4^2 = 256$ pairs for the attack to have near perfect chance of success.\\
-\vspace{3mm}
-For a full 16-round DES, the best approximation has bias $\sim 1.19\times 10^{-22}$, thus we would need on the order of $2^{47}$ plaintext-ciphertext pairs.
-\end{frame}
-
-\begin{frame}
-The actual Attack on 4 round DES ... Demo ...
+\pause Thus we will need about $16\times 4^2 = 256$ pairs for the attack to have near perfect chance of success.
+
+\vspace{3mm}
+\pause For a full 16-round DES, the best approximation has bias $\sim 1.19\times 10^{-22}$, thus we would need on the order of $2^{47}$ plaintext-ciphertext pairs.
+\end{frame}
+
+\begin{frame}
+DEMO: Attack on 4-round DES
 \end{frame}
@@
 \begin{frame}
-The chosen input difference:$\Delta P = 20 00 00 00 00 00 00 00$
-
-Round 1 Differential
-
-\[ \Delta P = 20 00 00 00 00 00 00 00 \rightarrow 20 00 00 00 00 00 00 00 = \Delta V_1 \]
-Probability: $p_1 = 1$ as Feistel gives output difference $0$ for input $0$
-
-Round 2 Differential
-
-\[ \Delta U_2 = \Delta V_1 = 20 00 00 00 00 00 00 00 \rightarrow \]
-
-this calculations and coding has not been completed yet ... will be done ...
-\end{frame}
-
-\begin{frame}
-Differentials for SBoxes
-
-\end{frame}
-
-\begin{frame}
-The actual Attack on 4 round DES ... Demo ...
-\end{frame}
-
-\subsection{Extension to full DES}
-\begin{frame}
-Extension of the same ideas to 6 / 8 / 16 round DES ... Probabilistic explanation ...
-
+\begin{columns}
+\begin{column}{0.4\textwidth}
+\begin{figure}
+\includegraphics[totalheight=0.8\textheight]{des_diff_char.jpg}
+\end{figure}
+\end{column}
+
+\begin{column}{0.6\textwidth}
+Differential Characteristic of DES
+
+\vspace{5mm}
+Input Difference: $\Delta P = 20\:00\:00\:00\:00\:00\:00\:00$
+
+\vspace{5mm}
+\pause Propagating Difference Pairs
+\begin{itemize}[<+->]
+\item{Round 1: $S_i$ : $00 \rightarrow 00$}
+\item{Round 2: $S_1$ : $04 \rightarrow 05$}
+\item{Round 3: $S_4$ : $01 \rightarrow 05$ \\ $\qquad \qquad \: S_5$ : $10 \rightarrow 07$ \\ $\qquad \qquad \: S_8$ : $04 \rightarrow 07$}
+\end{itemize}
+
+\vspace{5mm}
+\pause Differential Characteristic
+\[ 20\:00\:00\:00\:00\:00\:00\:00 \rightarrow A0\:06\:18\:A0\:00\:00\:80\:02 \]
+
+\end{column}
+\end{columns}
+\end{frame}
+
+\begin{frame}
+\begin{columns}
+\begin{column}{0.4\textwidth}
+\begin{figure}
+\includegraphics[totalheight=0.8\textheight]{des_diff_char.jpg}
+\end{figure}
+\end{column}
+
+\begin{column}{0.6\textwidth}
+\vspace{2mm}
+\pause Note: $00\:00\:80\:02$ implies difference of 2 bits which are both related to S-Box $S_1$ in round 4.
+
+\vspace{2mm}
+\pause Hence, in round 4
+\[ D' = \Delta C_L \oplus 00\:00\:80\:02 \]
+\pause and hence for $S_2, ... , S_8$, outputs are
+\[ \Delta Y = P^{-1}(\Delta C_L) = S_{box}(E(C_R) \oplus K_4) \]
+
+\vspace{2mm}
+So, just guess the {\it target partial subkey} for $S_2, ... S_8$ individually, \pause and compute their predicted $\Delta Y$ with a number of plaintext pairs.
+
+\vspace{2mm}
+\pause This will ``undo'' round 4 and  the maximum probability of satisfying the 3 round differential characteristic will indicate a correct guess.
+\end{column}
+\end{columns}
+\end{frame}
+
+\begin{frame}
+Probabilistic Analysis
+
+\vspace{5mm}
+Constructing the Difference Distribution tables for the S-Boxes and analyzing the probability distribution for differential propagation through SBoxes:
+
+\vspace{2mm}
+\pause
+\begin{itemize}[<+->]
+\item{Round 1: $S_i$ : $00 \rightarrow 00$ with $Pr = 1$}
+\item{Round 2: $S_1$ : $04 \rightarrow 05$ with $Pr = 10/64$}
+\item{Round 3: $S_4$ : $01 \rightarrow 05$  with $Pr = 16/64$ \\ $\qquad \qquad \: S_5$ : $10 \rightarrow 07$ with $Pr = 12/64$ \\ $\qquad \qquad \: S_8$ : $04 \rightarrow 07$ with $Pr = 12/64$}
+\end{itemize}
+
+\vspace{3mm}
+\pause Differential Characteristic for $\Delta C_L$: $00\:00\:00\:00 \rightarrow 00\:00\:80\:02$ holds with probability
+\[ p_D \: = \: 10/64 \: \approx \: 0.16 \: \gg 1/2^{32} \]
+
+\vspace{2mm}
+\pause
+\begin{itemize}[<+->]
+\item{How many pairs do we need?}
+\item{Approximately $2^4 = 16$ for $100\%$ success}
+\end{itemize}
+
+\end{frame}
+
+\begin{frame}
+DEMO: Attack on 4-round DES
 \end{frame}
@@
 
-\begin{frame}
-What if that is false ? ... Discussion ...
-
-\end{frame}
-
 \subsection{Links between the Attacks}
 \begin{frame}
-Links between Linear and Differential Attacks ...
-
-\end{frame}
-
-\begin{frame}
-Mathematically and while implementing ...
-
+Linear Resistance:\\
+\vspace{3mm}
+Let $F: GF(2)^p \rightarrow GF(2)^q$.\\
+\vspace{3mm}
+Defn: A function F is bent if its Walsh transform $\mathcal{W}\{F\}$ is $\pm 2^{p/2}$.\\
+\vspace{3mm}
+Thm: A function F is perfect nonlinear (ie bias is 0) if and only if it is \textit{bent}. \\
+\vspace{3mm}
+Thm: Bent functions exist only for $p \geq 2q$ and $p$ even. \\
+\vspace{5mm}
+To sumarize: When $p\geq 2q$ and $p$ even, differential-resistant is equivalent to linear resistant, and to vectorial Bentness.
 \end{frame}
@@
 \begin{frame}
-If you are building a cryptosystem ... To Do / Not To Do ...
-
-\end{frame}
-
-\begin{frame}
-Questions ... Doubts ... Coding ... Misc ... Discussion ...
-
+The problems of DES originated in two aspects:
+\begin{itemize}
+\item{6 bit S-Boxes are too easy to identify with linear or affine functions with high bias}
+\item{Diffusion between rounds is just not fast enough}
+\end{itemize}
+AES, which replaced DES, provides an 8-Bit S-Box in the SubBytes step and higher diffusion through the ShiftRows and MixColumns steps. 
+\end{frame}
+
+\begin{frame}
+Matsui rearrangement of S-Boxes ... Construction of S-Boxes ... IBM knew Diff 17 years earlier ... $10^44$ is closer to $10$ than to $10^45$ ...
+\end{frame}
+
+\begin{frame}
+References:\\
+\vspace{5mm}
+M. Matsui, "Linear Cryptanalysis Method for DES Cipher", \textit{Advances in Cryptology - EUROCRYPT '93 (Lecture Notes in Computer Science no. 765),} Springer-Verlag, pp. 38\\
+\vspace{2mm}
+E Biham And A. Shamir, "Differential Cryptanalysis of DES-like Cryptosystems", \textit{Journal of Cryptology}, vol 4, no 1, pp. 3-72, 1991.\\
+\vspace{2mm}
+H. Heys, "A Tutorial on Linear and Differential Cryptanalysis"\\
+\vspace{2mm}
+F.
+Chabaud And S.
+Vaudenay, "Links Between Differential and Linear Cryptanalysis", \textit{Advances in Cryptology - EUROCRYPT '94 (Lexture Notes in Computer Science no.
+950)}, Springer-Verlag, pp.
+356-365, 1995.\\
+\vspace{2mm}
+J. Manferdelli, UW Cryptography Lecture Notes
+\end{frame}
+
+\begin{frame}
+\begin{center}
+The End.\\
+Questions?
+\end{center}
 \end{frame}

### 2009-02-02 src/DifferentialCryptanalysis.java

@@ -1,4 +1,3 @@
-import java.io.IOException;
 import java.util.ArrayList;
 import java.util.BitSet;
 import java.util.List;
@@ -13,7 +12,6 @@ public class DifferentialCryptanalysis {
 
   public static void main(String[] args) {
     DesDifferentialAttack4Round();
-    // DesDifferentialAttack3Round();
     // PrintNSforAll();
   }
 
@@ -151,12 +149,13 @@ public class DifferentialCryptanalysis {
     return sets;
   }
 
+
   /**
  * 
  */
-  public static void DesDifferentialAttack3Round() {
-    // do analysis for 3-round DES
-    int numRounds = 3;
+  public static void DesDifferentialAttack4Round() {
+    // do analysis for 4-round DES
+    int numRounds = 4;
 
     // set the key
     BitSet key1 = Util.toBitSet(0x1234, 16);
@@ -169,229 +168,119 @@ public class DifferentialCryptanalysis {
 
     // print the actual key bits
     DesImpl des = new DesImpl();
-    BitSet K3 = des.KeySchedule(key, 3);
-    BitSet K3_S5 = K3.get(24, 30);
-
-    System.out.println("actual subkey is "
-        + Integer.toHexString(Util.toInteger(K3_S5, 6)));
-
-    // generate the pairs
-    int numPairs = 10000;
-    System.out.print("Generating " + numPairs + " pairs...");
-    List<BitSet[]> pairs = generateRandomPairs(numPairs, key, numRounds);
-    System.out.println("done");
-    System.out.println("Good pairs: " + pairs.size());
-
-    int[] counts = new int[64];
-    for (int i = 0; i < counts.length; i++) {
-      counts[i] = 0;
-    }
-    int count = 0;
-    for (BitSet[] pair : pairs) {
-      if (count++ % (numPairs / 100) == 0) {
-        System.out.print(((int) (count / ((double) numPairs) * 100)) + "%\n");
-      }
-
-      BitSet C1L = pair[0].get(0, 32);
-      BitSet C1R = pair[0].get(32, 64);
-      BitSet C2L = pair[1].get(0, 32);
-      BitSet C2R = pair[1].get(32, 64);
-
-      for (int i = 0; i < 64; i++) {
-        // guess the key for Sbox 5
-        BitSet partialSubkeyGuess = Util.toBitSet(i, 6);
-        // pad it
[…]

### 2009-02-02 trunk/talk/draft1.tex
Completed talk ... Final PDF included ... Some more figures ...

@@
 \[ 20\:00\:00\:00\:00\:00\:00\:00 \rightarrow A0\:06\:18\:A0\:00\:00\:80\:02 \]
+
+\end{column}
+\end{columns}
+\end{frame}
+
+\begin{frame}
+\begin{columns}
+\begin{column}{0.4\textwidth}
+\begin{figure}
+\includegraphics[totalheight=0.8\textheight]{des_diff_char.jpg}
+\end{figure}
+\end{column}
+
+\begin{column}{0.6\textwidth}
+\vspace{2mm}
+\pause Note: $00\:00\:80\:02$ implies difference of 2 bits which are both related to S-Box $S_1$ in round 4.
+
+\vspace{2mm}
+\pause Hence, in round 4
+\[ D' = \Delta C_L \oplus 00\:00\:80\:02 \]
+\pause and hence for $S_2, ... , S_8$, outputs are
+\[ \Delta Y = P^{-1}(\Delta C_L) = S_{box}(E(C_R) \oplus K_4) \]
+
+\vspace{2mm}
+So, just guess the {\it target partial subkey} for $S_2, ... S_8$ individually, \pause and compute their predicted $\Delta Y$ with a number of plaintext pairs.
+
+\vspace{2mm}
+\pause This will ``undo'' round 4 and  the maximum probability of satisfying the 3 round differential characteristic will indicate a correct guess.
 \end{column}
@@
 
-\vspace{5mm}
+\vspace{3mm}
 Constructing the Difference Distribution tables for the S-Boxes and analyzing the probability distribution for differential propagation through SBoxes:
@@
 \vspace{3mm}
-\pause Differential Characteristic: $20\:00\:00\:00\:00\:00\:00\:00 \rightarrow A0\:06\:18\:A0\:00\:00\:80\:02$ holds with probability
-\[ p_D \: = \: (10/64) \times (16/64) \times (12/64) \times (12/64) \: = \: 45/32768 \approx 0.001373 \]
-
-\vspace{2mm}
-\pause Note: $p_D$ is much higher than the random distribution probability $1/2^{64}$ and we hope to exploit this situation to break DES.
-
-\end{frame}
-
-\begin{frame}
-\begin{columns}
-\begin{column}{0.4\textwidth}
-\begin{figure}
-\includegraphics[totalheight=0.8\textheight]{des_diff_char.jpg}
-\end{figure}
-\end{column}
-
-\begin{column}{0.6\textwidth}
-Differential Characteristic of DES
-\[ 20\:00\:00\:00\:00\:00\:00\:00 \rightarrow A0\:06\:18\:A0\:00\:00\:80\:02 \]
-
-\pause Note: The right half of the round 3 characteristic output is $00\:00\:80\:02$.
-
-\vspace{2mm}
-\pause Thus, the right halves of the ciphertexts at this round differ by 2 bits, and these are related to S-Box $S_1$ in round 4.
-
-\vspace{2mm}
-\pause Hence, we just need to guess the {\it target partial subkey} $K_{4,1}, ..., K_{4,6}$ for $S_1$, i.e. the first 6 bits of $K_4$.
-
-\vspace{2mm}
-\pause Thereafter, we can ``undo'' round 4 using this guess and  the maximum probability of satisfying the 3 round differential characteristic will indicate a correct guess.
-\end{column}
-\end{columns}
-\end{frame}
-
-\begin{frame}
-How many pairs do we need? (Probabilistic justification)
-
+\pause Differential Characteristic for $\Delta C_L$: $00\:00\:00\:00 \rightarrow 00\:00\:80\:02$ holds with probability
+\[ p_D \: = \: 10/64 \: \approx \: 0.16 \: \gg 1/2^{32} \]
+
+\vspace{2mm}
+\pause
+\begin{itemize}[<+->]
+\item{How many pairs do we need?}
+\item{Approximately $2^4 = 16$ for $100\%$ success}
+\end{itemize}
+
+\pause For 16 round DES, $2^{47}$ pairs are needed. [$2^{36}$ ``good pairs'']
 \end{frame}
@@
 
-\subsection{Independence of S-Boxes}
-\begin{frame}
-Note that throughout this discussion we have assumed that the S-Boxes were independent. This allowed us to use the Piling Up Principle to combine linear approximations of S-Boxes across rounds. \\
-\vspace{3mm}
-This assumption worked well for us in practice, but is not necessarily true. We could have proceeded differently without this assumption - [see John Manferdelli's Boolean Functions slides]
-\end{frame}
-
-\subsection{Links between the Attacks}
-\begin{frame}
-Linear Resistance:\\
-\vspace{3mm}
+\subsection{A few important points}
+\begin{frame}
+Independence of S-Boxes
+
+\vspace{5mm}
+\pause Note that throughout this discussion we have assumed that the S-Boxes were independent. This allowed us to use the Piling Up Principle to combine linear approximations of S-Boxes across rounds.
+
+\vspace{3mm}
+\pause This assumption worked well for us in practice, but is not necessarily true. We could have proceeded differently without this assumption - [see John Manferdelli's Boolean Functions slides]
+\end{frame}
+
+\begin{frame}
+Linear Resistance
+
+\vspace{5mm}
 Let $F: GF(2)^p \rightarrow GF(2)^q$.\\
 \vspace{3mm}
-Defn: A function F is bent if its Walsh transform $\mathcal{W}\{F\}$ is $\pm 2^{p/2}$.\\
-\vspace{3mm}
-Thm: A function F is perfect nonlinear (ie bias is 0) if and only if it is \textit{bent}. \\
-\vspace{3mm}
-Thm: Bent functions exist only for $p \geq 2q$ and $p$ even. \\
-\vspace{5mm}
-To sumarize: When $p\geq 2q$ and $p$ even, differential-resistant is equivalent to linear resistant, and to vectorial Bentness.
+\pause Defn: A function F is bent if its Walsh transform $\mathcal{W}\{F\}$ is $\pm 2^{p/2}$.\\
+\vspace{3mm}
+\pause Thm: A function F is perfect nonlinear (ie bias is 0) if and only if it is \textit{bent}. \\
+\vspace{3mm}
+\pause Thm: Bent functions exist only for $p \geq 2q$ and $p$ even. \\
+\vspace{5mm}
+\pause To sumarize: When $p\geq 2q$ and $p$ even, differential-resistant is equivalent to linear resistant, and to vectorial Bentness.
 \end{frame}
@@
 \begin{beamercolorbox}[ht=2.5ex,dp=1.125ex,center,rounded=true,shadow=true]{author in head/foot}
-`Take Home'
-\end{beamercolorbox}
-\end{frame}
-
-\begin{frame}
-If you are mounting the attacks ... To Do / Not To Do ...
-
-\end{frame}
-
-\begin{frame}
-The problems of DES originated in two aspects:
-\begin{itemize}
+`Moral of the story'
+\end{beamercolorbox}
+\end{frame}
+
+\begin{frame}
+If you are mounting the attacks, remember the following:
+
+\vspace{5mm}
+\pause
+\begin{itemize}[<+->]
+\item{Analyze the main non-linear component, S-Boxes for example}
+\item{For linear attack, try to find the linear/affine approximation}
+\item{For differential attack, find the best characteristic}
+\item{Combine through levels to get the total approximation/characteristic}
+\item{Analyze piling up probabilities to estimate number of required pairs}
+\item{Verify if this is better than any other existing attack}
+\end{itemize}
+
+\end{frame}
+
+\begin{frame}
+If you are designing a block cipher, note:
+
+\vspace{5mm}
+\pause The problems of DES originated in two aspects
+\begin{itemize}[<+->]
 \item{6 bit S-Boxes are too easy to identify with linear or affine functions with high bias}
@@
 \end{itemize}
-AES, which replaced DES, provides an 8-Bit S-Box in the SubBytes step and higher diffusion through the ShiftRows and MixColumns steps. 
+
+\vspace{2mm}
+\pause Prescription
+\begin{itemize}[<+->]
+\item{Create S-Boxes as bent functions [inefficient as $p \geq 2q$]}
+\item{Make the diffusion step faster to create an avalanche}
+\end{itemize}
+
+\vspace{3mm}
+\pause AES, which replaced DES, provides an 8-Bit S-Box in the SubBytes step and higher diffusion through the ShiftRows and MixColumns steps. 
+\end{frame}
+
+\subsection{Fun facts}
+\begin{frame}
+\begin{beamercolorbox}[ht=2.5ex,dp=1.125ex,center,rounded=true,shadow=true]{author in head/foot}
+Just for Fun!
+\end{beamercolorbox}
+\end{frame}
+
+\begin{frame}
+What if you play around with the S-Box ordering?
+
+\vspace{5mm}
+\pause
+\begin{itemize}[<+->]
+\item{Matsui prescribed a rearrangement of S-Boxes specifically against linear attack.}
+\item{Matsui prescribed another rearrangement of the S-Boxes, which is good enough to prevent both linear and differential attacks.}
+\item{Biham and Shamir illustrated a rearrangement where differential attack becomes way more efficient.}
+\end{itemize}
+
+\end{frame}
+
+\begin{frame}
+\begin{figure}
+\includegraphics[width=0.6\textwidth]{ibm.jpg}
+\end{figure}
+
+\pause
+S-Box Design criteria
+\begin{itemize}[<+->]
+\item{1 bit input difference produces 2 bits output difference}
+\item{Minimize the difference between the numbers of 1's and 0's when any input bit is held constant}
+\item{$S(X) \neq S(X \oplus 11ab00)$} 
+\end{itemize}
+
+\vspace{2mm}
+\pause Other good stuff
+\begin{itemize}[<+->]
+\item{2 input difference bits mapped to 3 by $E(R)$}
+\item{Each output difference bit permuted to 1 each to S-boxes of next round}
+\end{itemize}
+
+\end{frame}
+
+\begin{frame}
+\begin{figure}
+\includegraphics[totalheight=0.7\textheight]{coppersmith.jpg}
+\end{figure}
+\pause
+\centerline{Average of $10$ and $10^{45}$ is $5.10^{44} + 5 > 10^{44} \approx 2^{47}$}
+\pause
+\begin{figure}
+\includegraphics[width=0.7\textwidth]{eli.jpg}
+\end{figure}
+\hfill - Eli Biham
 \end{frame}
@@
 \begin{frame}
-\begin{center}
-The End.\\
+\begin{beamercolorbox}[ht=2.5ex,dp=1.125ex,center,rounded=true,shadow=true]{author in head/foot}
+The End.
+\end{beamercolorbox}
+\end{frame}
+
+\begin{frame}
+\begin{beamercolorbox}[ht=2.5ex,dp=1.125ex,center,rounded=true,shadow=true]{author in head/foot}
 Questions?
-\end{center}
+\end{beamercolorbox}
 \end{frame}

### 2009-02-02 trunk/src/LinearCryptanalysis.java
Pause in DES Differential re-inserted

@@ -290,7 +290,7 @@ public class LinearCryptanalysis {
     BitSet K4 = des.KeySchedule(key, 4);
 
     
-    int numPairs = 1000;
+    int numPairs = 10000;
     
     // print info 
     System.out.println("4-round DES Linear Cryptanalysis.");

### 2009-02-02 src/DifferentialCryptanalysis.java
Pause in DES Differential re-inserted

@@ -1,3 +1,4 @@
+import java.io.IOException;
 import java.util.ArrayList;
 import java.util.BitSet;
 import java.util.List;
@@ -169,16 +170,31 @@ public class DifferentialCryptanalysis {
     // print the actual key bits
     DesImpl des = new DesImpl();
     BitSet K4 = des.KeySchedule(key, 4);
+    int numPairs = 16;
     
-    System.out.print("Actual key is   ");
-    	    
+    // print info 
+    System.out.println("4-round DES Linear Cryptanalysis.");
+    System.out.println("Using " + numPairs + " pairs.");
+    System.out.println("****************");
+    System.out.println("Using Key:");
+    Util.printBitSet(key, 64);
+    System.out.println("****************");
+    System.out.println("Real round 4 key:");
     for (int i = 0; i < 8; i++){
-    	BitSet K4_S = K4.get(6*i, 6*(i+1));
-    	System.out.print(Integer.toHexString(Util.toInteger(K4_S, 6)) + " ");
+      BitSet K4_S = K4.get(6*i, 6*(i+1));
+      System.out.print("\tS-Box " + (i+1) + ":\t");
+      System.out.print(Integer.toHexString(Util.toInteger(K4_S, 6)) + " ");
+      System.out.println();
+    }
+    System.out.println("****************");
+    System.out.println("Press Enter to run linear attack...");
+    try {
+      System.in.read();
+    } catch(IOException exn) {
+      
     }
     
     // generate the pairs
-    int numPairs = 16;
     System.out.println("");
     System.out.println("Generating " + numPairs + " pairs...");
     List<BitSet[]> pairs = generateRandomPairs(numPairs, key, numRounds);
@@ -191,7 +207,6 @@ public class DifferentialCryptanalysis {
 	    }
     }
     
-    int count = 0;
     for (BitSet[] pair : pairs) {
 //      if (count++ % (numPairs / 100) == 0) {
 //        System.out.print(((int) (count / ((double) numPairs) * 100)) + "%\n");

### 2009-02-02 talk/draft1.tex
Now there is the final final version ... :)

@@
 \begin{itemize}[<+->]
-\item{How many pairs do we need?}
-\item{Approximately $2^4 = 16$ for $100\%$ success}
-\end{itemize}
-
+\item{Question: How many pairs do we need?}
+\item{Answer: $N \approx c/p_D$ where $p_D$ is the probability of the differential characteristic, and $c > 0$ is a constant}
+\end{itemize}
+\pause I used $2^4 = 16$ pairs for $100\%$ success in this case. For 16 round DES, $2^{47}$ pairs are needed, out of which $2^{36}$ are ``good pairs''.
 \end{frame}
@@
 \end{itemize}
-AES, which replaced DES, provides an 8-Bit S-Box in the SubBytes step and higher diffusion through the ShiftRows and MixColumns steps. 
-\end{frame}
-
-\begin{frame}
-Matsui rearrangement of S-Boxes ... Construction of S-Boxes ... IBM knew Diff 17 years earlier ... $10^44$ is closer to $10$ than to $10^45$ ...
+
+\vspace{2mm}
+\pause Prescription
+\begin{itemize}[<+->]
+\item{Create S-Boxes as bent functions [inefficient as $p \geq 2q$]}
+\item{Make the diffusion step faster to create an avalanche}
+\end{itemize}
+
+\vspace{3mm}
+\pause AES, which replaced DES, provides an 8-Bit S-Box in the SubBytes step and higher diffusion through the ShiftRows and MixColumns steps. 
+\end{frame}
+
+\subsection{Fun facts}
+\begin{frame}
+\begin{beamercolorbox}[ht=2.5ex,dp=1.125ex,center,rounded=true,shadow=true]{author in head/foot}
+Just for Fun!
+\end{beamercolorbox}
+\end{frame}
+
+\begin{frame}
+What if you play around with the S-Box ordering?
+
+\vspace{5mm}
+\pause
+\begin{itemize}[<+->]
+\item{Matsui prescribed a rearrangement of S-Boxes specifically against linear attack.}
+\item{Matsui prescribed another rearrangement of the S-Boxes, which is good enough to prevent both linear and differential attacks.}
+\item{Biham and Shamir illustrated a rearrangement where differential attack becomes way more efficient.}
+\end{itemize}
+
+\end{frame}
+
+\begin{frame}
+\begin{figure}
+\includegraphics[width=0.6\textwidth]{ibm.jpg}
+\end{figure}
+
+\pause
+S-Box Design criteria
+\begin{itemize}[<+->]
+\item{1 bit input difference produces 2 bits output difference}
+\item{Minimize the difference between the numbers of 1's and 0's when any input bit is held constant}
+\item{$S(X) \neq S(X \oplus 11ab00)$} 
+\end{itemize}
+
+\vspace{2mm}
+\pause Other good stuff
+\begin{itemize}[<+->]
+\item{2 input difference bits mapped to 3 by $E(R)$}
+\item{Each output difference bit permuted to 1 each to S-boxes of next round}
+\end{itemize}
+
+\end{frame}
+
+\begin{frame}
+\begin{figure}
+\includegraphics[totalheight=0.7\textheight]{coppersmith.jpg}
+\end{figure}
+\pause
+\centerline{Average of $10$ and $10^{45}$ is $5.10^{44} + 5 > 10^{44} \approx 2^{47}$}
+\pause
+\begin{figure}
+\includegraphics[width=0.7\textwidth]{eli.jpg}
+\end{figure}
+\hfill - Eli Biham
 \end{frame}

### 2009-03-09 talk/draft1.tex
Final Updates

@@
 \vspace{3mm}
-\pause For a full 16-round DES, the best approximation has bias $\sim 1.19\times 10^{-22}$, thus we would need on the order of $2^{47}$ plaintext-ciphertext pairs.
+\pause For a full 16-round DES, the best approximation has bias $\sim 1.19\times 10^{-22}$, thus we would need on the order of $2^{47}$ plaintext-ciphertext pairs.\\
+\vspace{3mm}
+The 15-round best expression has the following form:
+\[ P_L[8] \oplus P_L[14] \oplus P_L[25] \oplus P_R[16] \oplus P_R[20] \oplus \]
+\[  U_{16L}[17] \oplus U_{16R}[3] \oplus U_{16R}[8] \oplus U_{16R}[14] \oplus U_{16R}[25] = \Sigma K \]
 \end{frame}
